\chapter{What we're building (and why)}
\label{ch:what-and-why}

\section{What you'll build}
In this chapter, you will build the top-level contract for a deterministic UI runtime:

\begin{itemize}
  \item A crisp definition of what a UI runtime must do (model \textrightarrow mount \textrightarrow patch \textrightarrow destroy).
  \item The two execution paths used in this book (VDOM baseline and HRBR compiler-assisted fast path).
  \item A practical workflow for connecting claims to evidence (verification contracts encoded as tests).
\end{itemize}

\section{What you'll learn}
\begin{itemize}
  \item Why UI libraries exist: the DOM is an output device, not a state container.
  \item The minimum set of concepts you need to build a correct renderer.
  \item How to interpret benchmark results without lying to yourself (preview).
\end{itemize}

\section{Prereqs}
No JavaScript/TypeScript tricks are required.
I only assume you can read code, and you understand that DOM nodes are mutable objects.

\section{Start from first principles (not from frameworks)}
We start from a logical contract, not from an API surface.
Here is the problem statement that drives the architecture:

\begin{quote}
We want a deterministic way to turn \emph{state} into \emph{DOM}, and then update that DOM efficiently as state changes.
\end{quote}

This gives us four responsibilities.
If you understand these, you understand 80\% of any UI runtime:

\begin{enumerate}
  \item \textbf{Model}: represent desired UI in memory.
  \item \textbf{Mount}: turn the model into DOM for the first time.
  \item \textbf{Patch}: update DOM when the model changes.
  \item \textbf{Destroy}: cleanup (remove nodes, listeners, and references).
\end{enumerate}

Relax.js is one concrete implementation of those responsibilities.
Throughout the book, we will judge every design decision against three runtime properties:

\begin{itemize}
  \item \textbf{Correctness}: stable invariants, validated by tests.
  \item \textbf{Performance}: minimize avoidable DOM work and repeated discovery.
  \item \textbf{Predictability}: explicit scheduling and lifecycle boundaries.
\end{itemize}

\section{The elevator pitch (Relax in one minute)}
Relax.js is a small UI runtime built around a Virtual DOM (VDOM) and a component model.
What makes it a useful systems case study is that it contains \emph{two} execution paths:

\begin{enumerate}
  \item A traditional VDOM runtime under \texttt{src/} (general case).
  \item A compiler-assisted runtime (HRBR) that updates the DOM through precomputed \textbf{slots} (fast path for stable structure).
\end{enumerate}

The goal of this book is not to teach a framework API.
It is to teach a buildable architecture: contracts, invariants, and failure modes.
We will still read real source code, but the emphasis is on transferable design.

This isn't a conceptual ``frameworks 101'' book.
It's a guided runtime engineering walk-through:

\begin{itemize}
  \item You will learn the smallest set of abstractions needed to describe UI.
  \item You will see how those abstractions are represented in memory.
  \item You will trace how a rendering pass turns them into real DOM.
  \item You will follow how updates are scheduled and applied.
  \item You will use the test suite as the specification that locks behavior in place.
\end{itemize}

By the end, you should be able to (1) understand the architecture end-to-end, and (2) rebuild a similar deterministic runtime with confidence.

\section{A tiny contract: what does Relax.js promise?}
At a high level, Relax.js promises:

\begin{itemize}
  \item You describe UI as a tree (VNode tree or compiled block).
  \item The runtime mounts it into a DOM host element.
  \item When state changes, only the necessary DOM updates happen.
  \item Components have lifecycle hooks and can emit/handle events.
\end{itemize}

\subsection*{Diagram: the end-to-end VDOM update loop (mount \textrightarrow patch \textrightarrow destroy)}
Even if you later adopt the HRBR compiled fast path, the VDOM loop is the baseline contract. Most tests in \texttt{src/\_\_tests\_\_/*} are describing one of these arrows.

\begin{center}
\begin{tikzpicture}[
  node distance=10mm and 14mm,
  box/.style={draw, rounded corners, align=center, inner sep=6pt, fill=RelaxLight},
  arrow/.style={-Latex, thick}
]
  \node[box] (vnode) {VNode tree\\(plain objects)};
  \node[box, right=of vnode] (mount) {\texttt{mountDOM(vdom, host)}};
  \node[box, right=of mount] (dom) {Real DOM\\\texttt{vdom.el} refs};
  \node[box, below=of mount] (patch) {\texttt{patchDOM(old, next)}};
  \node[box, below=of dom] (destroy) {\texttt{destroyDOM(vdom)}};

  \draw[arrow] (vnode) -- (mount);
  \draw[arrow] (mount) -- (dom);
  \draw[arrow] (dom) -- node[right]{updates} (patch);
  \draw[arrow] (patch) -- node[left]{reuses identity when possible} (dom);
  \draw[arrow] (dom) -- (destroy);
\end{tikzpicture}
\end{center}

This is the book's core promise: if two VNodes represent the "same" thing, Relax tries to keep the same DOM node and just update its values.

\subsection*{Diagram: the two execution paths (VDOM vs HRBR)}
\begin{center}
\begin{tikzpicture}[
  node distance=12mm and 12mm,
  box/.style={draw, rounded corners, align=center, inner sep=6pt, fill=RelaxLight},
  arrow/.style={-Latex, thick}
]
\node[box] (author) {Authoring\\(TSX/JSX or \texttt{h()})};
\node[box, below left=of author] (vdom) {VDOM runtime\\\texttt{src/*}\\mount/patch/destroy};
\node[box, below right=of author] (compiler) {Compiler\\\texttt{compiler/*}\\(optional)};
\node[box, below=of compiler] (hrbr) {HRBR runtime\\\texttt{runtime/*}\\template + slots};
\node[box, below=of vdom, xshift=18mm] (dom) {Real DOM\\(browser/jsdom)};

\draw[arrow] (author) -- node[midway, left] {general case} (vdom);
\draw[arrow] (author) -- node[midway, right] {optimized path} (compiler);
\draw[arrow] (compiler) -- (hrbr);
\draw[arrow] (vdom) -- (dom);
\draw[arrow] (hrbr) -- (dom);
\end{tikzpicture}
\end{center}

This diagram is the mental model for the rest of the book: Relax always ends at \emph{real DOM updates}, but it can reach them either by interpreting a VNode tree (VDOM) or by executing precomputed slot writes (HRBR).

Those promises are verified by tests under \texttt{src/\_\_tests\_\_/} and (for the HRBR runtime) under \texttt{runtime/\_\_tests\_\_/}.

\subsection{A quick terminology primer}
You'll see these words throughout the code and tests:

\begin{itemize}
  \item \textbf{VNode}: a plain object that describes what DOM you want.
  \item \textbf{Mount}: create DOM from a VNode for the first time.
  \item \textbf{Patch}: update existing DOM by comparing old/new VNodes.
  \item \textbf{Destroy}: remove DOM nodes and detach listeners.
  \item \textbf{Host element}: the DOM element you mount the app into (often \texttt{document.body} in tests).
  \item \textbf{HRBR block}: a compiled representation: template HTML + dynamic \emph{slots}.
  \item \textbf{Slot}: a precomputed ``write location'' (text/attr/prop/class/style/event) in a template.
\end{itemize}

\section{Public API entry points (code references)}
The \emph{public} exports live in \texttt{src/index.ts}:

\begin{lstlisting}[style=relax,language=Java]
export { createApp } from './app'
export { defineComponent } from './component'
export { DOM_TYPES, h, hFragment, hSlot, hString } from './h'
export { nextTick } from './scheduler'
\end{lstlisting}

This tells you the learning order:

\begin{enumerate}
  \item \texttt{createApp()} (app lifecycle)
  \item \texttt{defineComponent()} (component model)
  \item \texttt{h()/hFragment()} (VNode model)
  \item \texttt{mountDOM()/patchDOM()/destroyDOM()} (DOM mechanics)
  \item \texttt{nextTick()} (async scheduling in the VDOM runtime)
\end{enumerate}

\section{Why a second runtime? HRBR as a justified architecture choice}

HRBR is the unique differentiator in this repository.
If it isn't justified early, it looks like a random experiment.
So let's state the motivation clearly and formally.

\subsection{Problem statement}

We want a UI runtime that is simultaneously:

\begin{itemize}
  \item \textbf{Correct by construction}: updates are deterministic and testable.
  \item \textbf{Ergonomic to author}: JSX/TSX stays readable and composable.
  \item \textbf{Fast for stable subtrees}: when a subtree's structure is stable, updates should avoid repeated interpretation work.
\end{itemize}

A classic VDOM runtime solves the first two nicely, but it pays a recurring per-update cost:
allocate VNodes, traverse them, and interpret/diff structure to discover \emph{where} to write.

HRBR (\textbf{H}ybrid \textbf{R}untime with \textbf{B}locks and \textbf{R}eactivity) is the opposite trade-off:
it moves some work earlier (compile/transform time) so the runtime can update a stable subtree through precomputed writes.

\subsection{The HRBR idea (one sentence)}

If the DOM \emph{shape} is stable, we can compile it into a stable template and update it by writing only the changing values (\emph{slots}).

\subsection{How HRBR works (contract-level view)}

HRBR has two core artifacts:

\begin{itemize}
  \item A \textbf{template} (static HTML) that is parsed once.
  \item A set of \textbf{slots} that specify \emph{what} to write (text/attr/prop/class/style/event) and \emph{where} to write it (a DOM path).
\end{itemize}

At runtime, an update becomes: \emph{evaluate value} \textrightarrow \emph{write value} to known nodes.

You'll see HRBR show up in VDOM code via \texttt{DOM\_TYPES.HRBR} and \texttt{hBlock()} in \texttt{src/h.ts},
and then fully defined in the HRBR runtime chapters (template + slots, fallback, hydration).

\subsection{Why not just optimize the VDOM?}

You can (and we do), but there are two costs VDOM can't fully erase:

\begin{itemize}
  \item \textbf{Discovery cost:} you must discover where to write by walking/diffing the tree.
  \item \textbf{Allocation cost:} you repeatedly allocate VNodes/arrays/closures as part of rendering.
\end{itemize}

For structurally stable subtrees, those costs are overhead.
HRBR exists to provide a fast path when the shape is known.

\subsection{Comparison (where HRBR sits relative to React/Vue/Solid)}

This book isn't a framework survey, but you do need a mental coordinate system:

\begin{itemize}
  \item \textbf{React (VDOM + reconciliation):} very flexible, pays an interpretation cost each update.
  \item \textbf{Vue (template compiler + runtime):} compiles templates to reduce work; still maintains a runtime update system.
  \item \textbf{Solid (fine-grained reactivity):} reduces re-render traversal by tracking dependencies; still needs DOM update sites.
  \item \textbf{Relax HRBR:} compiles stable templates into \emph{blocks} with explicit slot write locations, then drives updates through deterministic scheduling.
\end{itemize}

The punchline: HRBR is not ``better VDOM''.
It's a different point in the design space: pay compilation complexity to reduce runtime interpretation work for stable structure.

\subsection{Performance rationale (what HRBR is trying to buy)}

HRBR is motivated by a measurable goal:

\begin{itemize}
  \item When structure is stable, updates should trend toward \textbf{O(number of changed values)} rather than \textbf{O(size of subtree)}.
\end{itemize}

This does not make benchmarks magically faster in every case.
It is an architecture that can remove categories of overhead (tree traversal, diff discovery) when the compiler can guarantee the structure.

\subsection{Safety valve: fallback when structure isn't stable}

Compilers only work when the structure is predictable.
So the HRBR system includes a built-in safety valve:
when structure is dynamic (conditionals/lists/fragments with variable shape), compiled output routes to a fallback runtime.

That decision and boundary are explained later in Chapter~\ref{ch:compiler-architecture}.

\subsection*{Diagram: when HRBR is a good fit (static structure, dynamic data)}
This isn't a binary choice; think of HRBR as a selective optimization for subtrees that are structurally stable.

\begin{center}
\begin{tikzpicture}[
  node distance=9mm and 12mm,
  box/.style={draw, rounded corners, align=left, inner sep=6pt, fill=RelaxLight},
  arrow/.style={-Latex, thick},
  decision/.style={diamond, draw, aspect=2.2, align=center, inner sep=2pt, fill=RelaxLight}
]
  \node[box] (jsx) {Author TSX/JSX\\("what" UI looks like)};
  \node[decision, right=18mm of jsx] (stable) {Is structure\\stable?\\(same tags/shape)};
  \node[box, above right=10mm and 14mm of stable] (vdom) {Use VDOM\\diff + patch\\general case};
  \node[box, below right=10mm and 14mm of stable] (hrbr) {Use HRBR\\template + slots\\fast path};
  \node[box, right=22mm of stable] (dom2) {DOM updates\\(same result)};

  \draw[arrow] (jsx) -- (stable);
  \draw[arrow] (stable) -- node[above]{no / dynamic lists} (vdom);
  \draw[arrow] (stable) -- node[below]{yes / mostly values} (hrbr);
  \draw[arrow] (vdom) -- (dom2);
  \draw[arrow] (hrbr) -- (dom2);
\end{tikzpicture}
\end{center}

\section{How we'll prove correctness throughout this book}
This repository uses \textbf{Vitest} and \textbf{jsdom}.
Our rule is simple: every time we explain a behavior, we should be able to point to a unit test.

Examples:

\begin{itemize}
  \item Mounting basics: \texttt{src/\_\_tests\_\_/mount-dom.test.ts}
  \item Patching/diffing: \texttt{src/\_\_tests\_\_/patch-dom.test.ts}
  \item Components + lifecycle: \texttt{src/\_\_tests\_\_/component*.test.ts}
  \item HRBR block runtime: \texttt{runtime/\_\_tests\_\_/block.test.ts}
\end{itemize}

\subsection{What ``verification contract'' means in this codebase}
You should treat tests here as \emph{contracts}.
When you change runtime behavior, you don't just ``fix failing tests''---you decide whether the contract changed.

Most tests follow the same simple pattern:

\begin{enumerate}
  \item Set up a small DOM (usually \texttt{document.body}).
  \item Build a VNode tree (or HRBR block).
  \item Mount, patch, or destroy.
  \item Assert on resulting \texttt{innerHTML} or element identity.
\end{enumerate}

That gives you two superpowers as a reader:

\begin{itemize}
  \item You can infer semantics without reading every implementation detail.
  \item You can reproduce bugs by writing a tiny test first.
\end{itemize}

\section{Build-along: the smallest end-to-end loop (with real code)}
Let's walk through the smallest app lifecycle you can understand in one sitting.
This is the loop I use when I sanity-check any renderer:

\begin{enumerate}
  \item Build a model of UI (VNode tree).
  \item Mount it into a host element.
  \item Change state by producing a new model.
  \item Patch the DOM to match.
  \item Destroy and ensure cleanup is complete.
\end{enumerate}

\subsection{createApp(): mount and unmount}
The application API is defined in \texttt{src/app.ts}.
The critical part is that \texttt{createApp()} remembers three pieces of state:

\begin{itemize}
  \item \texttt{parentEl}: where we mounted
  \item \texttt{isMounted}: guard against double mount/unmount
  \item \texttt{vdom}: the root VNode tree
\end{itemize}

When you mount:
\begin{enumerate}
  \item it throws if already mounted
  \item it creates the root VNode by calling \texttt{h(RootComponent, props)}
  \item it calls \texttt{mountDOM(vdom, parentEl)}
\end{enumerate}

When you unmount:
\begin{enumerate}
  \item it throws if not mounted
  \item it calls \texttt{destroyDOM(vdom)}
  \item it resets internal references
\end{enumerate}

This is a small but important architectural choice: \emph{the App owns the root VNode}.
That makes ``unmount'' a reliable cleanup boundary.

\subsection{A small snippet you should internalize}
When you strip away everything else, most UI libraries are some version of this:

\begin{lstlisting}[style=relax,language=JavaScript]
let oldTree = null

function render(nextTree, host) {
  if (oldTree == null) {
    mountDOM(nextTree, host)
  } else {
    patchDOM(oldTree, nextTree, host)
  }
  oldTree = nextTree
}

function dispose() {
  if (oldTree != null) destroyDOM(oldTree)
  oldTree = null
}
\end{lstlisting}

	\textbf{Why this matters:}
if you can reason about this loop, you can reason about the entire Relax VDOM runtime.
All the complexity is in the details: fragments, keyed reorders, events, components, and scheduling.

\subsection{The test that proves it works}
Open \texttt{src/\_\_tests\_\_/app.test.ts}.
It reads like a story:

\begin{itemize}
  \item It creates an app via \texttt{createApp(App, \{ todos: [...] \})}.
  \item It mounts to \texttt{document.body}.
  \item It waits for \texttt{nextTick()} to flush async work.
  \item It asserts a stable HTML snapshot via \texttt{singleHtmlLine\ldots}.
\end{itemize}

Notice the discipline here: the tests make DOM updates predictable by explicitly awaiting \texttt{nextTick()}.
That foreshadows Chapter~\ref{ch:scheduling}, where we explain why scheduling exists and how it interacts with DOM patching.

\section{Why this book is written the way it is (my bias as the author)}
I built Relax by insisting on a few simple rules:

\begin{itemize}
  \item \textbf{Small surface area}: if it's not essential, it stays internal.
  \item \textbf{Contracts before cleverness}: tests come first, performance work comes second.
  \item \textbf{Measure what users feel}: tail latency and dropped frames matter more than averages.
\end{itemize}

This book follows the same rules.
You'll see me repeat the same pattern constantly: define the contract, show the smallest implementation, then harden it against edge cases.

\section{Common pitfalls (what usually goes wrong)}
When you build a UI runtime, these are the mistakes that create the nastiest bugs:

\begin{itemize}
  \item Confusing \textbf{attributes} with \textbf{properties} (it works for \texttt{div}, fails for \texttt{input}).
  \item Forgetting to remove event listeners on destroy (leaks + ghost behavior).
  \item Replacing DOM nodes on update instead of reusing identity when structure matches.
  \item Using only averages to judge performance (p95 and drops are the real story).
  \item Letting scheduling semantics become implicit (tests get flaky, bugs become nondeterministic).
\end{itemize}

\section{Mini-exercise}
Pick one of the test files below and write a one-paragraph ``contract'' description for it:

\begin{itemize}
  \item \texttt{src/\_\_tests\_\_/mount-dom.test.ts}
  \item \texttt{src/\_\_tests\_\_/patch-dom.test.ts}
  \item \texttt{src/\_\_tests\_\_/destroy-dom.test.ts}
\end{itemize}

Your goal isn't to understand every line.
Your goal is to extract the invariants: what must always be true after mount/patch/destroy.

\section{Online resources}
These sources are helpful for orienting yourself before diving into implementation details:

\begin{itemize}
  \item Virtual DOM as a concept (implementation details differ, but the mental model is useful):
    \href{https://react.dev/learn/render-and-commit}{React docs: Render and Commit}.
  \item DOM node basics (text nodes vs element nodes):
    \href{https://developer.mozilla.org/en-US/docs/Web/API/Node/nodeType}{MDN: Node.nodeType}.
  \item Microtasks vs macrotasks (relevant for \texttt{nextTick()} / \texttt{queueMicrotask}):
    \href{https://developer.mozilla.org/en-US/docs/Web/API/queueMicrotask}{MDN: queueMicrotask}.
\end{itemize}

\section{Walkthrough: patching is ``same element, new values''}
Relax's VDOM runtime is built around a simple principle:

\begin{quote}
If the old and new nodes are the ``same kind'' of node, we reuse the existing DOM element and update attributes/children.
If they are different, we replace.
\end{quote}

The test suite encodes this principle over and over.
For example, \texttt{src/\_\_tests\_\_/patch-dom.test.ts} starts with extremely small cases:

\begin{itemize}
  \item \textbf{no change}: patching identical trees should reuse \texttt{el}.
  \item \textbf{change the root node}: patching \texttt{div -> span} replaces the root element.
  \item \textbf{patch text}: \texttt{"foo" -> "bar"} updates only the Text node.
\end{itemize}

Those tests tell you what to look for when reading \texttt{src/patch-dom.ts}: element identity reuse is a correctness requirement, not an optimization.

\section{Mini-exercise (with intent)}
Before moving on, do this intentionally (don't just skim):

\begin{enumerate}
  \item Read \texttt{src/app.ts} and answer: what runtime calls does \texttt{createApp()} delegate to?
  \item Read \texttt{src/\_\_tests\_\_/app.test.ts} and find every \texttt{await nextTick()}.
    For each one, explain what DOM change it is waiting for.
  \item Read the first ~5 tests in \texttt{src/\_\_tests\_\_/patch-dom.test.ts}.
    For each, write a one-sentence rule (e.g., ``patching text updates textContent without replacing parent elements'').
\end{enumerate}

\section{Online resources (optional, but good context)}
This book is grounded in the Relax.js repository, but it helps to know the broader ecosystem it draws from.
These are references you can read alongside chapters; they are not prerequisites.

\begin{itemize}
  \item Virtual DOM background (high-level):
  \href{https://react.dev/learn/render-and-commit}{React --- Render and Commit}.
  \item The diffing strategy commonly used in VDOM children reconciliation:
    \href{https://neil.fraser.name/writing/diff/}{Neil Fraser, ``An O(ND) Difference Algorithm and Its Variations''}.
  \item Fine-grained reactivity conceptual background (signals):
    \href{https://www.solidjs.com/docs/latest#signals}{SolidJS docs: Signals}.
  \item Scheduling and keeping UIs responsive:
    \href{https://developer.mozilla.org/en-US/docs/Web/API/window/requestAnimationFrame}{MDN: requestAnimationFrame},
    \href{https://developer.mozilla.org/en-US/docs/Web/API/MessageChannel}{MDN: MessageChannel}.
  \item Testing DOM code in Node:
    \href{https://vitest.dev/guide/}{Vitest Guide},
    \href{https://github.com/jsdom/jsdom}{jsdom}. 
\end{itemize}
