\chapter{Implementation map \& constraints}

\section{What you'll build}
By the end of this chapter, you will have:

\begin{itemize}
  \item A system-level map of the Relax.js repository (what subsystems exist and why).
  \item A repeatable workflow to go from a question \textrightarrow tests \textrightarrow implementation.
  \item One complete mental trace of a user interaction from DOM event \textrightarrow component method \textrightarrow DOM patch.
\end{itemize}

\section{What you'll learn}
\begin{itemize}
  \item How the system splits into three cooperating subsystems (VDOM runtime, HRBR runtime, compiler).
  \item Why jsdom constraints make the runtime \emph{more} deterministic and teachable.
  \item The smallest set of files you should read to understand an end-to-end update.
\end{itemize}

\section{Prereqs}
Read Chapters~1--2.
You should be comfortable with the idea that DOM nodes have identity and can be mutated.

\section{Concept: read the system before you read the code}
When you're learning a new codebase, the worst mistake is to start at \texttt{src/index.ts} and keep scrolling.
You'll drown in details.

Instead, I want you to start with two maps:

\begin{itemize}
  \item a \textbf{module map} (what subsystems exist)
  \item a \textbf{causal map} (what happens when a user clicks / types)
\end{itemize}

This chapter gives you both.

\section{What you're looking at}
This codebase is intentionally small, but it isn't a single module.
It is a collection of cooperating subsystems, each with its own contract and its own tests.

At a high level, think of Relax.js as three subsystems that share a single correctness story:

\begin{enumerate}
  \item \textbf{Relax VDOM runtime} (the teaching-friendly baseline) under \texttt{src/}.
  \item \textbf{HRBR runtime} (experimental, compiled fast-path) under \texttt{runtime/}.
  \item \textbf{Compiler} (experimental transform from TSX/JSX subset to HRBR blocks) under \texttt{compiler/}.
\end{enumerate}

\subsection*{Diagram: repository architecture at a glance}
\begin{center}
\begin{tikzpicture}[
  node distance=10mm and 12mm,
  box/.style={draw, rounded corners, align=left, inner sep=6pt, fill=RelaxLight},
  arrow/.style={-Latex, thick}
]
\node[box] (src) {\textbf{VDOM runtime}\\\texttt{src/*}\\mount/patch/destroy\\components, events};
\node[box, right=of src] (tests) {\textbf{Tests-as-spec}\\\texttt{src/\_\_tests\_\_/*}\\\texttt{runtime/\_\_tests\_\_/*}\\\texttt{compiler/\_\_tests\_\_/*}};
\node[box, below=of src] (runtime) {\textbf{HRBR runtime}\\\texttt{runtime/*}\\signals, scheduler\\blocks, hydration};
\node[box, below=of runtime] (compiler) {\textbf{Compiler}\\\texttt{compiler/*}\\JSX\,$\rightarrow$\,BlockDef};
\node[box, right=of runtime] (bench) {\textbf{Benchmarks + docs}\\\texttt{benchmarks/*}\\\texttt{docs/*}};

\draw[arrow] (src) -- node[midway, above] {verified by} (tests);
\draw[arrow] (runtime) -- (tests);
\draw[arrow] (compiler) -- (tests);

\draw[arrow] (compiler) -- node[midway, left] {emits defs consumed by} (runtime);
\draw[arrow] (src) -- node[midway, left] {can embed} (runtime);
\draw[arrow] (bench) -- (runtime);
\draw[arrow] (bench) -- (src);
\end{tikzpicture}
\end{center}

The key idea: tests are the stable external contract, while the implementation is split into a flexible VDOM baseline and an optional compiled runtime.

Everything else exists to support one of these (docs, benchmarks, build tooling).

\subsection{Directory map (first pass)}
The codebase is split into a few major areas:

\begin{itemize}
  \item \texttt{src/}: the Relax.js Virtual DOM runtime.
  \item \texttt{runtime/}: an experimental runtime (HRBR) based on signals, scheduling, and compiled blocks.
  \item \texttt{compiler/}: a compiler/transform that converts a JSX subset into HRBR block definitions.
  \item \texttt{benchmarks/}: a browser harness to compare performance.
  \item \texttt{docs/}: design notes and whitepapers.
\end{itemize}

Two more folders matter when you're thinking about correctness:

\begin{itemize}
  \item \texttt{src/\_\_tests\_\_/}: the VDOM runtime spec.
  \item \texttt{runtime/\_\_tests\_\_/} and \texttt{compiler/\_\_tests\_\_/}: the HRBR + compiler spec.
\end{itemize}

\section{The constraint that shapes everything: DOM correctness in jsdom}
Almost every implementation choice is constrained by one hard rule:

\begin{quote}
The runtime must be correct in a headless DOM (jsdom), and tests are the contract.
\end{quote}

That affects:

\begin{itemize}
  \item Attribute vs property behavior (inputs/selects are tricky in jsdom).
  \item Event binding and cleanup.
  \item Diff algorithms (keyed moves must preserve node identity).
\end{itemize}

\subsection{Why jsdom is a meaningful constraint}
jsdom is \emph{not} a full browser, and that is exactly why it is useful here.
If your runtime depends on unspecified browser behavior or timing quirks, your test suite will become flaky.

Instead, tests in this repo focus on:

\begin{itemize}
  \item DOM structure (\texttt{innerHTML})
  \item identity (whether an element is reused or replaced)
  \item cleanup (event listeners and references are removed)
\end{itemize}

This shapes file boundaries: many helpers exist to make DOM updates deterministic and assertable.

\section{How to read code in this repo (recommended path)}
When you see a source file, immediately find its tests. That's the spec.

Recommended reading order:

\begin{enumerate}
  \item \texttt{src/index.ts} (public exports)
  \item \texttt{src/h.ts} (VNode model)
  \item \texttt{src/mount-dom.ts} (mount)
  \item \texttt{src/patch-dom.ts} (patch)
  \item \texttt{src/destroy-dom.ts} (cleanup)
  \item \texttt{src/component.ts} (components)
\end{enumerate}

\subsection{A second reading path: follow a user action}
The fastest way to build intuitions is to follow a single user interaction end-to-end.
For example: ``typing in an input and clicking a button adds a todo''.

You can trace that across:

\begin{itemize}
  \item event normalization: \texttt{src/events.ts}
  \item component event binding: \texttt{src/dispatcher.ts} and \texttt{src/component.ts}
  \item scheduling: \texttt{src/scheduler.ts} (\texttt{nextTick()})
  \item patching: \texttt{src/patch-dom.ts}
\end{itemize}

Doing this with the tests open beside you is the theme of this book.

\section{Build-along: follow one interaction end-to-end}
Now let's do the thing I personally do when I need to understand or debug a UI runtime:
I pick one user interaction and trace it until I can explain every boundary.

We'll use the Todo app test as our "story".
The specification is \texttt{src/\_\_tests\_\_/app.test.ts}.

\subsection{The spec (test) as the narrative}
In that test file, the interesting lines look like this:

\begin{lstlisting}[style=relax,language=JavaScript]
input.value = 'Buy milk'
input.dispatchEvent(new Event('input'))

button.click()

await nextTick()
expect(document.body.innerHTML).toBe(...)
\end{lstlisting}

\textbf{Why this matters:}
this is an executable description of the user story.
The only reason we can trust the runtime is because this story is repeatable.

In the rest of this chapter, I’m going to connect this story to the implementation.

\subsection{Step 1: the root lifecycle boundary (createApp)}
The first system boundary is the application lifecycle.
In Relax, \texttt{createApp()} is intentionally tiny.
Its entire job is to create the root VNode and unmount it.

Implementation: \texttt{src/app.ts}.

\begin{lstlisting}[style=relax,language=JavaScript]
export function createApp(RootComponent, props = {}) {
  let vdom = null

  return {
    mount(parentEl) {
      vdom = h(RootComponent, props)
      mountDOM(vdom, parentEl)
    },
    unmount() {
      destroyDOM(vdom)
      vdom = null
    },
  }
}
\end{lstlisting}

	\textbf{Why this design:}
I want exactly one owner for the root VNode.
If you don't have a clear ownership boundary, "unmount" becomes best-effort cleanup and bugs get subtle.

\subsection{Step 2: DOM event binding (how a handler gets installed)}
When a VNode contains an \texttt{on} map (events), Relax wires listeners during mount.
The low-level helper lives in \texttt{src/events.ts}:

\begin{lstlisting}[style=relax,language=JavaScript]
export function addEventListener(eventName, handler, el, hostComponent = null) {
  const listener = (event) => {
    if (hostComponent) handler.call(hostComponent, event)
    else handler(event)
  }

  el.addEventListener(eventName, listener)
  return listener
}
\end{lstlisting}

	\textbf{Why this matters:}
the listener wrapper is a tiny but crucial semantic guarantee.
It means component methods can rely on \texttt{this} being the component instance.
(We’ll return to this in the events chapter.)

\subsection{Step 3: component-to-component communication (Dispatcher)}
In Relax, components don't reach into each other directly.
They communicate through a tiny command dispatcher: \texttt{src/dispatcher.ts}.

\begin{lstlisting}[style=relax,language=JavaScript]
dispatch(commandName, payload) {
  if (this.#subs.has(commandName)) {
    this.#subs.get(commandName).forEach((handler) => handler(payload))
  }

  this.#afterHandlers.forEach((handler) => handler())
}
\end{lstlisting}

	\textbf{Why this matters:}
this gives me a clean abstraction boundary: event handlers emit events, components react, and then we run after-hooks (often ``schedule a patch'').

\subsection{Step 4: where does the DOM update happen?}
This is the final thing you must always locate:

\begin{quote}
Where is the code that turns a "new desired UI" into DOM mutations?
\end{quote}

In the VDOM runtime, that answer is \texttt{src/patch-dom.ts} (covered in depth later).
In this chapter, the important idea is simply: \emph{there is exactly one patching engine}, and everything funnels into it.

\section{A concrete example: createApp() is tiny on purpose}
\texttt{createApp()} exists so the user has a simple lifecycle API.
The full file is short: \texttt{src/app.ts}.

Key behavior:

\begin{itemize}
  \item It creates a root VNode with \texttt{h(RootComponent, props)}.
  \item It mounts it once.
  \item It can unmount and fully cleanup.
\end{itemize}

\subsection{Code reference}
See \texttt{src/app.ts}. The central calls are \texttt{mountDOM(...)} and \texttt{destroyDOM(...)}.

\subsection{Test reference (what proves the contract)}
The high-level application lifecycle is specified by \texttt{src/\_\_tests\_\_/app.test.ts}.
This test file shows a full mount-to-interaction flow:

\begin{itemize}
  \item It mounts the app.
  \item It waits for async work with \texttt{await nextTick()}.
  \item It simulates user input by setting \texttt{input.value} and dispatching an \texttt{input} event.
  \item It simulates clicks via \texttt{button.click()}.
  \item It asserts DOM structure with \verb|singleHtmlLine`...`|.
\end{itemize}

That single test is a great ``reading harness'' for learning how Relax wires components + events + patching.

\section{Behavioral contract (how to read tests)}
If you only remember one workflow from this chapter, make it this:

\begin{enumerate}
  \item Start from a test that matches the behavior you care about.
  \item Read the assertions and treat them as the contract.
  \item Jump to implementation and locate the code that makes those assertions true.
  \item If you're changing behavior, change the test first.
\end{enumerate}

Concrete anchors for this chapter:

\begin{itemize}
  \item App lifecycle story: \texttt{src/\_\_tests\_\_/app.test.ts}
  \item Event semantics: \texttt{src/\_\_tests\_\_/events.test.ts}
  \item Dispatcher semantics: \texttt{src/\_\_tests\_\_/dispatcher.test.ts}
\end{itemize}

\section{Common pitfalls}
\begin{itemize}
  \item Reading implementation first and assuming the behavior is intentional.
  \item Ignoring jsdom differences (inputs/selection are special).
  \item Treating "unmount" as optional cleanup.
  \item Not understanding what is synchronous vs deferred to \texttt{nextTick()}.
\end{itemize}

\section{Mini-exercise}
Open \texttt{src/\_\_tests\_\_/app.test.ts} and answer:

\begin{itemize}
  \item Why does the test call \texttt{await nextTick()} after mount?
  \item Which user actions are simulated (exact DOM APIs)?
  \item Which pieces of the runtime must exist for that simulation to work (mounting, events, patching)?
\end{itemize}

\section{Project constraint: small API surface}
Relax exports only a few primitives (h, fragments, components, scheduler).
Everything else is internal. That keeps the library small and also makes the tests more important.

\subsection{Where to see the supported surface area}
The public exports are in \texttt{src/index.ts} and mirrored by the \texttt{exports} map in \texttt{package.json}.

In other words:

\begin{itemize}
  \item if it's not in \texttt{src/index.ts}, it's not a stable public API
  \item tests are free to depend on internals, but users shouldn't
\end{itemize}

This book leans on that design: we can explain the internals without promising them as user-facing contracts.

\section{Tests: how this repo measures correctness}
This repository uses Vitest with a jsdom environment.
The configuration lives in \texttt{vitest.config.js}:

\begin{itemize}
  \item test patterns: \texttt{**/\_\_tests\_\_/**/*.test.ts(x)}
  \item reporters: verbose (good for learning)
  \item environment: jsdom
\end{itemize}

\subsection{The single most important helper: singleHtmlLine}
Many tests assert against HTML strings.
If you try to compare raw multi-line template literals to \texttt{innerHTML}, whitespace will fight you.

This repo solves that in \texttt{src/\_\_tests\_\_/utils.ts} with \texttt{singleHtmlLine()}:

\begin{itemize}
  \item it collapses whitespace
  \item it removes spaces between tags (for example, \texttt{>\textbackslash s+<} becomes \texttt{><})
  \item it trims
\end{itemize}

This is a micro-lesson in test design: you want assertions that are strict about structure, but forgiving about irrelevant formatting.

\subsection{Where to learn semantics fastest (test tour)}
If you're new to the codebase, these tests teach the contracts in the right order:

\begin{itemize}
  \item \texttt{src/\_\_tests\_\_/mount-dom.test.ts}:
    mount behavior, fragment behavior, saving \texttt{vdom.el} references.
  \item \texttt{src/\_\_tests\_\_/patch-dom.test.ts}:
    identity reuse vs replacement, attribute updates, keyed children, and edge cases.
  \item \texttt{src/\_\_tests\_\_/destroy-dom.test.ts}:
    cleanup and idempotency, especially ``remove event listeners''.
  \item \texttt{src/\_\_tests\_\_/component.test.ts}:
    component lifecycle, props/state updates, and event binding to the host component.
  \item \texttt{src/\_\_tests\_\_/jsx-react-syntax.test.tsx}:
    TSX compatibility contract (className/onClick/fragments/components).
\end{itemize}

Each of these will be revisited later in the book when we dive into the implementation files they protect.

\section{Mini-exercise}
Open these tests and write down what they prove (one sentence each). Then locate the implementation file they exercise.

\begin{itemize}
  \item \texttt{src/\_\_tests\_\_/app.test.ts}
  \item \texttt{src/\_\_tests\_\_/component.test.ts}
  \item \texttt{src/\_\_tests\_\_/mount-dom.test.ts}
  \item \texttt{src/\_\_tests\_\_/destroy-dom.test.ts}
\end{itemize}

If you can answer these, you're ready for Part II.

\section{Diagram: a "follow the tests" reading workflow}
When you don't know where to start, start from the tests and work backward to the implementation.

\begin{center}
\begin{tikzpicture}[
  node distance=9mm and 12mm,
  box/.style={draw, rounded corners, align=center, inner sep=6pt, fill=RelaxLight},
  arrow/.style={-Latex, thick}
]
  \node[box] (fail) {A behavior question\\("What should happen?")};
  \node[box, right=of fail] (test) {Find a test\\\texttt{src/\_\_tests\_\_/mount-dom.test.ts}};
  \node[box, right=of test] (assert) {Read assertions\\(DOM structure/identity)};
  \node[box, below=of assert] (impl) {Jump to implementation\\\texttt{src/*} / \texttt{runtime/*}};
  \node[box, left=of impl] (confirm) {Confirm contract\\(invariants, edge cases)};

  \draw[arrow] (fail) -- (test);
  \draw[arrow] (test) -- (assert);
  \draw[arrow] (assert) -- (impl);
  \draw[arrow] (impl) -- (confirm);
  \draw[arrow] (confirm) -- (test);
\end{tikzpicture}
\end{center}

This loop is also how you safely add features: you extend the spec (tests) and then evolve the implementation.

\section{Online resources}
This chapter is about reading \emph{this repo}, but a few external references help explain why these constraints and structures are common.

\begin{itemize}
  \item DOM living standard (event/bubbling/capturing model):
    \href{https://dom.spec.whatwg.org/}{WHATWG DOM Standard}.
  \item Event handling basics and differences between properties vs attributes:
    \href{https://developer.mozilla.org/en-US/docs/Web/API/EventTarget/addEventListener}{MDN: addEventListener},
    \href{https://developer.mozilla.org/en-US/docs/Web/API/HTMLElement}{MDN: HTMLElement}.
  \item Testing DOM in Node:
    \href{https://vitest.dev/guide/}{Vitest Guide},
    \href{https://github.com/jsdom/jsdom}{jsdom}.
  \item Module packaging and exports maps:
    \href{https://nodejs.org/api/packages.html#packages_exports}{Node.js docs: \texttt{exports}}.
\end{itemize}

For formal citations, you can map key links into the backmatter \texttt{References} list in \texttt{book/main.tex}.
