\chapter{Destroying DOM and Unmounting Components}

So far we've treated the renderer as a one-way pipeline:

\begin{center}
	texttt{VNode \textrightarrow mountDOM() \textrightarrow DOM \textrightarrow patchDOM()}
\end{center}

But a real UI runtime lives and dies by teardown. If teardown is incomplete or inconsistent, it manifests as:

\begin{itemize}
  \item \textbf{leaked event listeners} (handlers still firing after nodes are removed)
  \item \textbf{leaked component instances} (dispatch subscriptions never removed)
  \item \textbf{zombie DOM references} (stale \texttt{vdom.el} pointers used accidentally)
  \item \textbf{interop leaks} (third-party instances never disposed)
\end{itemize}

\section{Diagram: render lifecycle as a loop (mount \textrightarrow patch \textrightarrow destroy)}
It's useful to stop thinking of rendering as a straight line and instead as a loop that eventually ends in teardown:

\begin{center}
\begin{tikzpicture}[
  node distance=10mm,
  box/.style={draw,rounded corners,fill=RelaxLight,inner sep=5pt,align=center},
  arrow/.style={-Latex,thick}
]
  \node[box] (vdom) {VNode tree\\(created by \texttt{h()})};
  \node[box, right=of vdom] (mount) {\texttt{mountDOM}\\allocates DOM + listeners};
  \node[box, below=of mount] (patch) {\texttt{patchDOM}\\updates in place (when safe)};
  \node[box, left=of patch] (destroy) {\texttt{destroyDOM}\\teardown + delete refs};

  \draw[arrow] (vdom) -- (mount);
  \draw[arrow] (mount) -- (patch);
  \draw[arrow] (patch) -- (destroy);
  \draw[arrow] (destroy) -- (vdom);
\end{tikzpicture}
\end{center}

This chapter focuses on the \emph{only irreversible edge} in that loop: once you've destroyed, the old VNode tree should no longer be considered usable.

Relax centralizes teardown in \texttt{destroyDOM(vdom)} in \texttt{src/destroy-dom.ts}. The executable specs are primarily in:

\begin{itemize}
  \item \texttt{src/\_\_tests\_\_/destroy-dom.test.ts}
  \item \texttt{src/\_\_tests\_\_/component-lifecycle.test.ts}
  \item \texttt{src/\_\_tests\_\_/hrbr-integration.test.ts} (interop)
\end{itemize}

\section{A tiny contract for teardown}

We'll use a tight contract so you can re-derive the implementation from the tests:

\begin{itemize}
  \item \textbf{Input}: a \emph{mounted} VNode subtree root.
  \item \textbf{Effects}: remove DOM nodes, dispatch cleanup, interop cleanup, and clear runtime references.
  \item \textbf{Idempotence}: not guaranteed. Calling destroy twice is a programmer error.
  \item \textbf{Success criteria}: after teardown, the DOM no longer contains the rendered subtree \emph{and} the VNode tree no longer retains \texttt{.el} pointers.
\end{itemize}

That last bullet is important: removing DOM nodes is easy---cleaning the runtime state is the hard part.

\subsection{Two different things you must destroy (from scratch)}

When people say \emph{``unmount''} or \emph{``destroy''}, they're often mixing two separate jobs:

\begin{enumerate}
  \item \textbf{Destroy the DOM nodes} (the browser objects).
  \item \textbf{Destroy the runtime graph} (your VNodes, component instances, subscriptions, and cached references).
\end{enumerate}

In Relax, the first job is the visible one: the subtree disappears from the document.
The second job is the reliability one: you prevent memory leaks and you prevent \emph{use-after-destroy} bugs (where stale VNodes are accidentally reused).

This is why the contract mentions both:\ the DOM must no longer contain the subtree, \emph{and} the old VNode tree must no longer retain \texttt{.el} pointers.

\paragraph{Why this matters in practice}

Even when an element has been removed from the document, JavaScript can still hold a reference to it.
If a VNode still points at a removed element (\texttt{vdom.el}), you can accidentally:

\begin{itemize}
  \item call DOM APIs on a node you believe is ``gone''
  \item keep a whole subtree alive in memory because some object graph still references it
  \item keep event listeners alive (the browser keeps a reference from the element to the listener function)
\end{itemize}

For browser reference on why listener identity matters (same function object to remove), see MDN:\ \href{https://developer.mozilla.org/en-US/docs/Web/API/EventTarget/removeEventListener}{\texttt{removeEventListener()}}.

\section{destroyDOM(): the teardown dispatcher}

The public entry point is:\ \texttt{destroyDOM(vdom)}.

The implementation is built as a type-dispatcher over \texttt{vdom.type} (from \texttt{DOM\_TYPES} in \texttt{src/h.ts}).

\subsection*{Diagram: destroy dispatcher}
The file \texttt{src/destroy-dom.ts} mirrors \texttt{src/mount-dom.ts}: it is also a type-dispatcher.

\begin{center}
\begin{tikzpicture}[
  node distance=8mm,
  box/.style={draw,rounded corners,fill=RelaxLight,inner sep=5pt,align=left},
  arrow/.style={-Latex,thick}
]
  \node[box] (in) {\textbf{input:} mounted \texttt{vdom}};
  \node[box, below=of in] (sw) {\textbf{switch} (\texttt{vdom.type})};

  \node[box, below left=of sw, xshift=14mm] (text) {TEXT\\remove text node\\\texttt{el.remove()}};
  \node[box, below=of sw] (elem) {ELEMENT\\remove element\\remove listeners\\recurse children};
  \node[box, below right=of sw, xshift=-14mm] (frag) {FRAGMENT\\recurse children};
  \node[box, below=of elem, yshift=-2mm] (comp) {COMPONENT\\\texttt{component.unmount()}\\schedule \texttt{onUnmounted}};
  \node[box, below=of frag, yshift=-2mm] (hrbr) {HRBR\\dispose/destroy\\remove host};

  \node[box, below=18mm of comp] (out) {\textbf{post-step:} \texttt{delete vdom.el}\\(+ delete other refs)};

  \draw[arrow] (in) -- (sw);
  \draw[arrow] (sw) -- (text);
  \draw[arrow] (sw) -- (elem);
  \draw[arrow] (sw) -- (frag);
  \draw[arrow] (elem) -- (comp);
  \draw[arrow] (frag) -- (hrbr);
  \draw[arrow] (comp) -- (out);
\end{tikzpicture}
\end{center}

\subsection{The shared post-step: deleting \texttt{vdom.el}}

At the end of \texttt{destroyDOM}, Relax does:

\begin{verbatim}
delete vdom.el
\end{verbatim}

This isn't cosmetic.

\begin{itemize}
  \item It enforces a strong invariant in tests (``every destroyed node lost its \texttt{.el}'').
  \item It makes stale VNodes fail fast if reused accidentally.
  \item It reduces accidental retention of DOM nodes by application code holding onto VNodes.
\end{itemize}

\subsection{A quick mental trace: what gets touched?}

If you want a simple way to sanity-check teardown, trace what \texttt{destroyDOM(vdom)} touches:

\begin{itemize}
  \item the browser objects (DOM nodes and event listeners)
  \item the VNode object graph (delete \texttt{.el} and other stored pointers)
  \item the component graph (call \texttt{unmount()} and clear subscriptions)
  \item the scheduler queue (lifecycle hooks are queued, not executed inline)
\end{itemize}

If any one of those is skipped, teardown can look "fine" in the DOM while still leaking runtime state.

\section{Type-based teardown}

\texttt{destroyDOM} dispatches on \texttt{vdom.type} (\texttt{DOM\_TYPES} is defined in \texttt{src/h.ts}).

\subsection{Text nodes}

For \texttt{DOM\_TYPES.TEXT}, \texttt{removeTextNode} asserts \texttt{el instanceof Text} and calls \texttt{el.remove()}.

Test: \texttt{destroy a text element}.

\subsection{Element nodes: remove element + recurse + listener cleanup}

For \texttt{DOM\_TYPES.ELEMENT}, \texttt{removeElementNode} performs three distinct actions:

\begin{enumerate}
  \item Remove the element from the DOM: \texttt{el.remove()}.
  \item Recursively destroy children: \texttt{children.forEach(destroyDOM)}.
  \item If there are DOM event listeners stored on the VNode, detach them:
    \begin{verbatim}
removeEventListeners(listeners, el)
delete vdom.listeners
    \end{verbatim}
\end{enumerate}

The listener removal uses \texttt{removeEventListeners} from \texttt{src/events.ts}, which removes by the wrapper listener identity that was created during mount/patch.

Tests:

\begin{itemize}
  \item \texttt{destroy an html element and its children}
  \item \texttt{destroy an html element and its children recursively}
  \item \texttt{remove an html element event listeners}
\end{itemize}

\paragraph{Ordering note: remove first, then cleanup}

	texttt{removeElementNode()} removes the element from the DOM \emph{before} walking children.
This ordering is deliberate and safe:

\begin{itemize}
  \item In the DOM, removing a parent implicitly detaches descendants from the document.
  \item But Relax still needs to recursively teardown the VNode subtree (delete \texttt{.el}, detach listeners, unmount nested components).
  \item Running the recursion after removal avoids intermediate states where logic could observe partially-destroyed nodes in the live document.
\end{itemize}

\subsubsection{A short code reading (as JavaScript)}

Here is the important shape of \texttt{removeElementNode()} from \texttt{src/destroy-dom.ts}.
Notice how it does three things: DOM removal, recursive teardown, and listener cleanup.

\begin{lstlisting}[style=relax,language=JavaScript]
function removeElementNode(vdom) {
  const { el, children, listeners } = vdom

  el.remove()
  children.forEach(destroyDOM)

  if (listeners) {
    removeEventListeners(listeners, el)
    delete vdom.listeners
  }
}
\end{lstlisting}

Two details to internalize:

\begin{itemize}
  \item \textbf{The recursion is about runtime cleanup, not document removal.} The browser already detached child nodes when the parent was removed.
  \item \textbf{Listener cleanup depends on stored wrapper identity.} You must remove \emph{the exact function object} you added.
\end{itemize}

\subsection{Fragment nodes}

Fragments are a logical grouping; they don't own a single DOM node. For \texttt{DOM\_TYPES.FRAGMENT}, \texttt{removeFragmentNodes} just destroys each child.

Tests:

\begin{itemize}
  \item \texttt{destroy a fragment}
  \item \texttt{destroy a fragment recursively}
\end{itemize}

\subsection{Component nodes: unmount + schedule \texttt{onUnmounted}}

For \texttt{DOM\_TYPES.COMPONENT}, \texttt{destroyDOM} delegates to the component instance:

\begin{verbatim}
vdom.component.unmount()
enqueueJob(() => vdom.component.onUnmounted())
\end{verbatim}

Two important details are embedded here, and both are tested:

\begin{itemize}
  \item \textbf{Unmount is synchronous structural cleanup.} It destroys the component's rendered subtree and unsubscribes internal dispatcher subscriptions.
  \item \textbf{\texttt{onUnmounted} is scheduled.} It runs later via \texttt{enqueueJob} (\texttt{src/scheduler.ts}), matching the mount side where \texttt{onMounted} is also scheduled.
\end{itemize}

This scheduling is why the component lifecycle tests await \texttt{nextTick()} after calling \texttt{destroyDOM()}:

\begin{itemize}
  \item \texttt{a component reacts to being unmounted}
  \item \texttt{a component reacts to being unmounted asynchronously}
\end{itemize}

\subsection*{Diagram: component unmount is two-phase}
Component teardown is deliberately split into "structural cleanup now" and "user hook later":

\begin{center}
\begin{tikzpicture}[
  node distance=9mm,
  box/.style={draw,rounded corners,fill=RelaxLight,inner sep=5pt,align=left},
  arrow/.style={-Latex,thick}
]
  \node[box] (destroy) {\texttt{destroyDOM(componentVdom)}};
  \node[box, below=of destroy] (unmount) {\textbf{sync:} \texttt{component.unmount()}\\- destroys subtree\\- unsubscribes dispatcher\\- clears fields};
  \node[box, below=of unmount] (enqueue) {\textbf{async:} \texttt{enqueueJob(() => component.onUnmounted())}};
  \node[box, below=of enqueue] (tick) {\texttt{nextTick()} boundary\\(hooks are \emph{scheduled}, not inline)};

  \draw[arrow] (destroy) -- (unmount);
  \draw[arrow] (unmount) -- (enqueue);
  \draw[arrow] (enqueue) -- (tick);
\end{tikzpicture}
\end{center}

This pattern matches mount-side semantics too (\texttt{onMounted} is also scheduled). The intent is: the renderer finishes DOM surgery first, then user hooks run without re-entrancy hazards.

\subsubsection{From scratch: why is \texttt{onUnmounted} scheduled?}

Scheduling isn't just an implementation detail; it's an API promise.
If user hooks ran inline during\ \texttt{destroyDOM()}:

\begin{itemize}
  \item hooks could observe a partially-destroyed tree (some nodes already removed, others not yet)
  \item hooks could trigger state updates that attempt to patch while teardown is mid-flight
  \item hooks could throw errors that leave teardown half-complete
\end{itemize}

By deferring hooks to the scheduler queue (see \texttt{enqueueJob} in \texttt{src/scheduler.ts}), Relax makes teardown an atomic-ish structural operation: first it makes the DOM and runtime graph consistent, \emph{then} it runs user code.

For background on the reasoning, React's commit phase + effect cleanup is a helpful conceptual parallel:\ \href{https://react.dev/reference/react/useEffect#effects-and-cleanup}{Effects and cleanup}.

\section{Inside Component.unmount() (\texttt{src/component.ts})}

Components in Relax are created with \texttt{defineComponent(...)}. The generated class maintains several private fields that matter for unmount:

\begin{itemize}
  \item \texttt{\#vdom}: the last rendered subtree returned by \texttt{render()}
  \item \texttt{\#dispatcher}: a per-component \texttt{Dispatcher} used to implement \texttt{emit()} and event handler wiring
  \item \texttt{\#subscriptions}: unsubscribe closures returned from \texttt{dispatcher.subscribe()}
  \item \texttt{\#isMounted} and \texttt{\#hostEl}: mount state and mount container
\end{itemize}

\subsection{Unmount sequence}

Look at \texttt{unmount()} in \texttt{src/component.ts}:

\begin{enumerate}
  \item It throws if the component isn't mounted.
  \item It destroys the rendered subtree: \texttt{destroyDOM(this.\#vdom)}.
  \item It unsubscribes from internal dispatcher subscriptions: \texttt{this.\#subscriptions.forEach(unsubscribe)}.
  \item It clears state: \texttt{\#vdom = null}, \texttt{\#isMounted = false}, \texttt{\#hostEl = null}, \texttt{\#subscriptions = []}.
\end{enumerate}

This means after unmount, both DOM and book-keeping references are gone.

\subsection{Unmount in one annotated snippet (as JavaScript)}

Here is the sequence from \texttt{src/component.ts}, rewritten in JavaScript style so you can read it as control flow:

\begin{lstlisting}[style=relax,language=JavaScript]
unmount() {
  if (!this.#isMounted) {
    throw new Error('Component is not mounted')
  }

  // 1) structural teardown of rendered subtree
  destroyDOM(this.#vdom)

  // 2) detach component-owned subscriptions (not VNode-owned)
  this.#subscriptions.forEach((unsubscribe) => unsubscribe())

  // 3) clear references so GC can reclaim memory and stale use fails fast
  this.#vdom = null
  this.#isMounted = false
  this.#hostEl = null
  this.#subscriptions = []
}
\end{lstlisting}

The separation between (1) and (2) is the same architectural split as earlier chapters:
  	exttt{destroyDOM()} is VNode-centric, while \texttt{unmount()} is component-centric.

\paragraph{Why unsubscribe in \texttt{unmount()} and not in \texttt{destroyDOM()}?}

Because subscriptions are a property of the component instance, not a property of the rendered VDOM nodes.
    exttt{destroyDOM()} is intentionally ``VNode-centric''; \texttt{unmount()} is ``component-centric''.

\section{HRBR nodes: dispose, destroy, and remove host}

Relax also supports an experimental HRBR interop node type (\texttt{DOM\_TYPES.HRBR}).
These nodes represent ``something else'' mounting into a host container (a block, micro-framework component, etc.).

In \texttt{removeHrbrNode(vdom)} (\texttt{src/destroy-dom.ts}):

\begin{itemize}
  \item If the instance provides \texttt{dispose()} and/or \texttt{destroy()}, Relax calls them (in that order).
  \item Then it removes the HRBR host element from the DOM.
  \item Finally, it deletes \texttt{vdom.instance} and \texttt{vdom.host}.
\end{itemize}

\subsection{From scratch: HRBR is about \emph{ownership boundaries}}

HRBR nodes exist for one reason: sometimes the thing you're mounting isn't a Relax VNode subtree.
It might be a compiled template block, a micro-framework widget, or a third-party imperative library.

To make teardown predictable, you need a rule for \textbf{who owns what}:

\begin{itemize}
  \item \textbf{Relax owns the host container} (a DOM element that acts as a mounting surface).
  \item \textbf{The interop system owns the instance} (whatever object the mount function returns).
  \item \textbf{Relax promises a small teardown protocol:}
    call optional \texttt{dispose()}, then optional \texttt{destroy()}, then remove the host.
\end{itemize}

This is very similar to a resource-management pattern in lower-level systems: \emph{if you allocate it, you must free it}.
The difference is that in UI you often split responsibility: Relax allocates the DOM ``socket'' (host), while the interop layer allocates the instance.

\subsection*{Diagram: HRBR ownership + lifecycle (mount, patch, destroy)}

The easiest way to think about HRBR is: Relax plugs an \emph{opaque instance} into a \emph{host container}, and then manages the host's lifetime.
The instance itself is treated like a resource with optional cleanup hooks.

\begin{center}
\begin{tikzpicture}[
  node distance=8mm,
  box/.style={draw,rounded corners,fill=RelaxLight,inner sep=5pt,align=left},
  arrow/.style={-Latex,thick}
]
  \node[box] (vnode) {HRBR VNode\\\texttt{type = DOM\_TYPES.HRBR}\\\texttt{mountFactory}\ (function identity)};

  \node[box, below=of vnode] (mount) {\textbf{mount:}\\Relax creates \texttt{host}\ (owned container)\\calls \texttt{mountFactory(host)}\\stores \texttt{vdom.instance}};

  \node[box, below=of mount] (patch) {\textbf{patch:} compare \texttt{mountFactory} identity\\\textbf{same} \textrightarrow reuse instance\\\textbf{changed} \textrightarrow dispose+destroy old\\then mount new};

  \node[box, below=of patch] (destroy) {\textbf{destroy:}\\if \texttt{instance.dispose} call it\\if \texttt{instance.destroy} call it\\remove \texttt{host}\\delete \texttt{instance/host refs}};

  \draw[arrow] (vnode) -- (mount);
  \draw[arrow] (mount) -- (patch);
  \draw[arrow] (patch) -- (destroy);
\end{tikzpicture}
\end{center}

The test suite drives this design: HRBR patching is intentionally based on \emph{factory identity} (stable identity enables reuse; changing identity forces remount).

\subsection{The destroy sequence (as JavaScript)}

Here is \texttt{removeHrbrNode()} from \texttt{src/destroy-dom.ts}, rewritten into JavaScript style so you can focus on the protocol:

\begin{lstlisting}[style=relax,language=JavaScript]
function removeHrbrNode(vdom) {
  const inst = vdom.instance
  if (inst?.dispose) inst.dispose()
  if (inst?.destroy) inst.destroy()

  const host = vdom.host
  if (host) host.remove()

  delete vdom.instance
  delete vdom.host
}
\end{lstlisting}

This ordering is deliberate:

\begin{itemize}
  \item \textbf{Call hooks before removing the host} so the instance can still query/update the DOM it owns.
  \item \textbf{Remove the host last} so Relax reliably removes the container regardless of how well-behaved the instance is.
\end{itemize}

For background on DOM node removal semantics, see MDN:\ \href{https://developer.mozilla.org/en-US/docs/Web/API/Element/remove}{\texttt{Element.remove()}}.

This mirrors the mounting/patching behavior from earlier chapters: HRBR nodes are treated like a mounted instance with an owned host container.

\paragraph{Interop rule of thumb}

Relax doesn't assume what ``destroy'' means.
It only promises it will call well-known lifecycle hooks (\texttt{dispose}/\texttt{destroy}) if present, and remove the host container it created.

\subsection{How HRBR teardown relates to patching (what the tests prove)}

The HRBR integration tests (\texttt{src/\_\_tests\_\_/hrbr-integration.test.ts}) implicitly define a second contract:\ how HRBR behaves under \emph{patch}, not just under destroy.

\begin{itemize}
  \item \textbf{Stable mount function identity means reuse.}
    If a component re-renders and returns an HRBR node created with the \emph{same} mount function, Relax preserves the existing instance.
  \item \textbf{Changed mount function identity means remount.}
    If the mount function identity changes, Relax must treat it like a different interop node: it tears down the old instance (\texttt{dispose}/\texttt{destroy}) and mounts a new one.
\end{itemize}

Why identity? Because it's the simplest safe signal that the meaning of ``this interop node'' changed.
If Relax tried to "diff" two opaque interop instances, it would be forced to guess.
So the system picks a clean architectural line: \emph{either you're the same factory (reuse), or you're a new factory (remount)}.

This is the same philosophy you saw in keyed VDOM reconciliation earlier: stable identity enables preservation.

For a conceptual parallel in mainstream UI libraries, see React's note on identity and preserving state across renders (keys and element identity):\
\href{https://react.dev/learn/rendering-lists#keeping-list-items-in-order-with-key}{Keeping list items in order with \texttt{key}}.

\section{A practical mental model}

It helps to treat teardown as the inverse of mounting:

\begin{itemize}
  \item mounting creates DOM nodes and stores refs in \texttt{vdom.el}
  \item patching preserves DOM nodes and updates refs on the new VDOM
  \item destroying removes DOM nodes and deletes refs from the old VDOM
\end{itemize}

The explicit \texttt{delete vdom.el} makes it hard to accidentally keep using stale VNodes after tear-down.

\subsection{A leak-prevention checklist you can reuse}

When teardown bugs happen in real apps, they usually violate one of these checks.
You can use this list as a debugging worksheet:

\begin{enumerate}
  \item \textbf{DOM nodes removed:} is the subtree gone from the host element?
  \item \textbf{Listener identities removed:} are \texttt{removeEventListener} calls using the exact wrapper that was added?
  \item \textbf{VNode references cleared:} are old \texttt{vdom.el} pointers deleted so stale objects can't keep nodes alive?
  \item \textbf{Component subscriptions cleared:} do per-component subscriptions get unsubscribed on unmount?
  \item \textbf{Interop disposed:} do embedded instances get \texttt{dispose}/\texttt{destroy} and does the host container get removed?
\end{enumerate}

The browser-side references are the subtle part. Even after DOM removal, a reference chain like:
    exttt{appState -> handler -> closure -> element}
can keep things alive. This is why tests focus on listener removal and reference deletion.

\section{Takeaways}

\begin{itemize}
  \item Relax teardown is explicit and recursive.
  \item Element teardown removes event listeners using the stored wrapper identities.
  \item Component teardown is two-phase: synchronous unmount + scheduled \texttt{onUnmounted} hook.
  \item Deleting \texttt{.el} and clearing instance fields is a core anti-leak strategy.
\end{itemize}

\section{Walking the tests: what they prove}

This section treats the tests as a teardown requirements document.

\subsection{\texttt{destroy a text element}}

From \texttt{src/\_\_tests\_\_/destroy-dom.test.ts}:

\begin{itemize}
  \item Mount a text VNode via \texttt{mountDOM} and assert \texttt{vdom.el} is a \texttt{Text} node.
  \item Call \texttt{destroyDOM(vdom)}.
  \item Assert the document is empty \emph{and} \texttt{allElementsHaveBeenDestroyed(vdom)} returns true.
\end{itemize}

The helper \texttt{allElementsHaveBeenDestroyed()} recursively checks for the absence of \texttt{.el}. This is the strongest teardown invariant in this test suite.

\subsection{\texttt{destroy an html element and its children}}

This test ensures an \texttt{ELEMENT} node:

\begin{itemize}
  \item removes the parent element from DOM
  \item still destroys (cleans up) its children VNodes
\end{itemize}

It's the ``DOM removed'' + ``runtime graph cleaned'' pairing again.

\subsection{\texttt{remove an html element event listeners}}

This is the leak-prevention test.

\begin{enumerate}
  \item Mount a \texttt{button} with \texttt{on: \{ click: handler \}}.
  \item Trigger a click and assert the handler is called.
  \item Destroy the VNode.
  \item Trigger another click on the previously captured \texttt{buttonEl} and assert the handler isn't called again.
\end{enumerate}

This works because during mount, Relax installs a wrapper listener and stores it in \texttt{vdom.listeners}. During destroy, \texttt{removeEventListeners(vdom.listeners, el)} uses those wrapper identities to detach correctly.

\subsection{Recursive destruction and fragments}

	texttt{destroy an html element and its children recursively}, \texttt{destroy a fragment}, and \texttt{destroy a fragment recursively} collectively prove that:

\begin{itemize}
  \item recursion always visits nested VNodes
  \item fragment nodes are teardown-neutral: they simply forward destruction to children
\end{itemize}

\subsection{\texttt{destroy a component with subcomponents}}

This test ensures that the recursion crosses component boundaries.
The starting VNode is an element \texttt{div} containing a component.
Destroying the root must unmount the component and destroy its rendered subtree, including nested child components.

\subsection{Lifecycle scheduling tests (\texttt{component-lifecycle.test.ts})}

The component lifecycle tests focus on \texttt{enqueueJob}/\texttt{nextTick()} behavior:

\begin{itemize}
  \item \texttt{onMounted} is not expected to run synchronously with \texttt{mountDOM}; the test awaits \texttt{nextTick()}.
  \item Symmetrically, \texttt{onUnmounted} is not expected to run synchronously with \texttt{destroyDOM}; the test awaits \texttt{nextTick()}.
\end{itemize}

This enforces a stable mental model: lifecycle hooks run in the scheduler, not ``inline'' within the DOM mutation operations.

\subsection{HRBR teardown behavior (\texttt{hrbr-integration.test.ts})}

The HRBR integration suite demonstrates that \texttt{destroyDOM()} participates in interop cleanup.

\begin{itemize}
  \item In the ``can mount an HRBR block'' test, the block's \texttt{destroy()} clears the host; afterwards, Relax removes the host itself.
  \item In ``patch preserves HRBR instance'', destroying triggers the instance's \texttt{destroy()} and the test asserts it was called.
\end{itemize}

\section{Online resources}

\begin{itemize}
  \item MDN: \href{https://developer.mozilla.org/en-US/docs/Web/API/Element/remove}{\texttt{Element.remove()}}
  \item MDN: \href{https://developer.mozilla.org/en-US/docs/Web/API/EventTarget/removeEventListener}{\texttt{removeEventListener()}} and listener identity
  \item MDN: \href{https://developer.mozilla.org/en-US/docs/Web/JavaScript/Memory_Management}{JavaScript memory management (GC and references)}
  \item React docs (conceptual parallel): \href{https://react.dev/reference/react/useEffect#effects-and-cleanup}{Effects and cleanup}
  \item Vue docs (conceptual parallel): \href{https://vuejs.org/guide/essentials/lifecycle.html}{Lifecycle hooks}
\end{itemize}

\section{Failure modes: the bugs you\'ll actually hit}

When teardown is wrong, the DOM often still ``looks fine'' because it\'s gone.
The failures show up later as leaks, stale references, and re-entrancy.

Use this mapping as a debugging guide.

\subsection{Zombie event handlers (detached element still fires)}

	extbf{Symptom:} you remove a view, but an old click handler still runs (or memory grows after repeated mount/unmount cycles).

	extbf{Why it happens:} a DOM node can be detached from the document yet still exist as a JavaScript object.
If some code keeps a reference to the node (like the test does with \texttt{buttonEl}), it can still dispatch events and hold listener closures.

	extbf{Root cause:} listener identity mismatch. The function you pass to \texttt{removeEventListener} must be the same function object you passed to \texttt{addEventListener}.
Relax solves this by storing listener wrappers on the VNode (\texttt{vdom.listeners}) during mount/patch, then reusing them during destroy.

Reference: \href{https://developer.mozilla.org/en-US/docs/Web/API/EventTarget/removeEventListener}{\texttt{removeEventListener()}}.

\subsection{Use-after-destroy (stale VNodes still referenced)}

	extbf{Symptom:} something tries to patch using an old tree, throws because \texttt{el} is missing, or silently corrupts the UI.

	extbf{Root cause:} application code treated VNodes as durable objects instead of ephemeral render outputs.
VNodes are only valid while they are mounted; after \texttt{destroyDOM}, that VNode graph should be considered dead.

	extbf{Relax\'s guard rail:} the shared post-step \texttt{delete vdom.el} is a fail-fast mechanism.

\subsection{Unmount hook re-entrancy}

	extbf{Symptom:} an \texttt{onUnmounted} hook triggers an update which tries to patch while destroy is mid-flight.

	extbf{Design fix in Relax:} hooks are scheduled via \texttt{enqueueJob}.
Teardown becomes ``do all structural cleanup first, then run user code''.

Conceptual parallel: React\'s effect cleanup happens after commit, not in the middle of DOM mutations:\
\href{https://react.dev/reference/react/useEffect#effects-and-cleanup}{Effects and cleanup}.

\subsection{HRBR interop leaks}

	extbf{Symptom:} embedded widgets keep timers/observers alive after unmount.

	extbf{Root causes:}

\begin{itemize}
  \item the mounted instance didn\'t implement \texttt{dispose}/\texttt{destroy}
  \item the factory identity changes on every render, forcing remount and skipping the ``reuse'' mental model
\end{itemize}

	extbf{Fix:} implement at least one cleanup hook and keep the mount factory identity stable when you intend reuse.

\section{Edge cases worth understanding}

\subsection{Fragments and multi-root components}

Fragments don\'t have a single \texttt{el}. So teardown means:

\begin{itemize}
  \item recursively destroy children
  \item ensure each child loses its \texttt{.el} pointer
\end{itemize}

This is also why \texttt{Component.elements} returns an array: a component might render a fragment, and its DOM footprint is multiple siblings.

\subsection{Double destroy}

Relax does not promise idempotent teardown.
In practice, calling destroy twice usually means ownership got confused (two subsystems believe they own the same subtree).

If your application needs idempotence, implement it at a higher level (router/view manager) by tracking whether a screen is currently mounted.

\subsection{Destroy order (parent removed before child teardown)}

Relax removes the parent element first, then recurses.
That feels reversed until you internalize the key idea:

\begin{itemize}
  \item removing the parent is the browser-side detach
  \item recursion is runtime cleanup (listeners, hooks, deleting references)
\end{itemize}

So this ordering reduces intermediate observable states in the live document, while still guaranteeing runtime cleanup.

\section{Practical patterns for leak-free components}

Relax provides hooks, but you still need discipline.
Most real apps leak not because the renderer forgot something, but because application code created resources without cleaning them up.

\subsection{Common resources you must clean up}

\begin{itemize}
  \item \textbf{Timers:} \texttt{setInterval}/\texttt{setTimeout} \textrightarrow clear them in \texttt{onUnmounted}
  \item \textbf{Global listeners:} \texttt{window.addEventListener} \textrightarrow remove in \texttt{onUnmounted}
  \item \textbf{Observers:} \texttt{ResizeObserver}/\texttt{IntersectionObserver} \textrightarrow \texttt{disconnect()} in \texttt{onUnmounted}
  \item \textbf{Fetches:} cancel via \texttt{AbortController} in \texttt{onUnmounted}
\end{itemize}

Reference: \href{https://developer.mozilla.org/en-US/docs/Web/API/AbortController}{\texttt{AbortController}}.

\subsection{A small \texttt{resourceBag} pattern (plain JavaScript)}

When a component grows, cleanup code tends to scatter.
A tiny pattern keeps it centralized: store cleanup callbacks and run them in the teardown hook.

\begin{lstlisting}[style=relax,language=JavaScript]
function createResourceBag() {
  const cleanups = []
  return {
    add(fn) {
      cleanups.push(fn)
      return fn
    },
    run() {
      for (const fn of cleanups.splice(0)) fn()
    },
  }
}

// Example usage:
const bag = createResourceBag()
const onResize = () => {}

window.addEventListener('resize', onResize)
bag.add(() => window.removeEventListener('resize', onResize))

// later, onUnmounted:
bag.run()
\end{lstlisting}

The important rule is ownership: only add resources that \emph{this component created}.

\section{Exercises}

\begin{enumerate}
  \item \textbf{Detached-element click:}
    recreate the listener test mentally. Why is clicking a detached element a meaningful leak test?

  \item \textbf{Idempotent destroy wrapper:}
    design \texttt{safeDestroy(vdom)}. Where should an \texttt{isDestroyed} flag live: VNode, component instance, or app router?

  \item \textbf{HRBR cleanup contract:}
    write a mount factory that starts an interval timer. Implement \texttt{dispose} or \texttt{destroy} so teardown stops it.

  \item \textbf{Scheduling thought experiment:}
    list two ways inline \texttt{onUnmounted} execution could corrupt teardown correctness.
\end{enumerate}

\section{Chapter summary}

\begin{itemize}
  \item Teardown is two jobs: destroying DOM nodes and destroying the runtime graph.
  \item \texttt{destroyDOM} is a type-dispatcher that recursively tears down VNode trees.
  \item Element teardown removes event listeners using stored wrapper identities.
  \item Component teardown is two-phase: synchronous \texttt{unmount()} + scheduled \texttt{onUnmounted}.
  \item HRBR teardown is a small resource protocol (optional \texttt{dispose}/\texttt{destroy} + remove host + delete refs).
\end{itemize}
