\chapter{Scheduling, nextTick, and Deterministic Work}
\label{ch:scheduling}

Scheduling is one of those topics that every UI developer feels they "kind of" understands
even while still getting surprised by it weekly.
So we\'ll treat the scheduler as a real subsystem with a contract, not a vague "async happens later" box.

Relax has \emph{two} schedulers:

\begin{itemize}
  \item A small microtask-based job queue used by the classic VDOM layer (\texttt{src/scheduler.ts}).
  \item A more capable lane-based scheduler used by the runtime/reactive layer (\texttt{runtime/scheduler.ts}).
\end{itemize}

They serve different purposes, but they share one theme: \textbf{make UI work predictable}.

Scheduling discussions tend to get vague fast ("it’s async, it’ll happen later").
In Relax, scheduling is treated as part of the renderer’s public contract: tests depend on it, and lifecycle semantics are defined by it.

This chapter is anchored in:

\begin{itemize}
  \item \texttt{src/scheduler.ts} and \texttt{src/\_\_tests\_\_/scheduler.test.ts}
  \item \texttt{runtime/scheduler.ts} and \texttt{runtime/\_\_tests\_\_/scheduler.test.ts}
\end{itemize}

\section{Diagram: two schedulers, two contracts}
Relax has two schedulers because it has two kinds of work with different needs:

\begin{center}
\begin{tikzpicture}[
  node distance=10mm,
  box/.style={draw,rounded corners,fill=RelaxLight,inner sep=5pt,align=left},
  arrow/.style={-Latex,thick}
]
  \node[box] (vdom) {\textbf{VDOM scheduler}\\\texttt{src/scheduler.ts}\\goal: lifecycle hooks + small deferred jobs};
  \node[box, right=18mm of vdom] (rt) {\textbf{Runtime scheduler}\\\texttt{runtime/scheduler.ts}\\goal: priority lanes + budgets + large reactive work};

  \node[box, below=of vdom] (vdomc) {\textbf{contract}\\- FIFO job queue\\- one microtask flush per tick\\- doesn't await async jobs};
  \node[box, below=of rt] (rtc) {\textbf{contract}\\- lane priority\\- deterministic ordering\\- time slicing via budget};

  \draw[arrow] (vdom) -- (vdomc);
  \draw[arrow] (rt) -- (rtc);
\end{tikzpicture}
\end{center}

\section{The VDOM microtask job queue (\texttt{src/scheduler.ts})}

Before we read Relax\'s scheduler, we need one shared mental model: \textbf{the JavaScript event loop}.

\subsection{Mental model: one thread, a call stack, and two queues}

In a browser (and in many JS environments), your code runs on one main thread.
Execution is a loop:

\begin{enumerate}
  \item run synchronous code on the call stack
  \item when the stack is empty, run microtasks (promises / \texttt{queueMicrotask}) until empty
  \item then allow the next macrotask (timers, IO, UI events) to run
\end{enumerate}

Two important consequences for UI frameworks:

\begin{itemize}
  \item \textbf{Microtasks run sooner than timers.} If you want "run after this render, but before the browser handles the next event",
    microtasks are the sharp tool.
  \item \textbf{Microtasks can starve the browser if abused.} If you keep queueing microtasks from microtasks,
    you can delay input/paint.
\end{itemize}

References: \href{https://developer.mozilla.org/en-US/docs/Web/JavaScript/Event_loop}{MDN: JavaScript event loop}
and \href{https://developer.mozilla.org/en-US/docs/Web/API/queueMicrotask}{MDN: \texttt{queueMicrotask()}}.

The VDOM layer uses a minimal scheduling primitive: \texttt{enqueueJob(job)}.

\subsection{Why it exists}

Several operations in Relax are intentionally deferred:

\begin{itemize}
  \item component \texttt{onMounted()} is scheduled after mount completes
  \item component \texttt{onUnmounted()} is scheduled after unmount begins
\end{itemize}

Deferring hook execution avoids re-entrancy surprises and gives the renderer a clean boundary: "render/mount now, run hooks shortly after".

\paragraph{The hidden constraint: DOM consistency and re-entrancy}

If \texttt{onMounted()} ran inside \texttt{mountDOM()}, a component could synchronously call \texttt{updateState()}, which triggers patching while the initial mount is still underway.
By scheduling hooks, Relax ensures the mount/patch/destroy primitives finish their structural work before user hook code runs.

\subsection{Data model}

\texttt{src/scheduler.ts} maintains:

\begin{itemize}
  \item \texttt{jobs}: an array of queued callbacks
  \item \texttt{isScheduled}: a flag to ensure we schedule processing only once per tick
\end{itemize}

\subsection{enqueueJob and queueMicrotask}

\texttt{enqueueJob(job)} pushes the job and calls \texttt{scheduleUpdate()}.

\texttt{scheduleUpdate()} is the key:

\begin{verbatim}
if (isScheduled) return
isScheduled = true
queueMicrotask(processJobs)
\end{verbatim}

So no matter how many jobs you enqueue, you’ll process them in a single microtask flush.

\subsection*{Diagram: microtask batching timeline}
This is the behavior the tests rely on:

\begin{center}
\begin{tikzpicture}[
  node distance=7mm,
  box/.style={draw,rounded corners,fill=RelaxLight,inner sep=5pt,align=left},
  arrow/.style={-Latex,thick}
]
  \node[box] (sync) {\textbf{sync JS turn}\\enqueueJob(A)\\enqueueJob(B)\\enqueueJob(C)};
  \node[box, right=14mm of sync] (micro) {\textbf{microtask}\\processJobs():\\A then B then C};
  \node[box, right=14mm of micro] (macro) {\textbf{macrotask}\\\texttt{flushPromises()} / timers};

  \draw[arrow] (sync) -- (micro);
  \draw[arrow] (micro) -- (macro);
\end{tikzpicture}
\end{center}

This is also why hook scheduling avoids re-entrancy: hooks run after the renderer finishes its synchronous DOM work.

\paragraph{Batching rule}

The \texttt{isScheduled} latch is the batching mechanism:

\begin{itemize}
  \item enqueue one job \textrightarrow schedule one microtask
  \item enqueue many more jobs before the microtask runs \textrightarrow still one microtask
\end{itemize}

This makes lifecycle hooks cheap even when many components mount in the same synchronous turn.

\subsection{processJobs: ordering and async behavior}

\texttt{processJobs()} drains \texttt{jobs} in FIFO order. For each job:

\begin{itemize}
  \item it calls \texttt{job()}
  \item it wraps the result with \texttt{Promise.resolve(result).then(...)} to log async errors
\end{itemize}

Important nuance (documented in the code): \textbf{async jobs do not block the queue}.
The queue keeps running other jobs; the promise continuation runs later.

\subsubsection{A code walk (as JavaScript)}

Here is the scheduler rewritten into minimal JavaScript so the control flow is obvious.
It mirrors \texttt{src/scheduler.ts} exactly, just without TypeScript types:

\begin{lstlisting}[style=relax,language=JavaScript]
let isScheduled = false
const jobs = []

export function enqueueJob(job) {
  jobs.push(job)
  scheduleUpdate()
}

function scheduleUpdate() {
  if (isScheduled) return
  isScheduled = true
  queueMicrotask(processJobs)
}

function processJobs() {
  while (jobs.length > 0) {
    const job = jobs.shift()
    const result = job()

    // Important: we don't await async jobs.
    // We only attach a rejection handler so async errors are visible.
    Promise.resolve(result).then(
      () => {},
      (error) => console.error(`[scheduler]: ${error}`)
    )
  }
  isScheduled = false
}
\end{lstlisting}

From this we can state the contract precisely:

\begin{itemize}
  \item Jobs run \textbf{after} the current synchronous call stack.
  \item Jobs run in \textbf{FIFO order} within a single flush.
  \item The queue does \textbf{not await} async jobs.
  \item Async errors are surfaced (logged) instead of turning into silent unhandled rejections.
\end{itemize}

\paragraph{Why the queue doesn’t await}

The VDOM scheduler is designed for hook-style work ("start it soon, don’t stall unrelated hooks, but surface errors").
If it awaited each job, one slow async hook could hold up every other component’s hook.

This is tested in \texttt{src/\_\_tests\_\_/scheduler.test.ts}:

\begin{itemize}
  \item \texttt{Asynchronous jobs enqueue more jobs} expects order \texttt{[1, 3, 2]}.
\end{itemize}

That test shows:

\begin{enumerate}
  \item async job pushes \texttt{1}
  \item it awaits a microtask, so it pauses
  \item the next (sync) job runs and pushes \texttt{3}
  \item then the async job resumes and pushes \texttt{2}
\end{enumerate}

\subsection{nextTick(): what it guarantees (and what it doesn’t)}

\texttt{nextTick()} is the testing and integration helper:

\begin{verbatim}
scheduleUpdate()
return flushPromises()
\end{verbatim}

\texttt{flushPromises()} is implemented as \texttt{setTimeout(resolve)}.
That means \textbf{nextTick resolves after the microtask flush and one macrotask turn}.

The comment in \texttt{src/scheduler.ts} is explicit:

\begin{quote}
If the jobs are asynchronous, the promise will resolve before all the jobs have completed.
\end{quote}

So the contract is:

\begin{itemize}
  \item It ensures the queued jobs have been \emph{started} and synchronous effects have run.
  \item It does not await completion of async work inside a job.
\end{itemize}

\paragraph{Practical testing guidance}


    \texttt{nextTick()} is a boundary for "scheduler-driven work started", not "all async completed".
If you write a hook that does work after an \texttt{await}, you may need an additional wait (or design your test so the awaited promise resolves immediately).

\subsubsection{Why does \texttt{flushPromises()} use \texttt{setTimeout}?}

This is subtle but important.
Relax\'s \texttt{nextTick()} is:

\begin{verbatim}
scheduleUpdate()
return flushPromises() // setTimeout(resolve)
\end{verbatim}

If \texttt{nextTick()} only awaited a microtask, you could end up in situations where:

\begin{itemize}
  \item the scheduler flush runs in a microtask
  \item a job itself queues more microtasks
  \item and your test resumes before the environment has advanced a real "turn" (macrotask)
\end{itemize}

Using \texttt{setTimeout(resolve)} forces the promise to settle on the next macrotask,
which is a stronger "breathing point" for tests.

Reference: \href{https://developer.mozilla.org/en-US/docs/Web/API/setTimeout}{MDN: \texttt{setTimeout()}}.

\subsubsection{Batching correctness: a tiny proof sketch}

The batching mechanism is the \texttt{isScheduled} latch.
Here\'s the key claim:

\begin{quote}
Within a single synchronous JS turn, regardless of how many times \texttt{enqueueJob()} is called,
      \texttt{processJobs()} will be scheduled at most once.
\end{quote}

Reasoning:

\begin{enumerate}
  \item The first \texttt{enqueueJob()} sees \texttt{isScheduled === false}, flips it to true, and schedules one microtask.
  \item Every later \texttt{enqueueJob()} in the same turn hits \texttt{if (isScheduled) return} and schedules nothing.
  \item Only after the microtask flush completes does \texttt{processJobs()} set \texttt{isScheduled = false} again.
\end{enumerate}

So: one turn \textrightarrow one microtask flush \textrightarrow all jobs drained.

\subsubsection{Timeline: why the tests produce those arrays}

The scheduler tests are easiest to understand as timelines.

\paragraph{Jobs run after synchronous code (expects \texttt{[4, 1, 2, 3]})}

\begin{enumerate}
  \item Sync: enqueue jobs \texttt{1,2,3}. Scheduler schedules one microtask flush.
  \item Sync: push \texttt{4} directly into the array.
  \item Microtask: scheduler flush runs and executes \texttt{1,2,3}.
\end{enumerate}

So the sync side-effect (\texttt{4}) happens before the microtask side-effects.

\paragraph{Asynchronous jobs enqueue more jobs (expects \texttt{[1, 3, 2]})}

This is the "async jobs don\'t block" rule.

\begin{enumerate}
  \item Microtask flush starts.
  \item Job A runs: pushes \texttt{1}, then hits \texttt{await Promise.resolve()} and yields.
  \item Scheduler continues the flush loop and runs Job B (sync): pushes \texttt{3}.
  \item Later microtask: Job A continuation runs and pushes \texttt{2}.
\end{enumerate}

So \texttt{3} appears between \texttt{1} and \texttt{2}.

Reference: \href{https://developer.mozilla.org/en-US/docs/Web/JavaScript/Reference/Global_Objects/Promise}{MDN: \texttt{Promise}} (microtask semantics).

Tests in \texttt{src/\_\_tests\_\_/scheduler.test.ts} cover:

\begin{itemize}
  \item jobs run after \texttt{nextTick}
  \item jobs run in order
  \item jobs run after synchronous code
\end{itemize}

\section{The runtime lane scheduler (\texttt{runtime/scheduler.ts})}

The runtime scheduler is built for reactive workloads and larger apps.
Instead of a single FIFO queue, it manages multiple priority lanes:

\begin{itemize}
  \item \texttt{sync}
  \item \texttt{input}
  \item \texttt{default}
  \item \texttt{transition}
  \item \texttt{idle}
\end{itemize}

This scheduler is deterministic by design.
It uses stable tie-breakers, explicit budgets, and a promotion system (aging + deadlines) so low-priority work eventually runs.

This section is grounded directly in \texttt{runtime/scheduler.ts} and \texttt{runtime/\_\_tests\_\_/scheduler.test.ts}.

\subsection{Creating a scheduler}

The entry points:

\begin{itemize}
  \item \texttt{createScheduler(options)}
  \item \texttt{createRequestFlush(strategy)} for how flush is triggered
  \item \texttt{createBrowserScheduler(...)} convenience wrapper
\end{itemize}

\subsection{Budgets and time slicing}

A scheduler flush is run with a budget (in milliseconds). The flush loop:

\begin{itemize}
  \item picks the next task
  \item runs it
  \item checks elapsed time
  \item if budget is exceeded and work remains, schedules another flush
\end{itemize}

This behavior is tested in \texttt{runtime/\_\_tests\_\_/scheduler.test.ts}:

\begin{itemize}
  \item \texttt{budget enforcement: flush splits work across multiple flush calls when budget is tight}
  \item \texttt{setFrameBudget caps the budget used by scheduled flushes}
\end{itemize}

\subsection{Lane priority and deterministic ordering}

Lane priority is declared in \texttt{LANE\_PRIORITY} and selection order is:

\begin{verbatim}
sync > input > default > transition > idle
\end{verbatim}

\subsection*{Diagram: lane selection and budgeted flushing}

\begin{center}
\begin{tikzpicture}[
  node distance=8mm,
  box/.style={draw,rounded corners,fill=RelaxLight,inner sep=5pt,align=left},
  arrow/.style={-Latex,thick}
]
  \node[box] (pick) {Pick next lane\\(sync \textrightarrow input \textrightarrow default \textrightarrow transition \textrightarrow idle)};
  \node[box, below=of pick] (run) {Run tasks in lane\\(deterministic order)};
  \node[box, below=of run] (budget) {Check elapsed time\\budget exceeded?};
  \node[box, below left=of budget, xshift=14mm] (yield) {yes \textrightarrow yield\\schedule another flush};
  \node[box, below right=of budget, xshift=-14mm] (done) {no \textrightarrow continue\\or finish if empty};

  \draw[arrow] (pick) -- (run);
  \draw[arrow] (run) -- (budget);
  \draw[arrow] (budget) -- (yield);
  \draw[arrow] (budget) -- (done);
\end{tikzpicture}
\end{center}

Within a lane, tasks are sorted with deterministic tie-breakers:

\begin{enumerate}
  \item earlier deadline
  \item earlier timestamp
  \item lower task id
\end{enumerate}

Test: \texttt{orders tasks within a lane by explicit deadline (then timestamp/id)}.

In code, the deterministic ordering is implemented by sorting each lane queue using:

\begin{verbatim}
(a.deadline - b.deadline) || (a.timestamp - b.timestamp) || (a.id - b.id)
\end{verbatim}

The final tie-breaker (\texttt{id}) is the determinism anchor: it makes equal-time scheduling stable.

\subsection{Aging and promotion (starvation prevention)}

To prevent starvation, the scheduler promotes work that is overdue or old.

The implementation:

\begin{itemize}
  \item defines \texttt{AGING\_MS = 250}
  \item in \texttt{pickNextTask()}, checks queue heads in weaker lanes
  \item if a head task is overdue (deadline reached) or aged, it is promoted one lane upward
\end{itemize}

Tests:

\begin{itemize}
  \item \texttt{ages/pushes starving tasks upwards (idle -> transition)}
  \item \texttt{promotes a task when its deadline is reached (even if it is not old enough)}
  \item \texttt{starvation prevention: repeated promotions eventually let idle work run even under continuous input load}
\end{itemize}

This is the scheduler equivalent of "don’t let background work get stuck forever".

\subsection{withBudget(): integrating user work}

The scheduler also offers \texttt{withBudget(budgetMs, fn)}.
It executes user code and, if that work consumes the budget while tasks remain, schedules continuation.

Test: \texttt{withBudget schedules continuation when user code exhausts budget and tasks remain}.

\section{How these schedulers relate to the book}

In Part II and Part IV we saw that the VDOM layer uses \texttt{enqueueJob} to schedule lifecycle hooks.

In later parts of the book (signals, blocks, hydration), we’ll lean more on the runtime scheduler.
The runtime lane model is a foundation for:

\begin{itemize}
  \item fair scheduling of effects
  \item batching updates
  \item making performance tunable (frame budgets, flush strategy)
\end{itemize}

\section{Takeaways}

\begin{itemize}
  \item \texttt{src/scheduler.ts} is tiny but intentional: microtask batching + safe error logging.
  \item \texttt{nextTick()} is a boundary for tests and integration, but does not await async job completion.
  \item \texttt{runtime/scheduler.ts} adds lanes, deadlines, aging, and budgets for scalable reactive workloads.
\end{itemize}

\section{Walking the tests: what they prove}

You can treat these as the scheduler’s executable specification.

\subsection{VDOM scheduler tests (\texttt{src/\_\_tests\_\_/scheduler.test.ts})}

\begin{description}
  \item[Enqueued jobs run after nextTick] Proves \texttt{enqueueJob()} is deferred and does not run synchronously.
  \item[Enqueued jobs run in order] Proves FIFO execution order in \texttt{processJobs()}.
  \item[Jobs run after synchronous code] Proves microtasks run after the current synchronous call stack.
  \item[Asynchronous jobs enqueue more jobs] Proves the "async doesn’t block" rule: a job that yields allows later jobs to run, producing \texttt{[1, 3, 2]}.
\end{description}

\subsection{Runtime scheduler tests (\texttt{runtime/\_\_tests\_\_/scheduler.test.ts})}

The runtime scheduler suite uses a manual clock (\texttt{now()}) and a manual flush trigger (\texttt{requestFlush(cb)} pushes \texttt{cb} into an array).
That technique eliminates timing flakiness and makes time-based scheduling deterministic.

Key tests and the guarantees they enforce:

\begin{description}
  \item[runs tasks by lane priority] Proves lane selection order and that \texttt{sync} runs immediately.
  \item[budget enforcement / setFrameBudget] Proves time slicing across multiple flush calls when the budget is tight.
  \item[ages/pushes starving tasks upwards] Proves aging-based promotion (\texttt{AGING\_MS = 250}).
  \item[promotes a task when its deadline is reached] Proves deadline-based promotion even when a task isn’t old.
  \item[withBudget schedules continuation] Proves user code that consumes budget can trigger a continuation flush if tasks remain.
  \item[createRequestFlush(messageChannel/raf)] Proves flush strategy selection when platform APIs exist.
\end{description}

\section{Online resources}

\begin{itemize}
  \item MDN: \href{https://developer.mozilla.org/en-US/docs/Web/API/queueMicrotask}{\texttt{queueMicrotask()}}
  \item MDN: \href{https://developer.mozilla.org/en-US/docs/Web/JavaScript/Event_loop}{JavaScript event loop (microtasks vs. macrotasks)}
  \item WHATWG HTML: \href{https://html.spec.whatwg.org/multipage/webappapis.html#microtask-queue}{Microtask queue (spec)}
  \item React (conceptual parallel): \href{https://react.dev/learn/queueing-a-series-of-state-updates}{Queued updates and batching}
  \item MDN (conceptual parallel): \href{https://developer.mozilla.org/en-US/docs/Web/API/Background_Tasks_API}{Cooperative background work}
\end{itemize}

\section{Pitfalls and debugging (senior notes)}

This section is where real-world scheduler bugs live.

\subsection{Mistaking \texttt{nextTick()} for "wait for all async"}

      \texttt{nextTick()} does not promise to await async jobs.
It promises the scheduler had a chance to flush and start them.

If a lifecycle hook does:

\begin{verbatim}
async onMounted() {
  await fetch(...)
  this.updateState(...)
}
\end{verbatim}

then \texttt{await nextTick()} only guarantees the hook started.
It does \emph{not} guarantee the \texttt{fetch} completed.

\subsection{Microtask storms (starving input/paint)}

If jobs keep enqueueing more jobs in a tight cycle, you can monopolize the microtask queue.
In production this looks like:

\begin{itemize}
  \item input lag
  \item delayed paints
  \item "the browser froze" reports
\end{itemize}

This is one reason the runtime scheduler exists: it introduces budgets and yielding.

\subsection{Re-entrancy and why hooks are scheduled}

The component lifecycle tests (\texttt{src/\_\_tests\_\_/component-lifecycle.test.ts}) enforce the contract:

\begin{itemize}
  \item \texttt{mountDOM}/\texttt{destroyDOM} do structural work synchronously
  \item \texttt{onMounted}/\texttt{onUnmounted} run later via the scheduler
\end{itemize}

This keeps the renderer from entering "mount triggers patch triggers mount" spirals.

\section{Exercises}

\begin{enumerate}
  \item \textbf{Order reasoning:} explain why the scheduler test expects \texttt{[4, 1, 2, 3]} in
      \texttt{Jobs run after synchronous code}. Write the timeline in terms of call stack vs microtasks.

  \item \textbf{Async ordering:} explain why \texttt{Asynchronous jobs enqueue more jobs} yields \texttt{[1, 3, 2]}.
    What would have to change in the scheduler for it to yield \texttt{[1, 2, 3]} instead?

  \item \textbf{Design exercise:} if you wanted a variant \texttt{nextTickAll()} that awaited async jobs too,
    what bookkeeping would you need? What new failure mode could that introduce (hint: one slow job blocks all hooks)?

  \item \textbf{Lane intuition:} when would you choose \texttt{transition} lane over \texttt{input} lane?
    Give a UI example (typing vs expensive filtering).
\end{enumerate}

\section{Chapter summary}

\begin{itemize}
  \item The VDOM scheduler is intentionally tiny: FIFO microtask batching for lifecycle hooks.
  \item The key mechanism is the \texttt{isScheduled} latch, collapsing many enqueues into one flush.
  \item \texttt{nextTick()} is a useful flush boundary for tests, but it is not an "await all async" primitive.
  \item The runtime scheduler adds lanes, deadlines, budgets, and starvation prevention for large reactive workloads.
\end{itemize}
