\chapter{SSR Authoring (v1): Rendering Blocks to HTML Strings}

Server-side rendering (SSR) is the process of producing HTML \emph{before} your client JavaScript boots.
In practical terms, SSR answers one very specific question:

\begin{quote}
What HTML should the server send so the client can \emph{attach} to it without rewriting?
\end{quote}

In HRBR v1, SSR is intentionally minimal.
It primarily exists to make hydration testable and to provide a predictable authoring workflow.
The key property is not raw server performance; it’s \textbf{compatibility} with hydration.

\section{What you’ll build}

By the end of this chapter you’ll be able to:

\begin{itemize}
  \item render a \texttt{BlockDef} to a string via \texttt{renderBlockToString}
  \item understand why SSR v1 is implemented as \emph{mount into a detached host $\rightarrow$ serialize}
  \item reason about what breaks hydration, and how the SSR/hydration tests detect those breaks
\end{itemize}

\section{What you’ll learn}

\begin{itemize}
  \item the “ABI” idea: \texttt{templateHTML} is a stable structure shared by server render and client hydrate
  \item why HRBR treats SSR as a special case of the same slot-write machinery used on the client
  \item what SSR v1 explicitly does \emph{not} attempt (DOM-free HTML emission, streaming, partial hydration)
\end{itemize}

\section{Prerequisites}

This chapter builds on:

\begin{itemize}
  \item Chapter 14 (blocks: \texttt{templateHTML}, slots, and stable paths)
  \item Chapter 15 (hydration: validate + attach, else bailout/remount)
\end{itemize}

\begin{center}
\begin{tikzpicture}[
  node distance=9mm,
  box/.style={draw, rounded corners, align=left, inner sep=5pt},
  arrow/.style={-Latex, thick}
]
\node[box] (def) {\textbf{BlockDef}\\templateHTML + slots};
\node[box, right=18mm of def] (ssr) {\textbf{SSR v1}\\mountBlock in detached host\\serialize \texttt{innerHTML}};
\node[box, right=18mm of ssr] (html) {\textbf{HTML string}\\server output};
\node[box, below=of html] (client) {\textbf{Client}\\insert HTML into host\\hydrateBlock attach};

\draw[arrow] (def) -- (ssr);
\draw[arrow] (ssr) -- (html);
\draw[arrow] (html) -- (client);
\end{tikzpicture}
\end{center}

This chapter is grounded in:

\begin{itemize}
  \item \texttt{runtime/ssr.ts}
  \item \texttt{runtime/\_\_tests\_\_/ssr.test.ts}
  \item and the follow-on behavior in \texttt{runtime/hydration.ts} (Chapter 15).
\end{itemize}

\section{Why SSR exists in HRBR}

If hydration is “attach to existing DOM”, SSR is “produce DOM that is safe to attach to”.
So SSR in this runtime is less about doing rendering on the server in the abstract, and more about enforcing a strong invariant:

\begin{quote}
The server must output markup whose structure matches what \texttt{mountBlock()} would create.
\end{quote}

This is why SSR v1 reuses the block runtime: it reduces the number of places where HTML/slot semantics can diverge.

\section{A tiny contract (what SSR v1 guarantees)}

SSR is not “render an app” in the abstract. In HRBR v1 it is narrowly scoped:

\begin{enumerate}
  \item You provide a \texttt{BlockDef} (template + slots).
  \item You optionally provide \texttt{initialValues} (a slot-key \(\rightarrow\) value map).
  \item The runtime returns an HTML string that matches the same template structure the client will mount/hydrate.
\end{enumerate}

Put differently:

\begin{quote}
SSR v1 is \emph{mountBlock + serialize}. Hydration v1 is \emph{validate + attach, else remount}.
\end{quote}

That quote is a useful mental checksum: if you ever change block slot semantics (Chapter 14), you should expect SSR and hydration tests to fail until all three layers agree again.

This tight coupling is deliberate. It makes SSR and hydration “round-trip compatible” by construction.

\section{The contract: \texttt{renderBlockToString(def, options?) -> string}}

The only exported function is:

\begin{verbatim}
renderBlockToString(def: BlockDef, options?: { initialValues?: Record<string, unknown> }): string
\end{verbatim}

The options only support one thing today: \texttt{initialValues}.

This mirrors the core block update contract:

\begin{quote}
A block update is just a map \texttt{\{ [slotKey]: value \}}.
\end{quote}

SSR is simply the "apply values, then serialize" version of that idea.

\subsection{A minimal end-to-end usage (SSR $\rightarrow$ hydrate $\rightarrow$ update)}

The fastest way to internalize SSR v1 is to look at how it composes with hydration.
The test suite models this exact authoring loop:

\begin{lstlisting}[language=JavaScript]
import { defineBlock, mountCompiledBlock } from 'relaxjs/runtime/block'
import { renderBlockToString } from 'relaxjs/runtime/ssr'
import { hydrateBlock } from 'relaxjs/runtime/hydration'

const Counter = defineBlock({
  templateHTML: `<div class="c"><button id="inc">+</button><span id="t"> </span></div>`,
  slots: {
    value: { kind: 'text', path: [1, 0] },
  },
})

// "Server" side: first paint
const html = renderBlockToString(Counter, { initialValues: { value: 'A' } })

// "Client" side: insert + attach
const host = document.getElementById('host')
host.innerHTML = html
const hydrated = hydrateBlock(Counter, host)

// Later: incremental updates can happen via hydration...
hydrated.update({ value: 'B' })

// ...or via compiled slots (Chapter 14/15 composition story)
const mounted = mountCompiledBlock(Counter, host, [{ key: 'value', read: () => 'C' }])
mounted.update({ value: 'C' })

mounted.dispose()
hydrated.destroy()
\end{lstlisting}

Two important observations:

\begin{itemize}
  \item SSR returns \emph{just a string}. The correctness part is whether the client can attach without rewriting.
  \item Both hydration and compiled mounts boil down to the same contract: \texttt{update(\{ [slotKey]: value \})}.
\end{itemize}

One way to think about it:

\begin{itemize}
  \item \textbf{Client mount:} template \(\rightarrow\) DOM instance \(+\) slot node references.
  \item \textbf{Client update:} slot values \(\rightarrow\) DOM writes.
  \item \textbf{Server render:} slot values \(\rightarrow\) DOM writes \(\rightarrow\) serialize.
\end{itemize}

\section{Implementation: DOM-based SSR for correctness}

\texttt{runtime/ssr.ts} implements SSR using the platform DOM:

\begin{enumerate}
  \item create a detached host element: \texttt{document.createElement('div')}
  \item mount the block into that host via \texttt{mountBlock(def, host, initialValues)}
  \item read \texttt{host.innerHTML}
  \item destroy the mounted block to avoid leaks
  \item return the HTML string
\end{enumerate}

This is documented directly in the function comment:

\begin{itemize}
  \item it uses a detached DOM
  \item it serializes via \texttt{innerHTML}
  \item it’s primarily intended for tests/examples
  \item a future version could emit HTML without a DOM
\end{itemize}

The crucial correctness choice is that SSR applies \texttt{initialValues} by calling \texttt{mountBlock(def, host, initialValues)}.
That means:

\begin{itemize}
  \item all the slot semantics from Chapter 14 apply (boolean attributes, SVG namespaces, input properties, etc.)
  \item the generated markup shape matches what the client expects to attach to
\end{itemize}

\subsection{Why DOM-based SSR is acceptable in v1}

For a production SSR renderer, involving a full DOM implementation server-side can be expensive.

But for v1 HRBR, this has two big advantages:

\begin{itemize}
  \item it is hard to get HTML escaping and attribute semantics wrong
  \item it produces markup exactly consistent with the same block runtime used in the client
\end{itemize}

So correctness comes for free.

\begin{center}
\begin{tikzpicture}[
  node distance=8mm,
  box/.style={draw, rounded corners, align=left, inner sep=5pt},
  arrow/.style={-Latex, thick}
]
\node[box] (host) {detached \texttt{div} host};
\node[box, below=of host] (mount) {\texttt{mountBlock(def, host, initialValues)}};
\node[box, below=of mount] (serialize) {read \texttt{host.innerHTML}};
\node[box, below=of serialize] (destroy) {\texttt{mounted.destroy()}};
\node[box, below=of destroy] (ret) {return string};

\draw[arrow] (host) -- (mount);
\draw[arrow] (mount) -- (serialize);
\draw[arrow] (serialize) -- (destroy);
\draw[arrow] (destroy) -- (ret);
\end{tikzpicture}
\end{center}

There’s also a subtle lifecycle advantage: because SSR uses \texttt{mountBlock}, tearing down is just \texttt{mounted.destroy()}.
That ensures listeners are removed and the temporary root is cleaned up, even in a test environment.

\subsection{The implementation is short on purpose}

SSR v1 is deliberately a thin adapter.
The actual code in \texttt{runtime/ssr.ts} is essentially:

\begin{lstlisting}[language=JavaScript]
export function renderBlockToString(def, options = {}) {
  const host = document.createElement('div')
  const mounted = mountBlock(def, host, options.initialValues ?? {})
  const html = host.innerHTML
  mounted.destroy()
  return html
}
\end{lstlisting}

This design matters because it centralizes correctness.
Instead of re-implementing HTML escaping, boolean attributes, SVG namespaces, and slot kinds in a second renderer,
SSR simply uses the same mount/update semantics as the client.

\section{SSR must round-trip with hydration}

SSR isn't useful unless the client can hydrate it.

Hydration v1 makes strong assumptions (Chapter 15):

\begin{itemize}
  \item the root element tagName matches \texttt{templateHTML}
  \item slot paths resolve to the expected node types
  \item element slots match expected tagNames along the path
\end{itemize}

Therefore, SSR output must preserve the template’s structure.
Because SSR uses \texttt{mountBlock()}, it naturally meets the same invariant:

\begin{quote}
SSR uses the same template parser and slot patching semantics as client mount.
\end{quote}

This is the important “authoring rule” for SSR v1:

\begin{quote}
If you change \texttt{templateHTML} shape, you must change both server render and client hydrate together.
\end{quote}

You should treat \texttt{templateHTML} as a stable ABI between server and client.

\subsection{Authoring rule: structure changes are breaking changes}

Hydration v1 validates aggressively and bails out if anything looks unfamiliar.
That implies a simple rule of thumb:

\begin{itemize}
  \item Changing slot \emph{values} is safe.
  \item Changing \texttt{templateHTML} \emph{structure} is a breaking change for hydration.
\end{itemize}

One subtle (and common) structural foot-gun: paths are \texttt{childNodes}-based.
Whitespace can introduce text nodes and shift indices.
So “just add formatting” in the HTML string can become a breaking change if it changes the node tree.

\section{The end-to-end spec: render \(\rightarrow\) hydrate \(\rightarrow\) update}

The key test is \texttt{runtime/\_\_tests\_\_/ssr.test.ts}:

\begin{enumerate}
  \item Define a block with \texttt{templateHTML} and a text slot.
  \item Render HTML with \texttt{renderBlockToString(block, \{ initialValues \})}.
  \item Insert the returned string into the DOM.
  \item Call \texttt{hydrateBlock(block, host)}.
  \item Verify DOM identity is preserved (same root, same text node).
  \item Mount a compiled block via \texttt{mountCompiledBlock(...)} and update.
\end{enumerate}

This accomplishes two things:

\begin{itemize}
  \item it proves that SSR output is "hydrate compatible"
  \item it shows how hydration and compiled blocks can be composed (a realistic client story)
\end{itemize}

One implementation-level nuance in the test is worth calling out.
After updating the signal with \texttt{setCount('B')}, it also calls:

\begin{quote}
	\texttt{mounted.update(\{ value: 'B' \})}
\end{quote}

That makes the assertion deterministic and keeps the test independent of scheduler timing.
It’s modeling the fact that \texttt{mountCompiledBlock} can schedule work, and test environments often differ in how timers/microtasks flush.

\subsection{Why the test checks identity, not just text}

It’s easy to write a system that “hydrates” by discarding server HTML and remounting.
But HRBR’s hydration story is explicitly about preserving DOM identity when markup matches.
So the test stores references like:

\begin{itemize}
  \item \texttt{rootBefore = host.firstElementChild}
  \item \texttt{textBefore = host.querySelector('\#t')?.firstChild}
\end{itemize}

and asserts they’re the same after \texttt{hydrateBlock()}.
That’s the heart of the v1 contract: \emph{attach without rewriting, or remount fully}.

\section{Limitations and the intended direction}

The HRBR whitepaper explicitly lists a non-goal for v1:

\begin{quote}
A DOM-free string renderer (v1 SSR uses a DOM-based approach for tests/examples)
\end{quote}

A future SSR implementation could:

\begin{itemize}
  \item parse \texttt{templateHTML} once into a token stream
  \item apply slot writes into a string builder
  \item avoid constructing DOM nodes entirely
\end{itemize}

Even if SSR becomes DOM-free later, the invariants from hydration still apply:

\begin{itemize}
  \item the root tag must match
  \item the structure along slot paths must match
  \item text nodes must exist where text slots expect them
\end{itemize}

But even then, the contract remains:

\begin{quote}
SSR output must match the template structure so hydration can do a no-rewrite attach.
\end{quote}

\section{Takeaways}

\begin{itemize}
  \item SSR v1 is intentionally minimal: mount a block in a detached DOM and serialize \texttt{innerHTML}.
  \item \texttt{initialValues} uses the same slot update contract as client updates.
  \item The primary goal is hydration compatibility, not server performance.
  \item The end-to-end test shows SSR + hydration + compiled updates working together.
\end{itemize}

\section{Walking the tests (the executable spec)}

SSR v1 has one test file, but it covers the entire story.

\subsection{\texttt{runtime/\_\_tests\_\_/ssr.test.ts}}

The test \texttt{renders a block to HTML string, hydrates it, and updates via mountCompiledBlock} enforces these guarantees:

\begin{itemize}
  \item \textbf{SSR applies initial values:} \texttt{renderBlockToString(..., \{ initialValues \})} must produce HTML containing \texttt{'A'}.
  \item \textbf{Hydration preserves identity:} after \texttt{hydrateBlock}, the root element and the text node under \texttt{\#t} are the same objects as before hydration.
  \item \textbf{Compiled updates work on hydrated DOM:} wiring a signal through \texttt{mountCompiledBlock} and then updating should change text to \texttt{'B'}.
  \item \textbf{Cleanup is explicit:} the test disposes the compiled mount and destroys the hydrated block.
\end{itemize}

Because this file touches SSR, hydration, signals, and compiled blocks, it acts as a integration-level "safety rail" for the whole pipeline.

\section{Common pitfalls}

\begin{itemize}
  \item \textbf{Thinking SSR is string concatenation:} in HRBR, correctness is driven by slot semantics (attr vs prop vs SVG namespaces).
  \item \textbf{Forgetting paths are \texttt{childNodes}-based:} whitespace and formatting can change indices.
  \item \textbf{Assuming partial hydration:} v1 is conservative; mismatch triggers remount.
  \item \textbf{Forgetting cleanup in examples/tests:} SSR cleans up its own temporary mount; hydration/compiled mounts still require explicit \texttt{destroy()/dispose()}.
\end{itemize}

\section{Mini-exercise}

Extend the SSR test scenario conceptually by adding an attribute slot and updating it after hydration.

\begin{enumerate}
  \item Add a \texttt{hidden} \texttt{attr} slot on the \texttt{<button id="inc">}.
  \item SSR render with \texttt{hidden: false} (should not emit the attribute).
  \item Hydrate without rewriting.
  \item Update with \texttt{hidden: true} and verify the attribute exists.
\end{enumerate}

This exercise forces you to connect all three layers:
slot semantics (Chapter 14), hydration attach rules (Chapter 15), and SSR reuse of mount semantics (this chapter).

\section{Online resources}

\begin{itemize}
  \item HTML serialization via DOM:
  \href{https://developer.mozilla.org/en-US/docs/Web/API/Element/innerHTML}{MDN: \texttt{Element.innerHTML}}.
  \item Server rendering and hydration overview:
  \href{https://web.dev/rendering-on-the-web/}{web.dev: Rendering on the Web}.
  \item Template element (used by mounting/hydration):
  \href{https://developer.mozilla.org/en-US/docs/Web/HTML/Element/template}{MDN: \texttt{<template>}}.
  \item A compiler-first argument for skipping VDOM in the steady state:
  \href{https://svelte.dev/blog/virtual-dom-is-pure-overhead}{Svelte: “Virtual DOM is pure overhead”}.
\end{itemize}
