\chapter{VDOM \texorpdfstring{$\leftrightarrow$}{<->} HRBR interop (mixed trees)}
\label{ch:vdom-hrbr-interop}

This chapter covers the experimental interop layer that allows compiled HRBR output to live \emph{inside} a Relax VDOM tree.
The goal is practical: you can adopt HRBR incrementally (or only for hot paths) without rewriting your whole app.

\subsection*{What you\'ll build}

By the end of the chapter you\'ll be able to:

\begin{itemize}
\item return an HRBR block from a normal Relax VDOM component (mixed trees)
\item understand why HRBR needs a wrapper host element and why that host is the thing the VDOM ``owns''
\item predict when patching will reuse an existing HRBR instance vs dispose/destroy/remount it
\item reason about event handler binding semantics when HRBR uses VDOM-style component instances
\end{itemize}

\subsection*{Prerequisites}

You should already be comfortable with:

\begin{itemize}
\item the Relax vnode model (Chapter~\ref{ch:vnode-model})
\item the mount/patch/destroy split in the DOM runtime (Chapters on \texttt{mountDOM} and \texttt{patchDOM})
\item the HRBR mental model: ``mount into a host, then the runtime owns what\'s inside''
\end{itemize}

\subsection*{Why this matters (the real-world motivation)}

In production UIs, you rarely get to rewrite everything.
Interop gives you a migration path:

\begin{itemize}
\item keep most of the app in VDOM where the ergonomics are great
\item drop in HRBR for hotspots (large lists, heavy patching, micro-interactions) where you want tighter control
\item avoid building ``two apps'' and stitching them manually
\end{itemize}

This design is also a useful general lesson: if you ever embed one rendering runtime into another, you need a \emph{boundary object} that both runtimes can agree on.
In Relax, that boundary object is a wrapper host element.

\section*{Diagram: where HRBR sits inside a VDOM tree}
\begin{center}
\begin{tikzpicture}[
	node distance=10mm and 14mm,
	box/.style={draw, rounded corners, align=center, inner sep=6pt, fill=RelaxLight},
	arrow/.style={-Latex, thick}
]
\node[box] (vdom) {VDOM parent element\\\texttt{<div>} (owned by VDOM)};
\node[box, below=of vdom] (host) {HRBR wrapper host\\\texttt{<span data-hrbr="1">}\\(owned by VDOM)};
\node[box, below=of host] (inner) {Inner DOM produced by HRBR\\(owned by HRBR runtime)};

\draw[arrow] (vdom) -- node[midway, right] {childNode} (host);
\draw[arrow] (host) -- node[midway, right] {contains} (inner);
\end{tikzpicture}
\end{center}

This “wrapper host” trick is what makes generic VDOM replace/move logic still work: VDOM only needs a single stable child node to own.

\section{The tiny contract: an HRBR vnode is just a mount factory}

In the VDOM layer, HRBR is represented by a new vnode type, \texttt{DOM\_TYPES.HRBR}, implemented in \texttt{src/h.ts}.

\subsection{VNode shape}

The HRBR vnode type is defined as \texttt{HrbrVNode}:

\begin{itemize}
\item \texttt{type: DOM\_TYPES.HRBR}
\item \texttt{mount(host: Element) -> instance}
\item internal bookkeeping fields: \texttt{host}, \texttt{instance}, \texttt{el}
\end{itemize}

The important part is that \texttt{mount} is referentially meaningful. Two HRBR vnodes are treated as the ``same'' subtree iff:\\
	texttt{old.mount === new.mount}.

In other words, \emph{mount factory identity is the key}. This identity flows through three separate places that must agree:

\begin{itemize}
\item \texttt{src/nodes-equal.ts}: HRBR equality is \texttt{old.mount === new.mount}.
\item \texttt{src/patch-dom.ts}: HRBR patching reuses the mounted instance on that same condition.
\item \texttt{src/destroy-dom.ts}: HRBR teardown calls the instance hooks and removes the host wrapper.
\end{itemize}

This isn't an accident: it lets the VDOM runtime treat HRBR the same way it treats components. ``Same function identity'' means ``keep state/instance''.

\subsection{A tiny contract you can hold in your head}

If you forget everything else, remember this:

\begin{quote}
The VDOM runtime treats an HRBR vnode like a component instance whose ``type'' is the \texttt{mount} function identity.
\end{quote}

That yields a very concrete set of guarantees (and constraints):

\begin{itemize}
\item If you keep returning the same \texttt{mount} function (same reference), you keep the same mounted HRBR instance.
\item If you return a new \texttt{mount} function reference, you are asking the VDOM runtime to throw away and rebuild that subtree.
\item The VDOM runtime does \emph{not} diff the DOM inside the HRBR host. That\'s HRBR\'s job.
\end{itemize}

This kind of contract shows up in many UI runtimes:

\begin{itemize}
\item React treats function component identity as the key to preserving state (``same function, same hooks state''). See \href{https://react.dev/learn/preserving-and-resetting-state}{https://react.dev/learn/preserving-and-resetting-state}.
\item Solid treats compiled templates as stable ``owners'' that update reactive expressions without remounting DOM. See \href{https://www.solidjs.com/docs/latest\#the-compiler}{https://www.solidjs.com/docs/latest\#the-compiler}.
\end{itemize}

\subsection{HRBR instances: a minimal lifecycle interface}

The VDOM runtime does \emph{not} understand HRBR internals. It only requires that \texttt{mount(host)} returns an object with the following hooks:

\begin{itemize}
\item \texttt{destroy(): void} --- required (DOM and internal cleanup).
\item \texttt{dispose?: () => void} --- optional (reactivity teardown, unsubscribe).
\item \texttt{update?: (values: any) => void} --- optional (to push new values).
\end{itemize}

This interface matches the objects returned from the HRBR runtimes (compiled blocks and fallback mounts), but it's also easy to fake in tests.

\subsection{Creating HRBR vnodes: \texttt{hBlock()}}

The helper \texttt{hBlock()} in \texttt{src/h.ts} wraps a mount factory into an HRBR vnode:

\begin{quote}
\texttt{hBlock((host) => mountCompiledBlock(def, host, compiledSlots))}
\end{quote}

This is the shape the compiler produces (or that you can use manually for experiments).

\subsubsection*{Code: the actual \texttt{hBlock()} helper}

The source implementation is intentionally tiny: it\'s just a tagged vnode:

\begin{lstlisting}[language=JavaScript]
function hBlock(mount) {
	return { type: DOM_TYPES.HRBR, mount }
}
\end{lstlisting}

That simplicity is a feature:

\begin{itemize}
\item the VDOM runtime doesn\'t need to know about blocks, templates, or signals
\item HRBR is ``just another vnode type'' as far as mount/patch/destroy are concerned
\end{itemize}

\section{Mounting: why HRBR gets its own host container}

Mounting happens in \texttt{src/mount-dom.ts}. The main \texttt{mountDOM()} switch includes a new case:

\begin{quote}
\texttt{case DOM\_TYPES.HRBR: createHrbrNode(vdom, parentEl, index)}
\end{quote}

\subsection{\texttt{createHrbrNode()}: mount into a dedicated \texttt{<span>}}

\texttt{createHrbrNode()} creates a dedicated host element (currently a \texttt{span}) and inserts it into the parent:

\begin{itemize}
\item It sets \texttt{host.dataset.hrbr = '1'} to make the host easily recognizable in the DOM.
\item It stores the host on the vnode: \texttt{vdom.host = host}.
\item It calls \texttt{vdom.mount(host)} and stores the returned instance.
\item It sets \texttt{vdom.el = host}.
\end{itemize}

The decision \emph{to set \texttt{vdom.el} to the host container} is subtle but critical.

\subsubsection{Why \texttt{vdom.el = host} matters}

Relax's patcher has a generic ``replace'' path:

\begin{quote}
if nodes aren't equal $\Rightarrow$ find index of \texttt{oldVdom.el} in \texttt{parentEl.childNodes} $\Rightarrow$ destroy old $\Rightarrow$ mount new at the same index.
\end{quote}

That index scan only works if \texttt{oldVdom.el} is actually a direct child of \texttt{parentEl}. For HRBR, the direct child is the wrapper host, not the inner DOM nodes produced by the HRBR runtime.

So \texttt{createHrbrNode()} intentionally makes the wrapper the unit of ownership:

\begin{itemize}
\item moves are moves of the wrapper
\item removals remove the wrapper
\item inner churn stays encapsulated in HRBR
\end{itemize}

Even though the vnode type includes an \texttt{el?: Element} field described as ``first element produced by the block'' in \texttt{src/h.ts}, the mount code currently treats \texttt{el} as the wrapper host. This aligns with ``VDOM owns the wrapper, HRBR owns the inside''.

\subsection{Edge cases this design avoids}

Owning the wrapper host solves a bunch of subtle problems you\'ll hit otherwise:

\begin{itemize}
\item \textbf{Replace correctness:} VDOM replace logic can always locate the subtree by scanning for \texttt{oldVdom.el} among direct children.
\item \textbf{Move correctness:} keyed reorders can move one DOM node (the wrapper) instead of trying to reason about a variable-length range of nodes.
\item \textbf{Encapsulation:} HRBR can freely insert/remove/patch children inside the host without violating VDOM invariants.
\item \textbf{Teardown safety:} removing the host element guarantees the whole subtree is gone, even if an inner runtime makes mistakes.
\end{itemize}

\subsection*{Diagram: mount path (VDOM creates host, HRBR fills it)}
\begin{center}
\begin{tikzpicture}[
	node distance=10mm and 12mm,
	box/.style={draw, rounded corners, align=left, inner sep=6pt, fill=RelaxLight},
	arrow/.style={-Latex, thick}
]
\node[box] (mountdom) {\texttt{mountDOM(vdom, parentEl, index)}};
\node[box, below=of mountdom] (create) {\texttt{createHrbrNode(vdom, parentEl, index)}};
\node[box, below=of create] (host) {Create \texttt{host = document.createElement('span')}\\set \texttt{data-hrbr="1"}\\insert into parent};
\node[box, below=of host] (call) {Call mount factory\\\texttt{instance = vdom.mount(host)}};
\node[box, below=of call] (store) {Store refs\\\texttt{vdom.host = host}\\\texttt{vdom.instance = instance}\\\texttt{vdom.el = host}};

\draw[arrow] (mountdom) -- (create);
\draw[arrow] (create) -- (host);
\draw[arrow] (host) -- (call);
\draw[arrow] (call) -- (store);
\end{tikzpicture}
\end{center}

\subsubsection{A small but useful marker: \texttt{data-hrbr="1"}}

The wrapper uses \texttt{host.dataset.hrbr = '1'}. This isn't required for correctness, but it's very handy when inspecting mixed trees in DevTools: you can quickly spot the boundaries between VDOM-managed and HRBR-managed DOM.

\section{Patching: stability by mount factory identity}

Patching happens in \texttt{src/patch-dom.ts}. The HRBR case delegates to \texttt{patchHrbr(old, next)}.

\subsection{Fast path: stable mount function means reuse}

\texttt{patchHrbr()} first checks referential identity:

\begin{quote}
\texttt{if (old.mount === next.mount) \{ copy host/instance/el; return; \}}
\end{quote}

This mirrors component identity rules: if the ``type'' of the subtree hasn't changed, we preserve the mounted instance.

In the actual implementation, the fast path copies internal bookkeeping from the old vnode to the new vnode:

\begin{itemize}
\item \texttt{host}
\item \texttt{instance}
\item \texttt{el}
\end{itemize}

That copy step is important because the patcher returns the \emph{new} vnode object, and subsequent patches/destroys must be able to reach the original HRBR host and instance.

\subsubsection*{Code: fast-path reuse in \texttt{patchHrbr()}}

The happy path is just a bookkeeping transfer:

\begin{lstlisting}[language=JavaScript]
function patchHrbr(oldVdom, newVdom) {
	if (oldVdom.mount === newVdom.mount) {
		// Keep the existing already-mounted instance.
		newVdom.host = oldVdom.host
		newVdom.instance = oldVdom.instance
		newVdom.el = oldVdom.el
		return
	}

	// ...remount path...
}
\end{lstlisting}

When you get this right, an HRBR subtree behaves like a normal component subtree with preserved state.

\subsection*{Diagram: patch decision (reuse vs remount)}
\begin{center}
\begin{tikzpicture}[
	node distance=10mm and 14mm,
	box/.style={draw, rounded corners, align=left, inner sep=6pt, fill=RelaxLight},
	decision/.style={draw, diamond, aspect=2, align=center, inner sep=2pt, fill=RelaxLight},
	arrow/.style={-Latex, thick}
]
\node[box] (in) {Inputs\\\texttt{oldHrbrVNode, nextHrbrVNode}};
\node[decision, below=of in] (same) {\texttt{old.mount === next.mount}?};
\node[box, below left=of same] (reuse) {Reuse instance\\copy \texttt{host/instance/el}\\no DOM churn};
\node[box, below right=of same] (remount) {Remount into same host\\\texttt{dispose?()}\\\texttt{destroy()}\\clear host\\\texttt{next.mount(host)}};

\draw[arrow] (in) -- (same);
\draw[arrow] (same) -- node[midway, left] {yes} (reuse);
\draw[arrow] (same) -- node[midway, right] {no} (remount);
\end{tikzpicture}
\end{center}

\subsection{Slow path: mount changed means dispose+destroy+remount}

When \texttt{mount} changes, \texttt{patchHrbr()} treats it as a subtree replacement \emph{inside the same outer HRBR vnode type}:

\begin{itemize}
\item reuse the same \texttt{host} container
\item if the instance supports \texttt{dispose()}, call it
\item call \texttt{destroy()} (required)
\item clear the host's children
\item call \texttt{next.mount(host)} and store the new instance
\end{itemize}

Note the interplay with \texttt{areNodesEqual()}:

\begin{itemize}
\item If \texttt{areNodesEqual(old, next)} returns \emph{false}, \texttt{patchDOM()} goes through the full replace path (destroy old vnode + mount new vnode at the same index).
\item If equality says they are the ``same type'' (both HRBR), then \texttt{patchHrbr()} is responsible for remounting when the mount factory changes.
\end{itemize}

This makes HRBR a first-class citizen in the patcher without teaching the patcher HRBR internals.

\subsubsection*{Code: remount ordering (dispose, destroy, clear, mount)}

The slow path is still small, but order matters:

\begin{lstlisting}[language=JavaScript]
function patchHrbr(oldVdom, newVdom) {
	if (oldVdom.mount === newVdom.mount) return

	const host = oldVdom.host
	const inst = oldVdom.instance

	// 1) detach reactivity / cancel scheduled work
	if (inst && inst.dispose) inst.dispose()

	// 2) cleanup DOM and internal state
	if (inst && inst.destroy) inst.destroy()

	// 3) ensure the host is empty even if destroy() was partial
	while (host.firstChild) host.removeChild(host.firstChild)

	// 4) mount the new subtree into the same wrapper host
	const nextInst = newVdom.mount(host)
	newVdom.host = host
	newVdom.instance = nextInst
}
\end{lstlisting}

There\'s a deliberate ``belt and suspenders'' step here: even if \texttt{destroy()} forgets to remove some nodes, the host is forcibly emptied.
That makes remounts deterministic.

\subsection{A note on where remounts happen}

There are two remount scenarios and they get handled in different layers:

\begin{itemize}
\item \textbf{VNode type changes} (or HRBR \texttt{mount} identity changes per \texttt{areNodesEqual()}): handled by the generic replace-path in \texttt{patchDOM()} (destroy + mount at the same index).
\item \textbf{Same vnode type} but HRBR \texttt{mount} changes: handled by \texttt{patchHrbr()} (dispose/destroy/clear/remount into the same wrapper).
\end{itemize}

The integration tests (Section~\ref{ch:vdom-hrbr-interop}) mostly focus on the second path, because the first isn't HRBR-specific.

\section{Destroying: one call site, two optional hooks}

Destroying happens in \texttt{src/destroy-dom.ts} via \texttt{removeHrbrNode(vdom)}.

The HRBR instance contract is designed to work for both compiled blocks and fallback runtime objects:

\begin{itemize}
\item \texttt{destroy(): void} is required.
\item \texttt{dispose(): void} is optional (signals cleanup / reactive graph unsubscribe).
\end{itemize}

The destroy path:

\begin{enumerate}
\item call \texttt{instance.dispose()} if present
\item call \texttt{instance.destroy()} if present
\item remove the host element itself from the DOM
\end{enumerate}

Because the host wrapper is removed, the HRBR subtree can't leak DOM nodes even if the inner runtime forgets to clean up.

\subsection*{Diagram: destroy ordering (dispose is optional, destroy is required)}
\begin{center}
\begin{tikzpicture}[
	node distance=10mm and 12mm,
	box/.style={draw, rounded corners, align=left, inner sep=6pt, fill=RelaxLight},
	arrow/.style={-Latex, thick}
]
\node[box] (start) {\texttt{destroyDOM(vdom)} hits HRBR case};
\node[box, below=of start] (dispose) {If present: \texttt{instance.dispose()}};
\node[box, below=of dispose] (destroy) {Then: \texttt{instance.destroy()}};
\node[box, below=of destroy] (remove) {Remove wrapper host\\\texttt{host.remove()}};

\draw[arrow] (start) -- (dispose);
\draw[arrow] (dispose) -- (destroy);
\draw[arrow] (destroy) -- (remove);
\end{tikzpicture}
\end{center}

\subsection{Why both \texttt{dispose()} and \texttt{destroy()}?}

This pattern matches the rest of the HRBR runtime in this book:

\begin{itemize}
\item \texttt{dispose()} is for disconnecting from reactive graphs (signals/effects) and releasing scheduled work.
\item \texttt{destroy()} is for deterministic DOM cleanup, including nested mounts.
\end{itemize}

\section{Tests as specs: what the integration tests guarantee}

The behavior in this chapter is backed by a small integration test suite in:

\begin{quote}
	exttt{src/\_\_tests\_\_/hrbr-integration.test.ts}
\end{quote}

It\'s worth reading these tests like a contract.
Here are the key guarantees they encode.

\subsection{Mounting an HRBR block returned from a VDOM component}

The first test verifies a basic shape: a VDOM component can return \texttt{hBlock(mount)}, and mounting the VDOM tree calls \texttt{mount(host)} exactly once.

\begin{lstlisting}[language=JavaScript]
const mount = (host) => {
	const el = document.createElement('div')
	el.textContent = 'HRBR'
	host.appendChild(el)
	return { destroy() { host.innerHTML = '' } }
}

const Comp = defineComponent({
	render() {
		return hBlock(mount)
	},
})
\end{lstlisting}

The point of this test isn\'t the DOM details.
It\'s the ownership boundary: the VDOM runtime gives you a host element, and you\'re free to fill it.

\subsection{Patching: stable mount function preserves instance}

Another test asserts the most important property: if the mount factory is stable, patching doesn\'t remount.

Implementation detail: this works because \texttt{src/nodes-equal.ts} defines HRBR vnode equality as \texttt{old.mount === new.mount}.

\subsection{Patching: changing mount function forces remount}

Finally, a test changes the function identity and expects a dispose/destroy/remount into the same host.
This is the contract you lean on when you want to intentionally reset an HRBR subtree.

\section{Common pitfalls (and how to avoid them)}

\subsection{Pitfall: creating a new mount function every render}

If you write:

\begin{lstlisting}[language=JavaScript]
render() {
	return hBlock((host) => mountBlock(def, host, slots))
}
\end{lstlisting}

then you\'re creating a new function every render, so patching will treat it as a different subtree and remount.

Fix: hoist the mount function to a stable reference (module scope, constructor/init, memo), or structure your API so the mount function is stable and you pass changing values via an \texttt{update()} hook.

\subsection{Pitfall: forgetting that VDOM owns the wrapper, not the inner nodes}

When writing debugging tools or DOM traversal helpers, don\'t assume the HRBR subtree\'s ``root element'' is a direct child of the VDOM parent.
The direct child is always the wrapper host (currently \texttt{<span data-hrbr="1">}).

\subsection{Pitfall: leaking event handlers without a teardown hook}

If an HRBR runtime attaches global listeners (\texttt{window}, \texttt{document}) or schedules work, it should put the cleanup in \texttt{dispose()} and/or \texttt{destroy()}.
The VDOM layer will faithfully call them.

\section{Mini-exercise: design an HRBR block API that updates without remounting}

Goal: avoid remounting on every render.

\begin{enumerate}
\item Define a block mount factory that returns \texttt{\{ update(values), destroy() \}}.
\item Keep the \texttt{mount} function reference stable.
\item On VDOM component updates, call \texttt{instance.update(nextValues)} instead of returning a brand new mount function.
\end{enumerate}

This is the same idea as ``props update'' in a component system, just expressed as a tiny imperative contract at the boundary.

The VDOM runtime calls both (when present) so HRBR implementations can keep those concerns separate.

\section{Tests-as-spec: the integration suite}

The file \texttt{src/\_\_tests\_\_/hrbr-integration.test.ts} is the executable contract for everything above.
It contains four key behaviors.

\subsection{Test harness: what this suite is actually exercising}

This isn't a unit test of the HRBR runtime itself; it's testing the integration points:

\begin{itemize}
\item \texttt{hBlock()} produces a vnode with \texttt{type: DOM\_TYPES.HRBR}.
\item \texttt{mountDOM()} correctly creates the wrapper host and calls the mount factory.
\item \texttt{patchDOM()} decides reuse vs remount using mount identity.
\item \texttt{destroyDOM()} tears down and removes the wrapper.
\end{itemize}

\subsection{1) Mount an HRBR block returned from a VDOM component}

The first test constructs a component whose render returns \texttt{hBlock(mount)}.
The mount function appends a child to the provided host and returns an object with \texttt{destroy()}.

Assertions:

\begin{itemize}
\item mounting the VDOM mounts the HRBR region (DOM contains \texttt{HRBR})
\item the mount function is called exactly once
\item destroying the VDOM destroys the HRBR region (DOM becomes empty)
\end{itemize}

\subsection{2) Event semantics: bind to the host component}

The second test verifies the \emph{host component binding} convention.
An HRBR block is mounted from inside a VDOM component, and the block defines an event slot:\\
\texttt{\{ kind: 'event', path: [], name: 'click' \}}.

The key detail is the \texttt{mountBlock(..., \{ hostComponent: this \})} option.
It ensures the handler's \texttt{this} matches the VDOM component instance, mirroring Relax's VDOM event behavior.

Assertions:

\begin{itemize}
\item clicking the button calls the handler
\item the handler sees \texttt{this === vdom.component}
\item the argument is an \texttt{Event}
\end{itemize}

This test is a key ``mixed tree'' guarantee: HRBR doesn't invent a new event binding model when embedded inside Relax VDOM; it can reuse Relax's convention of binding handlers to the component instance.

\subsection{3) Patch: stable mount identity preserves instance}

The third test mounts a component that always returns \texttt{hBlock(mount)} where \texttt{mount} is stable.
It then patches the VDOM with another render of the same component.

Assertions:

\begin{itemize}
\item \texttt{mount} was called only once
\item the instance array length remains 1
\item DOM remains \texttt{hello}
\end{itemize}

This is the ``no remount'' fast path.

Notice what is (and isn't) asserted:

\begin{itemize}
\item It doesn't require HRBR to implement any diffing---the integration layer simply promises not to tear it down.
\item It asserts the VDOM runtime carries the old instance forward by copying \texttt{host}/\texttt{instance} into the new vnode.
\end{itemize}

\subsection{4) Patch: changing mount identity remounts}

The fourth test swaps \texttt{currentMount} from \texttt{MountA} to \texttt{MountB} and triggers a component re-render via \texttt{updateState()}.

Assertions:

\begin{itemize}
\item \texttt{MountA} called once
\item \texttt{MountB} called once
\item DOM changes from \texttt{A} to \texttt{B}
\end{itemize}

The mechanism used here is also important: the test triggers a component re-render via \texttt{updateState()}, which causes the component's \texttt{render()} to run again and return an HRBR vnode with a different \texttt{mount} reference. This matches real application code where mount factories can accidentally change due to closures.

\section{Practical guidance (and foot-guns)}

\begin{itemize}
\item \textbf{Keep mount factories stable} if you want patch-time reuse. In practice, that means hoisting the mount function out of render (or memoizing it).
\item \textbf{Always implement \texttt{destroy()}} in your HRBR instance. Without it, detaching the host wrapper still removes DOM, but reactive resources and event hooks can leak.
\item \textbf{Pass \texttt{hostComponent}} when you mount blocks from components, if you want event handlers to behave like VDOM events.
\item \textbf{Don't assume DOM ownership beyond the host}. The VDOM layer treats the host wrapper as the unit of reconciliation.
\end{itemize}

\subsection{Common failure modes in mixed trees}

\begin{itemize}
\item \textbf{Accidental remounts:} if you create the mount function inline (a new closure every render), it will remount every patch. Prefer a stable function identity.
\item \textbf{Forgetting to clear:} \texttt{patchHrbr()} explicitly clears \texttt{host} before remount. If your custom HRBR instance assumes an empty host but doesn't clean on \texttt{destroy()}, you'll get surprising leftover nodes in custom setups.
\item \textbf{Destroying the wrong thing:} the integration treats the wrapper host as the owned node. Removing inner nodes manually from outside HRBR can break assumptions inside the HRBR runtime.
\end{itemize}

In Part~IV's later chapters, we'll connect this interop layer to the compiler output: the compiler emits \texttt{hBlock(() => mountCompiledBlock(...))} for static templates and \texttt{hBlock(() => mountFallback(...))} when it detects dynamic structure.

\section{Online resources}

\begin{itemize}
\item DOM node insertion and wrapper ownership patterns:\\
\href{https://developer.mozilla.org/en-US/docs/Web/API/Node/appendChild}{MDN: Node.appendChild},\\
\href{https://developer.mozilla.org/en-US/docs/Web/API/Node/insertBefore}{MDN: Node.insertBefore},\\
\href{https://developer.mozilla.org/en-US/docs/Web/API/ChildNode/remove}{MDN: ChildNode.remove}.

\item ``Identity is state'' in UI runtimes (keys and function identity are the rough analogs used here):\\
\href{https://react.dev/learn/rendering-lists#keeping-list-items-in-order-with-key}{React docs: list keys},\\
\href{https://vuejs.org/guide/essentials/list.html#maintaining-state}{Vue docs: maintaining state}.

\item Event handler binding and the meaning of \texttt{this} in JavaScript:\\
\href{https://developer.mozilla.org/en-US/docs/Web/JavaScript/Reference/Operators/this}{MDN: \texttt{this}}.

\item Data attributes used as lightweight, inspectable markers:\\
\href{https://developer.mozilla.org/en-US/docs/Learn/HTML/Howto/Use_data_attributes}{MDN: Using data attributes}.
\end{itemize}
