\chapter{Babel plugin in detail (JSX transform)}
\label{ch:babel-plugin}

This chapter explains the current HRBR Babel transform in \texttt{compiler/jsx/transform.ts}.
It is intentionally a \emph{scaffold}: it covers a useful subset, has crisp failure modes, and is wired end-to-end in jsdom tests.

The transform's job is to recognize TSX/JSX expressions that can be compiled into a stable \texttt{BlockDef} + \texttt{CompiledSlot[]} and to rewrite them into a \texttt{mount(host) => instance} factory that the VDOM interop layer can consume (see Chapter~\ref{ch:vdom-hrbr-interop}).

When it can't safely compile a JSX expression into a stable template, it routes to \texttt{mountFallback()}.

\section{Public entrypoint and options}

The plugin is exported as \texttt{hrbrJsxTransform}:

\begin{itemize}
\item \texttt{compiler/jsx/index.ts}: re-exports the default plugin
\item \texttt{compiler/index.ts}: re-exports it as part of the public compiler API
\end{itemize}

\subsection{Tiny contract: what this plugin must guarantee}

The transform is successful if it guarantees two things:

\begin{itemize}
\item The emitted program is \textbf{self-contained}: all runtime references are imported (or intentionally omitted when not needed).
\item Any transformed JSX expression becomes a \textbf{mount factory} of the form \texttt{(host: Element) => MountedInstance}, matching the HRBR vnode contract from Chapter~\ref{ch:vdom-hrbr-interop}.
\end{itemize}

Everything else (slot heuristics, fallback subset size, key readability) is an implementation choice that can evolve as long as the tests keep passing.

The options type \texttt{HrbrJsxTransformOptions} (in \texttt{compiler/jsx/transform.ts}) controls a few key behaviors:

\begin{itemize}
\item \texttt{runtimeImport}: module path for runtime helpers (default \texttt{"relax/hrbr"})
\item \texttt{mountWrapperName}: helper function name used to call \texttt{mountCompiledBlock} (default \texttt{\_\_hrbrBlock})
\item \texttt{dev}: enables more readable output and debug-oriented metadata
\item \texttt{stableSlotKeys}: when true, derive slot keys from source locations (defaults to \texttt{dev})
\end{itemize}

\section{High-level rewrite shape}

Given a JSX expression like:

\begin{quote}
\texttt{const App = () => <div className=\{cls()\}>Hi \{name()\}</div>}
\end{quote}

The plugin emits three conceptual things:

\begin{enumerate}
\item An import of runtime functions:\\
\texttt{import \{ defineBlock, mountCompiledBlock \} from "relax/hrbr"}
\item A module-scope \texttt{const \_\_hrbr\_block\_N = defineBlock (\ldots)} containing:
  \begin{itemize}
  \item \texttt{templateHTML: "<div ...>..."}
  \item \texttt{slots: \{ key: \{ kind, path, name?, lane? \}, ... \}}
  \end{itemize}
\item A replacement of the JSX expression with a mount factory:\\
\texttt{host => \_\_hrbrBlock(host, \_\_hrbr\_block\_N, [\{key, read\}, ...])}
\end{enumerate}

This shape matches the interop contract in \texttt{src/h.ts}: an HRBR vnode is just a mount function that returns an instance.

\subsection*{Diagram: one JSX expression becomes a hoisted block + mount factory}
\begin{center}
\begin{tikzpicture}[
  node distance=10mm and 14mm,
  box/.style={draw, rounded corners, align=left, inner sep=6pt, fill=RelaxLight},
  arrow/.style={-Latex, thick}
]
\node[box] (input) {Input code\\\texttt{const App = () => <div>Hi \{name()\}</div>}};
\node[box, right=of input] (visit) {Babel visitor\\\texttt{JSXElement(path)}};
\node[box, right=of visit] (emit) {Emit artifacts\\1) hoisted \texttt{defineBlock(...)}\\2) mount factory replaces JSX};

\node[box, below=of emit] (hoist) {Module scope\\\texttt{const \_\_hrbr\_block\_0 = defineBlock(...)};};
\node[box, below=of visit] (replace) {Expression position\\\texttt{(host) => \_\_hrbrBlock(host, \_\_hrbr\_block\_0, [\{key, read\}, ...])}};

\draw[arrow] (input) -- (visit);
\draw[arrow] (visit) -- (emit);
\draw[arrow] (emit) -- (hoist);
\draw[arrow] (emit) -- (replace);
\end{tikzpicture}
\end{center}

The key optimization is hoisting: the template is parsed once per module, while the mount factory runs per render/mount.

\section{Program-level state and synthesis (imports + wrappers)}

The plugin keeps per-file state variables:

\begin{itemize}
\item \texttt{needsRuntimeImport}: whether to inject \texttt{defineBlock/mountCompiledBlock}
\item \texttt{needsFallbackImport}: whether to also import \texttt{mountFallback}
\item \texttt{needsWrapper}: whether to emit the wrapper function \texttt{\_\_hrbrBlock}
\item \texttt{nextBlockId}: a counter used to generate \texttt{\_\_hrbr\_block\_0}, \texttt{\_\_hrbr\_block\_1}, ...
\item \texttt{currentFileName\/currentFileCode}: used for context-rich error messages
\end{itemize}

These are reset in the \texttt{Program.enter} visitor and materialized in \texttt{Program.exit}:

\begin{itemize}
\item \texttt{ensureRuntimeImport(programPath)} uses \texttt{unshiftContainer('body', ImportDeclaration)}.
\item \texttt{ensureWrapper(programPath)} appends \texttt{function \_\_hrbrBlock(host, def, slots) \{ return mountCompiledBlock(def, host, slots) \}}.
\end{itemize}

Implementation note: this scaffold avoids depending on Babel "types" helpers by emitting raw AST node objects.

\subsection*{Diagram: per-file state at \texttt{Program.enter/exit}}
\begin{center}
\begin{tikzpicture}[
  node distance=9mm and 12mm,
  box/.style={draw, rounded corners, align=left, inner sep=6pt, fill=RelaxLight},
  arrow/.style={-Latex, thick}
]
\node[box] (enter) {\texttt{Program.enter}\\reset file state:\\\texttt{needsRuntimeImport=false}\\\texttt{needsFallbackImport=false}\\\texttt{needsWrapper=false}\\\texttt{nextBlockId=0}};
\node[box, right=of enter] (walk) {Walk AST\\each \texttt{JSXElement}:\\\texttt{built = buildTemplateFromJsxElement(...)}\\toggle flags};
\node[box, right=of walk] (exit) {\texttt{Program.exit}\\materialize emissions:\\imports, wrapper helper,\\and \texttt{\_\_hrbr\_block\_N} constants};

\draw[arrow] (enter) -- (walk);
\draw[arrow] (walk) -- (exit);
\end{tikzpicture}
\end{center}

This structure prevents unused imports/helpers: nothing is emitted unless some subtree required it.

\subsection{Why inject at \texttt{Program.exit}?}

The plugin defers emissions until \texttt{Program.exit} so it can answer questions like:

\begin{itemize}
\item Did we ever successfully compile a block? If so, we need \texttt{defineBlock} and \texttt{mountCompiledBlock}.
\item Did any subtree route to fallback? If so, we need \texttt{mountFallback} too.
\item Did we generate a wrapper call? If so, we need to emit the wrapper function body.
\end{itemize}

This design keeps the output minimal: files that contain no transformable JSX don't get imports they don't use.

\section{The core visitor: \texttt{JSXElement(path)}}

The plugin's heart is the \texttt{JSXElement} visitor.
For each JSX element in an expression position, it calls:

\begin{quote}
\texttt{const built = buildTemplateFromJsxElement(path)}
\end{quote}

\subsection{Success path: compile into a block}

When \texttt{built} is non-null:

\begin{enumerate}
\item Mark \texttt{needsRuntimeImport = true} and \texttt{needsWrapper = true}.
\item Insert a module-scope \texttt{const \_\_hrbr\_block\_N = defineBlock(\{ templateHTML, slots \})}.
\item Replace the JSX expression with:\\
\texttt{host => \_\_hrbrBlock(host, \_\_hrbr\_block\_N, built.compiledSlots)}
\end{enumerate}

The slot objects are turned into AST literals through \texttt{literalizeSlot(slot)}.

\subsubsection{Hoisting: why \texttt{defineBlock(...)} lives at module scope}

The plugin uses \texttt{unshiftContainer('body', ...)} on the program node to emit:

\begin{quote}
	texttt{const \_\_hrbr\_block\_N = defineBlock({ templateHTML, slots })}
\end{quote}

Hoisting matters for two reasons:

\begin{itemize}
\item \textbf{Performance:} we want template parsing and slot-path validation to happen once per module, not once per render.
\item \textbf{Identity:} the mount factory will refer to a stable \texttt{\_\_hrbr\_block\_N} constant.
\end{itemize}

\subsection{Failure path: route to fallback}

When \texttt{built} is null, the plugin attempts a \emph{minimal fallback transform}.

\paragraph{Why fallback exists}

Block mode requires a stable template. Expressions that can produce different node shapes at runtime (conditionals, JSX inside expressions, etc.) must be executed by the fallback runtime, which uses the reconciler.

\subsection*{Diagram: fallback routing when template extraction fails}
\begin{center}
\begin{tikzpicture}[
  node distance=10mm and 14mm,
  box/.style={draw, rounded corners, align=left, inner sep=6pt, fill=RelaxLight},
  arrow/.style={-Latex, thick},
  decision/.style={draw, diamond, aspect=2, inner sep=2.5pt, align=center, fill=RelaxLight}
]
\node[box] (jsx) {JSX subtree\\\texttt{<div>...}};
\node[decision, right=of jsx] (ok) {Can build\\stable template?};
\node[box, above right=of ok] (block) {Block transform\\emit \texttt{defineBlock + compiledSlots}};
\node[box, below right=of ok] (fb) {Fallback transform\\emit \texttt{host => mountFallback(...)}\\\texttt{renderFn} returns specs};

\draw[arrow] (jsx) -- (ok);
\draw[arrow] (ok) -- node[midway, above] {yes} (block);
\draw[arrow] (ok) -- node[midway, below] {no} (fb);
\end{tikzpicture}
\end{center}

In other words: block mode is an optimization; fallback mode is the semantic firewall.

\paragraph{Current fallback subset}

The fallback routing in \texttt{transform.ts} is intentionally small:

\begin{itemize}
\item It only supports intrinsic lowercase root tags.
\item It only supports children that are \texttt{JSXText} or \texttt{JSXExpressionContainer}.
\item It allows only static string attributes (and ignores dynamic attrs).
\item It rejects spread attributes and event handlers.
\end{itemize}

In this subset, the plugin synthesizes a \texttt{render()} function (a closure-free data + expression composition) and emits:

\begin{quote}
\texttt{host => mountFallback(host, () => (\{ children: [spec] \}))}
\end{quote}

Each child spec contains:

\begin{itemize}
\item \texttt{create(): Element} -> creates the element and sets attributes/text
\item \texttt{patch(node)} -> updates textContent based on the latest expressions
\item optional \texttt{key} if a \texttt{key=} attribute is present
\end{itemize}

This is enough to prove the end-to-end pipeline and to connect the compiler to the fallback runtime from earlier chapters.

\subsection{Fallback synthesis: what gets captured (and what doesn't)}

The fallback compiler intentionally avoids capturing complex structures.
It builds a \texttt{renderFn} with the shape:\\
	texttt{() => ({ children: [spec] })}

Where \texttt{spec.create()} and \texttt{spec.patch(node)} both close over the user expressions.

Important implications:

\begin{itemize}
\item \textbf{Reactivity comes from runtime:} \texttt{mountFallback} re-runs \texttt{renderFn} under tracking and schedules reconciliation.
\item \textbf{This is not a general JSX renderer:} it is a deliberate minimal bridge used to validate the ``block or fallback'' architecture.
\end{itemize}

\section{Template extraction: \texttt{buildTemplateFromJsxElement()}}

The function \texttt{buildTemplateFromJsxElement()} is a best-effort compiler for a supported subset.
It returns either a full block build or \texttt{null}.

\subsection{Supported subset (today)}

In block mode, the transformer supports:

\begin{itemize}
\item intrinsic lowercase tags only (e.g. \texttt{div}, \texttt{section})
\item nested elements
\item static string attributes
\item expression attributes -> slot kinds \texttt{class/style/attr}
\item dynamic text children -> only when the expression is ``text-like''
\end{itemize}

So far this chapter has described the transform like a black box.
Now we\'ll build the mental model: how a Babel plugin works, which AST nodes it sees, and how this scaffold chooses to represent output.

\section{Mental model: Babel plugins (the minimum you need to know)}

If you\'ve never written a Babel plugin, the key idea is simple:

\begin{itemize}
\item Babel parses your file into an AST (a tree of node objects).
\item A plugin gives Babel a \emph{visitor}: a set of functions like \texttt{Program()}, \texttt{JSXElement()}, \texttt{Identifier()}, etc.
\item As Babel walks the tree, it calls your visitor functions with a \emph{path} object.
  A path is like ``a cursor pointing at this node'', with helpers to replace, insert, and inspect context.
\end{itemize}

In Relax\'s scaffold we rely on three path operations:

\begin{itemize}
\item \texttt{path.replaceWith(newNode)}: turn \texttt{<div/>} into \texttt{host => ...}.
\item \texttt{programPath.unshiftContainer('body', stmt)}: insert imports and hoisted constants.
\item \texttt{programPath.pushContainer('body', stmt)}: append helper functions like \texttt{\_\_hrbrBlock}.
\end{itemize}

What makes compiler work feel hard is not Babel itself.
It\'s the discipline of preserving semantics:

\begin{itemize}
\item when can we treat an expression as \emph{just a string}? (text slot)
\item when does it become \emph{structure}? (must use fallback)
\item how do we keep paths stable so runtime updates don\'t hit the wrong node?
\end{itemize}

\section{A tiny contract for this file (inputs/outputs/error modes)}

Treat \texttt{compiler/jsx/transform.ts} as implementing this contract:

\begin{itemize}
\item \textbf{Input:} a JS/TS program containing JSX elements.
\item \textbf{Output (block success):} the JSX expression is replaced by a mount factory
  	exttt{(host) => \_\_hrbrBlock(host, def, compiledSlots)}.
\item \textbf{Output (fallback):} the JSX expression is replaced by a mount factory
  	exttt{(host) => mountFallback(host, renderFn)}.
\item \textbf{Errors:} unsupported constructs in the scaffold (fragments, spreads, components, events) throw with a friendly code-frame.
\end{itemize}

Notice that both success and fallback produce the \emph{same outer type}: a mount function.
That\'s what makes the rest of the system clean.
The VDOM layer doesn\'t need to know which mode was chosen.

\section{Why this scaffold emits ``raw AST objects'' instead of Babel helpers}

If you\'re used to Babel plugins, you might expect to see heavy usage of \texttt{@babel/types} helpers like \texttt{t.identifier('x')}.
This scaffold goes the other direction: it constructs node objects like:

\begin{lstlisting}[language=JavaScript]
{
  type: 'ImportDeclaration',
  specifiers: [
    { type: 'ImportSpecifier', imported: { type: 'Identifier', name: 'defineBlock' }, local: { type: 'Identifier', name: 'defineBlock' } }
  ],
  source: { type: 'StringLiteral', value: runtimeImport }
}
\end{lstlisting}

Why it matters:

\begin{itemize}
\item It keeps the scaffold dependency surface small.
\item It makes it obvious what Babel is doing: everything is an AST node object.
\item It\'s easier to read snapshots because the output is less abstracted.
\end{itemize}

As the compiler grows, you may still migrate toward Babel helpers for correctness and less boilerplate.
But the raw-node style is an excellent teaching tool.

\section{Program-level synthesis: \texttt{ensureRuntimeImport} and \texttt{ensureWrapper}}

The transform maintains file-scoped flags:\\
  	exttt{needsRuntimeImport}, \texttt{needsFallbackImport}, \texttt{needsWrapper}.

These flags solve a real ``compiler ergonomics'' problem: we don\'t want to inject imports or helpers unless the file actually used JSX in a supported way.

\subsection{How imports are synthesized}

In block mode, we need \texttt{defineBlock} and \texttt{mountCompiledBlock}.
In fallback mode, we also need \texttt{mountFallback}.

So \texttt{ensureRuntimeImport(programPath)} emits exactly:

\begin{lstlisting}[language=JavaScript]
import { defineBlock, mountCompiledBlock } from "relax/hrbr"
// (plus mountFallback if needed)
\end{lstlisting}

This is asserted by \texttt{compiler/\_\_tests\_\_/babel-transform.test.ts}.

\subsection{Why we have a wrapper helper at all}

The wrapper is intentionally boring:

\begin{lstlisting}[language=JavaScript]
function __hrbrBlock(host, def, slots) {
  return mountCompiledBlock(def, host, slots)
}
\end{lstlisting}

So why keep it?
Because it gives you one stable call site to instrument later (devtools, tracing, invariants, feature flags).
A compiler that emits a dozen ad-hoc call sites is harder to evolve.

\section{Template extraction as a recursive renderer}

The heart of block mode is \texttt{buildTemplateFromJsxElement(elPath)}.
It\'s easiest to understand it as a recursive function that tries to \emph{render JSX into HTML} while simultaneously recording slots.

\subsection{Key data structures}

Inside \texttt{buildTemplateFromJsxElement()} the compiler builds three things:

\begin{itemize}
\item \texttt{templateHTML}: a string that can be parsed as an element.
\item \texttt{slotMap}: \texttt{\{ key -> \{ kind, path, name?, lane? \} \}}.
\item \texttt{compiledSlots}: an array of objects with \texttt{\{ key, read: () => expr \}}.
\end{itemize}

The first two belong to \texttt{defineBlock} (module scope).
The third belongs next to the mount call (runtime scope).

\subsection{Paths: tracked while rendering}

The recursive helper \texttt{renderElement(node, pathToHere, childIndexRef)} incrementally builds the DOM path.
Every time we emit a child node into the HTML string, we increment a counter.
When we encounter a dynamic expression that\'s allowed as a text slot, we record:

\begin{quote}
  	exttt{path: [...pathToHere, childIndex]}
\end{quote}

This is exactly the scheme you reverse-engineered in Chapter~\ref{ch:compiler-architecture} via the snapshot.

\subsection{The ``text-like'' gate: why \texttt{isTextLikeExpr} exists}

Remember: \textbf{block mode cannot represent structural variability}.
So if a child expression might yield an element/fragment/conditional structure, we must route to fallback.

In this scaffold the rule is encoded as:\\
  	exttt{if (!isTextLikeExpr(expr)) return null}.

The important thing here is not the exact heuristic.
It\'s the \emph{architectural pattern}:

\begin{itemize}
\item template extraction is best-effort
\item any ambiguous construct returns \texttt{null}
\item the caller chooses fallback
\end{itemize}

This makes the compiler safe by construction.

\section{Fallback synthesis: the semantic firewall (minimal but real)}

When block mode returns \texttt{null}, the transform builds a minimal fallback mount.
You can see this end-to-end in:

\begin{itemize}
\item \texttt{compiler/\_\_tests\_\_/babel-transform.fallback.e2e-jsdom.test.ts}
\end{itemize}

That test compiles:

\begin{lstlisting}[language=JavaScript]
const App = () => (
  <div>{show() ? name() : ''}</div>
)
\end{lstlisting}

Even though the child expression is ``text-like'' at runtime, it\'s a \texttt{ConditionalExpression}, and the current heuristic treats that as structural.
So the plugin emits \texttt{mountFallback} and the test asserts that updates are reactive and do not throw.

\subsection{Mechanics: what the fallback compiler is building}

The fallback compiler doesn\'t build a vdom tree.
It builds a tiny \emph{reconciliation spec} object with two operations:

\begin{itemize}
\item \texttt{create(): Node} --- build initial DOM.
\item \texttt{patch(node): void} --- mutate existing DOM in place.
\end{itemize}

Then it wraps those specs in a \texttt{renderFn}:

\begin{quote}
  	exttt{() => (\{ children: [spec] \})}
\end{quote}

And finally emits:\\
  	exttt{host => mountFallback(host, renderFn)}.

Why this matters: by giving fallback the same outer mounting shape, the rest of Relax stays composable.
Block mode and fallback mode become interchangeable implementations behind one vnode contract.

\section{Failure modes are part of the design (and are tested)}

Not everything should silently route to fallback---especially in a scaffold.
The project currently chooses to throw for several constructs.
That\'s intentional because it makes the supported subset crisp.

The tests in \texttt{compiler/\_\_tests\_\_/babel-transform.errors.test.ts} document the current hard constraints:

\begin{itemize}
\item event handlers like \texttt{onClick}
\item spread attributes like \texttt{\{...props\}}
\item fragments \texttt{<>...</>}
\item non-intrinsic tags like \texttt{<Foo />}
\end{itemize}

As the fallback runtime becomes more feature complete, you can change the policy (throw vs fallback)---but you should always encode it in tests first.

It explicitly rejects (in block mode):

\begin{itemize}
\item event handlers (\texttt{onClick=} etc.)
\item fragments (\texttt{<>...</>})
\item non-intrinsic tags (components, capitalized)
\item spread attributes (\texttt{\{...props\}})
\end{itemize}

\subsection{How paths are computed in the Babel transform}

Like the TS prototype compiler (Chapter~\ref{ch:compiler-architecture}), this transform must produce \texttt{childNodes}-based paths.

In \texttt{renderElement(...)} the transform:

\begin{itemize}
\item keeps a per-parent counter \texttt{childState.i} that increments for each child node
\item uses \texttt{pathToHere} plus the current index to form paths like \texttt{[3,1]}
\item emits a placeholder text node (a single space) when it needs to guarantee the existence of a text node for a dynamic text slot
\end{itemize}

This is why the snapshot test shows a template that contains \emph{extra whitespace} in places where text slots are inserted. That whitespace isn't an aesthetic choice; it's a structural anchor.

\subsection{Slot keys: sequential vs location-based}

Slot keys are allocated in \texttt{allocSlotKey()}.

\begin{itemize}
\item Default: sequential keys \texttt{s0}, \texttt{s1}, ...
\item Dev mode (\texttt{stableSlotKeys=true}): keys include source location: \texttt{s\_\textless{}kind\textgreater{}\_\textless{}line\textgreater:\textless{}col\textgreater}.
\end{itemize}

This is validated in \texttt{compiler/\_\_tests\_\_/babel-transform.test.ts}.

One nuance: location-based slot keys depend on Babel preserving \texttt{loc} information and comments; this is why dev mode is framed as "debug output" rather than a semantic requirement.

\subsection{Lane pragmas: \texttt{/*@lane transition*/}}

The helper \texttt{extractLanePragmaFromExpressionContainer()} scans leading comments and looks for:

\begin{quote}
\texttt{@lane (sync|input|default|transition|idle)}
\end{quote}

If present, the emitted slot metadata includes \texttt{lane: "..."}.
The tests verify that the transformed output contains the lane strings.

\subsection{Text-like safety}

In \texttt{isTextLikeExpr()}, the transform tries to reject expressions that can yield JSX elements or otherwise change DOM structure.

Examples treated as \emph{not text-like}:

\begin{itemize}
\item returning JSX elements/fragments
\item conditionals/logicals that could yield elements (conservatively rejected)
\end{itemize}

When an expression is accepted as text-like, the transform:

\begin{itemize}
\item emits a sentinel space in the template (so a text node exists)
\item records a \texttt{text} slot pointing to that position
\item emits a compiled slot \texttt{read: () => expr}
\end{itemize}

The predicate \texttt{isTextLikeExpr(expr)} is the key safety gate.
It explicitly rejects:

\begin{itemize}
\item \texttt{JSXElement} and \texttt{JSXFragment} (obviously structural)
\item \texttt{ConditionalExpression} and \texttt{LogicalExpression} (conservatively structural for now)
\end{itemize}

This conservative choice is why the fallback e2e test uses \texttt{show() ? name() : ''} to force fallback routing: it is not considered "text-like" yet.

\section{Failure modes are part of the API}

This compiler phase aims for clarity and determinism.
Unsupported syntax produces explicit errors via \texttt{fail(path, message)} which attempts to build Babel code-frame errors.

The file \texttt{compiler/\_\_tests\_\_/babel-transform.errors.test.ts} locks in these failure modes:

\begin{itemize}
\item event handlers should throw
\item spread attributes should throw
\item fragments should throw
\item non-intrinsic tags should throw
\end{itemize}

While these errors might eventually become "route to fallback" behaviors, they're still valuable today: they prevent the compiler from silently generating templates that would be incorrect (for example, by treating an event handler like an attribute).

\section{End-to-end tests: jsdom execution proves output shape}

Two e2e suites validate that the emitted output is not just syntactically correct but \emph{runnable}.

\subsection{Block mode e2e: transformed output mounts and updates}

\texttt{compiler/\_\_tests\_\_/babel-transform.e2e-jsdom.test.ts} compiles a small component, extracts the generated \texttt{defineBlock(...)} literal and the \texttt{compiledSlots} array literal, evaluates them with \texttt{new Function(...)} and mounts the result with \texttt{mountCompiledBlock}.

The key invariant: updating a signal updates the DOM without recreating the root element.

This suite also proves a deeper property: the emitted output can be \emph{partially evaluated}.
The test extracts only the \texttt{defineBlock(...)} object literal and the compiled slots array literal, then evaluates them with \texttt{new Function}.
So the plugin isn't just emitting runnable code; it's emitting code whose pieces behave like data.

\subsection{Fallback routing e2e: structural text expression goes to \texttt{mountFallback}}

\texttt{compiler/\_\_tests\_\_/babel-transform.fallback.e2e-jsdom.test.ts} compiles an expression with a structural conditional:

\begin{quote}
\texttt{<div>\{show() ? name() : ''\}</div>}
\end{quote}

This should include \texttt{mountFallback} in output and remain reactive under signal updates.

The spec here is intentionally small:

\begin{itemize}
\item it asserts that \texttt{mountCompiledBlock} should \emph{not} be called
\item it asserts that mounting succeeds and subsequent signal changes cause updates
\item it does \emph{not} assert perfect minimal DOM operations yet
\end{itemize}

That matches the current purpose of fallback in this compiler phase: correctness-first safety for dynamic structure.

\section{What to improve next (guided by tests)}

The current transform is intentionally conservative.
The next incremental steps typically look like:

\begin{itemize}
\item add safe event slot support (mapping \texttt{onClick} to HRBR event slots)
\item support fragments by lowering them to concatenated children in block mode
\item expand fallback routing to support nested children specs (not just a single root spec)
\item improve text-like analysis so common conditionals can stay in block mode when both arms are text
\end{itemize}

\section{Full trace: one \texttt{JSXElement} visitor call, end to end}

You now have all the pieces.
Let\'s glue them together by tracing a single JSX expression through the plugin.

We\'ll use the smallest example that still exercises the three interesting mechanics:

\begin{itemize}
\item a dynamic attribute (becomes a slot)
\item a nested element (forces recursion)
\item a dynamic text child (becomes a text slot)
\end{itemize}

\subsection{Author input (JSX in user code)}

\begin{lstlisting}[language=JavaScript]
const App = () => (
  <section id="root" className={cls()}>
    <h1>Title</h1>
    <p>Hello {name()}</p>
  </section>
)
\end{lstlisting}

This is the same structure used by the inline snapshot in
  exttt{compiler/\_\_tests\_\_/babel-transform.test.ts}.
That\'s intentional: we want a trace that is already pinned by tests.

\subsection{Step A: \texttt{Program.enter} resets per-file state}

Before any JSX is transformed, the plugin resets:

\begin{itemize}
\item \texttt{needsRuntimeImport = false}
\item \texttt{needsFallbackImport = false}
\item \texttt{needsWrapper = false}
\item \texttt{nextBlockId = 0}
\end{itemize}

This ensures a plugin run is deterministic per file.
If you ever see ``block ids jump'' across compiles, this is the first place to look.

\subsection{Step B: \texttt{JSXElement(path)} attempts block compilation}

When Babel visits the \texttt{<section ...>} node, the visitor calls:

\begin{quote}
  exttt{const built = buildTemplateFromJsxElement(path)}
\end{quote}

Think of \texttt{built} as three outputs:

\begin{itemize}
\item \texttt{built.templateHTML}: the static HTML string
\item \texttt{built.slots}: the slot metadata map (kind + path + name/lane)
\item \texttt{built.compiledSlots}: the runtime \texttt{[{ key, read }]} array
\end{itemize}

If \texttt{built} is null, the visitor routes to fallback.
But in this example it succeeds.

\subsection{Step C: \texttt{buildTemplateFromJsxElement} renders HTML and records slots}

Inside the template builder, \texttt{renderElement} walks:

\begin{enumerate}
\item Root: \texttt{<section id="root" className=\{cls()\}> ...}
\item Child element: \texttt{<h1>Title</h1>}
\item Child element: \texttt{<p>Hello \{name()\}</p>}
\end{enumerate}

Along the way it produces two slots:

\begin{itemize}
\item a \texttt{class} slot on the root element (path \texttt{[]})
\item a \texttt{text} slot inside the \texttt{<p>} node (a deeper path)
\end{itemize}

The important implementation trick is how text slots are anchored:
when a text expression is accepted, the builder emits a literal space in \texttt{templateHTML}.
That guarantees a real text node exists at mount time.

\subsection{Step D: the visitor hoists \texttt{defineBlock(...)} at module scope}

Once we have \texttt{built}, the visitor sets:

\begin{itemize}
\item \texttt{needsRuntimeImport = true}
\item \texttt{needsWrapper = true}
\end{itemize}

And it emits a module-scope constant:

\begin{lstlisting}[language=JavaScript]
const __hrbr_block_0 = defineBlock({
  templateHTML: "...",
  slots: {
    "s0": { kind: "class", path: [] },
    "s1": { kind: "text", path: [3, 1] }
  }
})
\end{lstlisting}

Why this is hoisted: it makes template parsing a one-time cost per module.
At runtime, multiple mounts just reference the same \texttt{\_\_hrbr\_block\_0}.

\subsection{Step E: the JSX expression becomes a mount factory}

Finally, the plugin replaces the original JSX expression with an arrow function:

\begin{lstlisting}[language=JavaScript]
const App = () => host => __hrbrBlock(host, __hrbr_block_0, [
  { key: "s0", read: () => cls() },
  { key: "s1", read: () => name() }
])
\end{lstlisting}

This is the key compiler/runtime boundary:

\begin{itemize}
\item \texttt{defineBlock(...)} is static data (safe to hoist)
\item \texttt{read: () => expr} is dynamic code (must live in runtime scope)
\end{itemize}

The snapshot in \texttt{babel-transform.test.ts} asserts that this shape stays stable.
If you change output shape, you should expect snapshot churn.

\subsection{Step F: \texttt{Program.exit} injects imports and the wrapper helper}

At the end of the file walk, \texttt{Program.exit} materializes any pending artifacts:

\begin{itemize}
\item an import from \texttt{runtimeImport}
\item the \texttt{\_\_hrbrBlock} wrapper function
\end{itemize}

This is the "emit only what you used" policy: no JSX compiled, no injected runtime surface.

\section{Exercises (tests-first, like a real compiler engineer)}

If you want to really internalize this transform, do these small, test-backed exercises.

\subsection{Exercise 1: explain a slot path without running code}

Open the snapshot in \texttt{compiler/\_\_tests\_\_/babel-transform.test.ts} and pick the \texttt{text} slot \texttt{path} (for example \texttt{[3, 1]}).
Write out the \texttt{childNodes} array layout that makes the path correct.

\subsection{Exercise 2: add a new slot kind (carefully)}

The builder already classifies \texttt{class}, \texttt{style}, and generic \texttt{attr}.
Add a test that forces a \texttt{prop} slot (for example \texttt{value=\{val()\}} on \texttt{<input>}).
Then implement the minimal compiler support in \texttt{buildTemplateFromJsxElement}.

\subsection{Exercise 3: widen \texttt{isTextLikeExpr} with a test}

Right now the transform routes \texttt{show() ? name() : ''} to fallback.
Add a unit test that asserts this can remain in block mode \emph{only when} both arms are "text-like".
Then update \texttt{isTextLikeExpr} to accept that case.

If you can complete Exercise 3 without breaking the E2E suites, you\'ve learned the right lesson: compiler changes are always negotiated via tests.

The rule is always: add a targeted test first (unit + e2e), then widen syntax support.

When you widen the subset, remember the three invariants all previous chapters rely on:

\begin{itemize}
\item slot paths must match \texttt{childNodes}
\item stable mount function identity enables reuse in VDOM interop
\item fallback must remain safe when structure can change
\end{itemize}

\section{Online resources}

\begin{itemize}
\item Babel plugin authoring and visitor lifecycle:\\
\href{https://babeljs.io/docs/en/plugins}{Babel docs: plugins},\\
\href{https://babeljs.io/docs/en/babel-plugin-api}{Babel docs: plugin API}.

\item JSX AST node shapes (useful when reading \texttt{JSXElement}, \texttt{JSXText}, and \texttt{JSXExpressionContainer} visitor logic):\\
\href{https://github.com/babel/babel/blob/main/packages/babel-parser/ast/spec.md}{Babel parser AST spec}.

\item Source maps and why transforms try to preserve locations:\\
\href{https://sourcemaps.info/spec.html}{Source Map Revision 3 Proposal}.

\item ``Static template + dynamic slots" as a general pattern (useful for intuition):\\
\href{https://svelte.dev/docs/svelte/what-is-svelte}{Svelte docs: compiler-first model}.
\end{itemize}
