\chapter{Benchmarks methodology (repeatability + fairness)}
\label{ch:benchmarks}

Benchmarks are where performance stories can go wrong fast.
It's easy to produce a chart that looks convincing while accidentally measuring something else (layout thrash, dev-mode checks, GC noise, or even the benchmark harness itself).

This chapter is the “source of truth” for the Relax.js benchmark harness:

\begin{itemize}
\item harness code: \texttt{benchmarks/harness.ts}
\item runner UI: \texttt{benchmarks/index.html} + \texttt{benchmarks/main.ts}
\item methodology notes: \texttt{docs/benchmarks.md}
\end{itemize}

The core goals are:

\begin{itemize}
\item \textbf{Repeatability}: seeded randomness, stable profiles, and a clear warmup/measure separation.
\item \textbf{Fairness}: compare like-for-like work (same DOM outcomes, comparable commit boundaries), and control build modes (dev vs prod).
\item \textbf{Interpretability}: report metrics that catch jank (p95 and frame drops), not just averages.
\end{itemize}

\section{A tiny contract for what we are measuring}

It helps to make the benchmark harness explicit as a small contract:

\begin{itemize}
\item \textbf{Input:} a benchmark case that can mount into a host element and perform one unit of work per tick.
\item \textbf{Output:} per-tick timings and a few summary statistics (avg, p95, drops).
\item \textbf{Success criterion:} when framework \emph{A} is faster than framework \emph{B}, they are doing comparable work and reaching a comparable ``committed'' state before the timer stops.
\item \textbf{Error mode:} if one framework returns early (async batching) while another blocks until DOM is updated, the comparison is not fair.
\end{itemize}

Most of this chapter explains how Relax's harness tries to stay on the right side of that contract.

\section{Benchmarks run in a real browser}

\section{Benchmarks run in a real browser}

The harness is designed to run in a browser environment:

\begin{itemize}
\item Uses \texttt{requestAnimationFrame} to align work with frames.
\item Uses \texttt{performance.now()} for timing.
\item Bundles through Rollup configs that target \texttt{browser: true}.
\end{itemize}

See \texttt{benchmarks/index.html} and the entry script \texttt{benchmarks/main.ts}.

\subsection{Why ``real browser'' matters}

Most UI framework performance costs are dominated by:

\begin{itemize}
\item DOM mutation costs (properties, attributes, insertion/reordering)
\item layout and style recalculation
\item paint/compositing
\item batching policies (microtasks, macrotasks, rAF)
\end{itemize}

These are browser behaviors.
Node + jsdom is useful for correctness tests, but it doesn't model layout/paint and it changes timing behavior.
That's why the canonical benchmark environment is a browser.

\section*{Diagram: how a benchmark tick is measured}
\begin{center}
\begin{tikzpicture}[
	node distance=9mm and 12mm,
	box/.style={draw, rounded corners, align=left, inner sep=6pt, fill=RelaxLight},
	arrow/.style={-Latex, thick}
]
\node[box] (frame) {Await next frame\\\texttt{await nextFrame()}};
\node[box, below=of frame] (t0) {Start timer\\\texttt{t0 = now()}};
\node[box, below=of t0] (tick) {Do one unit of work\\\texttt{controller.tick()}};
\node[box, below=of tick] (commit) {Optional commit barrier\\\texttt{await commit?.()}};
\node[box, below=of commit] (dt) {Stop timer + record sample\\\texttt{dt = now() - t0}};

\draw[arrow] (frame) -- (t0);
\draw[arrow] (t0) -- (tick);
\draw[arrow] (tick) -- (commit);
\draw[arrow] (commit) -- (dt);
\end{tikzpicture}
\end{center}

This is the heart of ``fairness'' in the harness: every framework is asked to do one \texttt{tick()}, and the measured duration only finishes once the case reports that it has committed.

\subsection{A small but important detail: Node/jsdom fallback exists}

The harness is designed for browsers first, but \texttt{benchmarks/harness.ts} contains a compatibility fallback:

\begin{itemize}
\item \texttt{now()} prefers \texttt{performance.now()} but falls back to \texttt{Date.now()}.
\item \texttt{nextFrame()} prefers \texttt{requestAnimationFrame} but falls back to \texttt{setTimeout(..., 16)}.
\end{itemize}

This is \emph{not} intended to make Node-based benchmarking ``accurate''. It's there so that small smoke runs and tooling can still function in non-browser environments.

\section{The harness contract}

The core harness is \texttt{benchmarks/harness.ts}.

\subsection{From scratch: the minimal mental model}

Before looking at the code, anchor the harness in two ideas:

\begin{enumerate}
\item \textbf{We measure per-frame work, not total throughput.} UI jank happens when you miss frame budgets. So the harness is intentionally frame-paced via \texttt{requestAnimationFrame}.
\item \textbf{We only stop the timer when the framework has committed.} Some frameworks batch work asynchronously; if you time only the synchronous part, you can accidentally benchmark \emph{how quickly a framework defers work}, not how quickly it completes it.
\end{enumerate}

You can treat the whole harness as a loop that runs for \(N\) frames:

\begin{itemize}
\item wait for a frame boundary
\item start a timer
\item do ``one tick'' of application work
\item optionally wait for a commit barrier
\item stop the timer and record the sample
\end{itemize}

Everything else in this chapter (profiles, seeds, smoke gates) exists to make that loop repeatable and interpretable.

\subsection{Benchmark case shape}

A benchmark case is a \texttt{BenchmarkCase}:

\begin{itemize}
\item \texttt{name}: stable identifier for reporting
\item \texttt{setup(host)}: returns a controller with\newline
  \texttt{tick()} -> perform one unit of work\newline
  \texttt{teardown()} -> clean up resources and DOM
\item optional \texttt{commit()}: a barrier that resolves only once the framework has committed for the tick
\end{itemize}

The \texttt{commit()} hook is how the harness stays fair across frameworks that batch work differently (e.g. React flush/paint boundaries).

\subsection{The actual types (spec-level)}

The best way to avoid misunderstandings is to show the exact shape of the harness types.
This is a faithful JavaScript-shaped transcription of \texttt{benchmarks/harness.ts} (the real source is TypeScript, but the runtime behavior is the same):

\begin{lstlisting}[language=JavaScript, caption={Harness types: what a benchmark must provide (from \texttt{benchmarks/harness.ts}).}]
// A case is what you plug into the harness.
// setup(host) returns a controller that can do one tick and then clean up.

// BenchmarkCase
const caseShape = {
	name: 'list-10k-1pct:relax-hrbr',
	setup(host) {
		return {
			tick() {
				// do one unit of work (may be sync or async)
			},
			teardown() {
				// remove timers/subscriptions if needed
			},
		}
	},

	// Optional: resolve only when the framework has committed this tick.
	// If a framework updates in a microtask, commit() should await that.
	commit() {
		// optional
	},
}

// runBenchmark options
const optionsShape = {
	warmupFrames: 30,
	measureFrames: 120,
	frameBudgetMs: 16.7,
}
\end{lstlisting}

Two practical notes:

\begin{itemize}
\item \textbf{\texttt{commit()} is a correctness hook, not a performance hack.} Its purpose is to ensure that when we say ``the tick cost is X ms'', we mean ``X ms until the UI is committed'', not ``X ms until the framework scheduled some work''.
\item \textbf{\texttt{teardown()} is part of fairness.} If a case leaks timers or subscriptions, later cases become noisy or slower for reasons unrelated to their own design.
\end{itemize}

\subsection{What counts as ``one tick''? (and why that matters)}

The harness intentionally measures a \emph{steady-state update loop}.
The goal is not to measure initial mount (which can be relevant, but is often a different set of costs), and not to measure bundling or download size.

Each tick should represent something plausible in UI work:

\begin{itemize}
\item update 1\% of a large list
\item shuffle keyed rows
\item update styles in a grid
\item toggle a boolean prop on many nodes
\end{itemize}

If a case is too synthetic (e.g. ``just mutate one text node a million times''), it can still be useful --- but only if we interpret it correctly as a microbenchmark.

\subsection{Tiny contract for a fair \texttt{BenchmarkCase}}

To keep comparisons honest, the benchmark cases should follow a simple contract:

\begin{itemize}
\item \texttt{tick()} should perform one repeatable unit of work and avoid random extra allocations.
\item \texttt{teardown()} must remove event handlers / timers / subscriptions (avoid cross-case leaks).
\item If a framework batches updates asynchronously, provide \texttt{commit()} that resolves after work is actually committed.
\end{itemize}

\subsection{Warmup vs measure windows}

\texttt{runBenchmark()} splits execution into two windows:

\begin{itemize}
\item \textbf{Warmup frames} (default 30): let JIT, memoization, layout caching, and internal queues settle.
\item \textbf{Measure frames} (default 120): record per-tick durations.
\end{itemize}

\section*{Diagram: warmup vs measure windows (and what gets recorded)}
\begin{center}
\begin{tikzpicture}[
	box/.style={draw, rounded corners, align=center, inner sep=6pt, fill=RelaxLight},
	arrow/.style={-Latex, thick}
]
\node[box, minimum width=36mm] (warm) {Warmup\\(no samples)};
\node[box, right=8mm of warm, minimum width=54mm] (measure) {Measure\\(record dt samples)};
\node[box, right=8mm of measure, minimum width=30mm] (report) {Report\\avg, p95, drops};

\draw[arrow] (warm) -- (measure);
\draw[arrow] (measure) -- (report);
\end{tikzpicture}
\end{center}

Warmup frames intentionally do the same work as measured frames; the only difference is whether the duration samples are kept.

Implementation details in \texttt{runBenchmark()}:

\begin{itemize}
\item For each tick: await \texttt{nextFrame()} , record \texttt{t0} , run \texttt{tick()} , await \texttt{commit()} , compute \texttt{dt}.
\item A \textbf{frame drop} is counted when \texttt{dt > frameBudgetMs} (default 16.7ms).
\end{itemize}

\subsection{Reported metrics: avg + p95 + drops}

The harness reports:

\begin{itemize}
\item \texttt{avgMs} , average time per tick
\item \texttt{p95Ms} , 95th percentile tick time (tail latency)
\item \texttt{frameDrops} , number of frames over budget
\end{itemize}

\subsection{How to read results (avg vs p95 vs drops)}

In UI work, \textbf{drops are the user experience}.
Averages can look great while still producing jank if there are occasional slow frames.

\begin{itemize}
\item \textbf{avgMs} answers: ``How much work do we typically do per frame?''
\item \textbf{p95Ms} answers: ``How bad are the spikes?''
\item \textbf{frameDrops} answers: ``How often do we miss our frame budget?''
\end{itemize}

At 60Hz, the typical budget is about \textbf{16.7ms}.
If a case shows \texttt{drops: 120} for \texttt{frames: 120}, it means \emph{every} measured tick exceeded the budget.

This is why the harness reports drops alongside avg/p95: it is the easiest way to distinguish ``fast enough to feel smooth'' from ``fast on average but still janky''.

Percentiles are computed on sorted samples with a simple index selection.
This is good enough for stable comparisons at the harness scale.

The harness computes p95 with:\\
	texttt{idx = floor(p * (n - 1))} (clamped).
This is a simple deterministic estimator, which is a feature here: the same sample list always produces the same percentile.

\subsection{The core measurement loop (the code that matters)}

The harness logic is intentionally small.
Reading it carefully is worth more than reading a dozen benchmark blog posts.
Here is the core loop as it exists in \texttt{benchmarks/harness.ts}, rewritten as JavaScript for the book:

\begin{lstlisting}[language=JavaScript, caption={Core harness loop: warmup then measure (from \texttt{benchmarks/harness.ts}).}]
function now() {
	return typeof performance !== 'undefined' ? performance.now() : Date.now()
}

function nextFrame() {
	if (typeof requestAnimationFrame !== 'undefined') {
		return new Promise((r) => requestAnimationFrame(() => r()))
	}
	// Node/jsdom fallback: approximate and not "benchmark-accurate".
	return new Promise((r) => setTimeout(() => r(), 16))
}

function percentile(sorted, p) {
	if (sorted.length === 0) return 0
	const idx = Math.min(sorted.length - 1, Math.max(0, Math.floor(p * (sorted.length - 1))))
	return sorted[idx]
}

export async function runBenchmark(b, host, options = {}) {
	const warmupFrames = options.warmupFrames ?? 30
	const measureFrames = options.measureFrames ?? 120
	const frameBudgetMs = options.frameBudgetMs ?? 16.7

	const ctrl = b.setup(host)

	// Warmup
	for (let i = 0; i < warmupFrames; i++) {
		await nextFrame()
		await ctrl.tick()
		if (b.commit) await b.commit()
	}

	const samples = []
	let drops = 0

	// Measure
	for (let i = 0; i < measureFrames; i++) {
		await nextFrame()
		const t0 = now()
		await ctrl.tick()
		if (b.commit) await b.commit()
		const dt = now() - t0
		samples.push(dt)
		if (dt > frameBudgetMs) drops++
	}

	ctrl.teardown()

	const avg = samples.reduce((a, x) => a + x, 0) / (samples.length || 1)
	const sorted = samples.slice().sort((a, b2) => a - b2)
	const p95 = percentile(sorted, 0.95)

	return {
		name: b.name,
		samplesMs: samples,
		avgMs: avg,
		p95Ms: p95,
		frames: samples.length,
		frameDrops: drops,
	}
}
\end{lstlisting}

If you only internalize one thing from this chapter, make it this: \textbf{the harness measures how long it takes to do one tick of work \emph{per frame}, including the framework's commit boundary}. That is the closest you can get to the user's experience without building a full product.

\section{Profiles and seeded runs}

Repeatability relies on standard profiles and seeded randomness.

\subsection{Profiles via URL parameters}

The profile parser lives in \texttt{benchmarks/profile.ts}.

\subsection{From scratch: why profiles exist at all}

Without profiles, people do this:

\begin{itemize}
\item one person runs 30 frames
\item another runs 600 frames
\item someone changes warmup from 0 to 60
\item everyone argues about results
\end{itemize}

Profiles solve that socially and technically.
Socially, you can say: ``I ran the \texttt{default} profile with seed=42.''
Technically, the profile defines warmup + measure windows and the frame budget in one place.

In this repo, profiles are intentionally simple: \texttt{quick}, \texttt{default}, \texttt{stress}.

It reads query params from \texttt{benchmarks/index.html}:

\begin{itemize}
\item \texttt{profile}: \texttt{quick | default | stress}
\item \texttt{seed}: integer seed for randomized update patterns
\item \texttt{warmup}, \texttt{frames}, \texttt{budget}: overrides for fine-grained control
\end{itemize}

\subsection{Benchmark variables (every knob you can turn)}

These are the key variables that affect results.
When you publish results, list them, because changing any of them can change the numbers:

\begin{itemize}
\item \textbf{\texttt{profile}}: selects baseline \texttt{warmup} + \texttt{frames} window sizes.
\item \textbf{\texttt{seed}}: makes randomized update patterns repeatable.
\item \textbf{\texttt{warmup}}: number of warmup frames (no samples recorded).
\item \textbf{\texttt{frames}}: number of measured frames (samples recorded).
\item \textbf{\texttt{budget}}: frame budget in milliseconds (drives ``drops'').
\item \textbf{build mode}: dev vs production bundles (especially affects React).
\item \textbf{browser + OS + CPU governor}: real-world environment noise.
\end{itemize}

This harness keeps the first five variables explicit and reproducible.

\subsection{Profile parsing (and the important clamps)}

The profile code is short, but it encodes several methodology decisions:

\begin{itemize}
\item Use named presets (quick/default/stress).
\item Allow overrides via URL query, but \emph{clamp} them to avoid runaway configs.
\item Force \texttt{seed} into uint32 range so the RNG behaves identically.
\end{itemize}

Here is a JavaScript version of \texttt{getBenchmarkProfile()} from \texttt{benchmarks/profile.ts}:

\begin{lstlisting}[language=JavaScript, caption={Profiles from URL params (from \texttt{benchmarks/profile.ts}).}]
function clampInt(n, min, max) {
	return Math.min(max, Math.max(min, n | 0))
}

function readInt(params, key, fallback) {
	const v = params.get(key)
	if (v == null) return fallback
	const n = Number(v)
	return Number.isFinite(n) ? (n | 0) : fallback
}

export function getBenchmarkProfile() {
	const url = typeof location !== 'undefined' ? new URL(location.href) : null
	const params = url?.searchParams ?? new URLSearchParams()

	const name = params.get('profile') ?? 'default'

	let warmupFrames = 30
	let measureFrames = 120
	let frameBudgetMs = 16.7
	let seed = 12345

	switch (name) {
		case 'quick':
			warmupFrames = 5
			measureFrames = 30
			break
		case 'stress':
			warmupFrames = 60
			measureFrames = 300
			break
		case 'default':
		default:
			break
	}

	warmupFrames = clampInt(readInt(params, 'warmup', warmupFrames), 0, 10_000)
	measureFrames = clampInt(readInt(params, 'frames', measureFrames), 1, 100_000)

	const budgetParam = params.get('budget')
	if (budgetParam != null) {
		const b = Number(budgetParam)
		if (Number.isFinite(b) && b > 0) frameBudgetMs = b
	}

	// Keep seed stable across JS engines: force into uint32.
	seed = (readInt(params, 'seed', seed) >>> 0)

	return { name, warmupFrames, measureFrames, frameBudgetMs, seed }
}
\end{lstlisting}

The clamps are boring, but they matter.
They prevent a simple copy/paste typo like \texttt{frames=100000000} from freezing your browser and producing meaningless results.

\subsection{Examples: URL parameters in practice}

The runner prints the profile and seed at the top of the output.
Example:

\begin{quote}
	texttt{Running...\newline profile: default\newline seed: 12345}
\end{quote}

Typical URLs:

\begin{itemize}
\item Quick sanity pass:\\
	texttt{benchmarks/index.html?profile=quick\&seed=1}
\item Longer run that makes tail latency clearer:\\
	texttt{benchmarks/index.html?profile=stress\&seed=42\&budget=16.7}
\item Debugging a suspicious regression by increasing sample size:\\
	texttt{benchmarks/index.html?profile=default\&frames=600\&seed=12345}
\end{itemize}

Defaults are intentionally ``reasonable and stable'' for local runs.

\section*{Diagram: profile parameters flow}
\begin{center}
\begin{tikzpicture}[
	node distance=9mm and 12mm,
	box/.style={draw, rounded corners, align=left, inner sep=6pt, fill=RelaxLight},
	arrow/.style={-Latex, thick}
]
\node[box] (url) {\texttt{index.html?profile=default\&seed=123\&frames=120}};
\node[box, below=of url] (parse) {\texttt{getBenchmarkProfile()} parses + clamps\\returns \texttt{\{ warmup, frames, budgetMs, seed \}}};
\node[box, below=of parse] (rng) {\texttt{rng = mulberry32(seed)}\\drives repeatable update patterns};
\node[box, below=of rng] (run) {\texttt{runBenchmark(profile, case)}};

\draw[arrow] (url) -- (parse);
\draw[arrow] (parse) -- (rng);
\draw[arrow] (rng) -- (run);
\end{tikzpicture}
\end{center}

Implementation note: the profile parser clamps overrides to avoid accidental runaway configurations:

\begin{itemize}
\item \texttt{warmup} is clamped to \texttt{[0, 10000]}
\item \texttt{frames} is clamped to \texttt{[1, 100000]}
\end{itemize}

Seed values are forced into uint32 range using \texttt{>>> 0}, which keeps the PRNG behavior consistent.

\subsection{Seeded RNG}

Randomized patterns (e.g. which list indices change each tick) should be deterministic.
Benchmark code uses a seeded RNG in \texttt{benchmarks/rng.ts}.

A key methodology rule: the seed must be part of the report output so results can be reproduced.

The benchmark RNG implementation is \texttt{Mulberry32} (\texttt{benchmarks/rng.ts}). It is simple, fast, and deterministic---perfect for selecting indices or toggling boolean flags in a way that remains repeatable across runs.

\subsection{The PRNG implementation (and why it is ``good enough'')}

For benchmarks, you do not need cryptographic randomness.
You need:

\begin{itemize}
\item fast number generation
\item stable determinism given the same seed
\item repeatable distributions (so the same indices flip each run)
\end{itemize}

That is why the repo uses Mulberry32 in \texttt{benchmarks/rng.ts}.
Here is the implementation in JavaScript form:

\begin{lstlisting}[language=JavaScript, caption={Mulberry32 PRNG used by benchmark cases (from \texttt{benchmarks/rng.ts}).}]
export function createRng(seed) {
	let a = seed >>> 0
	return {
		float() {
			a = (a + 0x6d2b79f5) >>> 0
			let t = a
			t = Math.imul(t ^ (t >>> 15), t | 1)
			t ^= t + Math.imul(t ^ (t >>> 7), t | 61)
			return ((t ^ (t >>> 14)) >>> 0) / 4294967296
		},
		int(maxExclusive) {
			if (maxExclusive <= 0) return 0
			return (this.float() * maxExclusive) | 0
		},
	}
}
\end{lstlisting}

This RNG is primarily used to answer questions like ``which 1\% of rows change this tick?'' in a deterministic way.
That makes results comparable across refactors: you can be confident you're still measuring the same update pattern.

\section{The benchmark runner UI}

The main entrypoint \texttt{benchmarks/main.ts} wires a button click to sequential benchmark runs.

Key details:

\begin{itemize}
\item It resets \texttt{host.innerHTML = ''} between runs so each case starts clean.
\item It prints formatted results to a \texttt{<pre>}.
\item It runs both HRBR and VDOM variants, and includes adapters for React/Solid.
\end{itemize}

Because different frameworks have different setup/teardown costs, the harness keeps the per-tick loop stable and encourages cases to isolate the steady-state behavior.

\subsection{Why \texttt{host.innerHTML = ''} between runs matters}

Clearing between runs gives you two strong properties:

\begin{itemize}
\item \textbf{Isolation:} DOM nodes and event listeners from a previous framework/case do not bleed into the next.
\item \textbf{Repeatability:} each case starts from the same initial DOM shape.
\end{itemize}

This is also why safe \texttt{teardown()} behavior is part of the case contract: the harness clears DOM, but frameworks may have timers/schedulers that need explicit disposal.

\section{Cases: what they measure}

Cases live under \texttt{benchmarks/cases/}.
The intent is to cover:

\begin{itemize}
\item \textbf{Pure text throughput}: \texttt{text-1m}
\item \textbf{Large lists with sparse updates}: \texttt{list-10k-1pct}
\item \textbf{Many small components/widgets}: \texttt{widgets-200}
\item Extra comparisons in \texttt{cases/more.ts}
\end{itemize}

Important fairness guideline: each case should produce the same DOM output across frameworks (or as close as is reasonable).

\subsection{Case study: \texttt{text-1m}}

In \texttt{benchmarks/cases/text-1m.ts}, the HRBR variant builds the smallest possible compiled block:

\begin{itemize}
\item \texttt{templateHTML: `<div><span> </span></div>`} includes an explicit space so there is a real text node.
\item Slot path \texttt{[0,0]} means: root \texttt{<div>} child 0 is \texttt{<span>}, and its child 0 is the \texttt{Text} node.
\item \texttt{tick()} updates a signal in chunks of \texttt{10,000} so work remains frame-visible.
\end{itemize}

This isn't a ``real app'' workload, but it is an excellent microbenchmark for measuring pure slot-write throughput.

\subsection{What should you expect from \texttt{text-1m}?}

	texttt{text-1m} is a microbenchmark.
It largely measures:

\begin{itemize}
\item how quickly a runtime can update a text node
\item how much overhead there is in the reactive/scheduling layer
\item whether the work produces long-tail spikes (p95) under repeated updates
\end{itemize}

Because it does not exercise reconciliation, keyed diffing, or lots of element updates, it should \emph{not} be used to claim ``framework X is faster for apps''.
It should be used to claim something narrower: ``The slot update path is efficient.''

\subsection{Case study: \texttt{list-10k-1pct}}

In \texttt{benchmarks/cases/list-10k-1pct.ts}, the cases delegate to example mount helpers:

\begin{itemize}
\item HRBR: \texttt{mountList10k1pctHRBR(host, 10000)}
\item VDOM: \texttt{mountList10k1pctVDOM(host, 10000)}
\end{itemize}

The contract is the same for both: return a controller with \texttt{update1pct()} and \texttt{dispose()}.
Methodologically, that matters: both implementations are being asked to do the same "one tick" of work.

\subsection{What \texttt{list-10k-1pct} tries to model}

This is closer to ``real UI pain'':

\begin{itemize}
\item 10,000 rows is a large but plausible virtualized/table workload.
\item Updating 1\% per tick models ``small changes in a large list'' (sorting indicators, status updates, counters).
\item The workload stresses reconciliation overhead and DOM patch strategy.
\end{itemize}

This is where HRBR often wins: it can apply targeted DOM mutations without re-diffing a full VDOM tree.

\subsection{Case study: \texttt{widgets-200}}

	texttt{widgets-200} is a component-shape workload:

\begin{itemize}
\item many small independent widgets
\item small updates in many places
\end{itemize}

This case helps distinguish:

\begin{itemize}
\item runtimes that do well when work is split across many small islands
\item runtimes that pay a fixed overhead per component update
\end{itemize}

Implementation: \texttt{benchmarks/cases/widgets-200.ts}.

\subsection{Extra cases in \texttt{cases/more.ts} (and what each means)}

The runner also includes additional comparisons in \texttt{benchmarks/cases/more.ts}:

\begin{itemize}
\item \texttt{props-toggle-5k}: toggles a boolean property (\texttt{disabled}) on 5,000 buttons.
\item \texttt{style-grid-1k}: updates inline styles on 1,000 elements.
\item \texttt{table-2k-shuffle}: reorders rows each tick (keyed diff/reorder stress).
\item \texttt{svg-1k}: updates attributes on 1,000 SVG circles.
\item \texttt{input-typing-1k}: updates 1,000 input values (a classic ``controlled input'' stress).
\end{itemize}

These give you a more complete story than a single macro statistic.
For example, a runtime can be great at list reconciliation but weaker at style updates, or vice versa.

\section{Build modes matter (dev vs prod)}

Some frameworks change scheduling and checks based on \texttt{process\.env\.NODE\_ENV}.
The benchmark Rollup config sets this value at bundle time.

\begin{itemize}
\item Rollup config: \texttt{rollup.benchmarks.config.mjs}
\item It uses \texttt{@rollup/plugin-replace} to define \texttt{process.env.NODE\_ENV}.
\end{itemize}

NPM scripts in \texttt{package.json} expose common modes:

\begin{itemize}
\item \texttt{build:bench:dev}
\item \texttt{build:bench:prod}
\end{itemize}

To be fair, comparisons should run with comparable modes (usually prod vs prod).

\subsection{What the Rollup configs actually do}

The benchmark bundles are built through two rollup configs:

\begin{itemize}
\item \texttt{rollup.benchmarks.config.mjs} (full UI runner)
\item \texttt{rollup.benchmarks.smoke.config.mjs} (perf-smoke gate)
\end{itemize}

Both use \texttt{@rollup/plugin-replace} to inject:\\
	texttt{process.env.NODE\_ENV = 'production'} by default.

This matters for fairness because some libraries (notably React) perform extra checks and different scheduling policies in development builds.

\section{Perf smoke (CI-friendly)}

For CI and ``obvious regression'' detection, the repo includes a reduced entry:

\begin{itemize}
\item \texttt{benchmarks/perf-smoke.ts}
\item \texttt{benchmarks/perf-smoke.html}
\end{itemize}

It builds via \texttt{rollup.benchmarks.smoke.config.mjs} into \texttt{benchmarks/dist-smoke/}.

\subsection{What perf-smoke is actually specifying}

You can read \texttt{benchmarks/perf-smoke.ts} as an executable specification of the harness:

\begin{itemize}
\item it obtains a profile via \texttt{getBenchmarkProfile()}
\item it forces short defaults unless overridden (default profile becomes \texttt{quick})
\item it runs a small fixed set of cases (currently \texttt{text-1m} and \texttt{list-10k-1pct}, HRBR variants)
\item it emits a JSON report that includes profile, timestamp, and result rows
\item it enforces very wide thresholds to catch catastrophic regressions
\end{itemize}

Even though this isn't a Vitest file, it plays the same role as tests in earlier chapters: it defines what "performance sanity" means for this repository.

\subsection{Perf smoke is a test, but for a different failure mode}

Correctness tests catch functional regressions.
Perf smoke catches a different kind of regression:

\begin{itemize}
\item accidental \(\mathcal{O}(n^2)\) behavior in a hot path
\item accidental "update everything" bug
\item infinite loops or runaway scheduling
\end{itemize}

It is not designed to validate small wins.
It is designed to catch ``something is badly broken'' early.

\subsection{Why the thresholds are wide}

\texttt{perf-smoke.ts} prints a JSON report to \texttt{console.log} and then checks very wide thresholds:

\begin{itemize}
\item fail only when \texttt{avgMs > 250} or \texttt{p95Ms > 500}
\end{itemize}

This is intentional: CI machines are noisy.
The smoke test is designed to catch catastrophic regressions (infinite loops, accidental \(\mathcal{O}(n^2)\) behavior), not micro-optimizations.
This is also why the gate uses two different lenses:

\begin{itemize}
\item \texttt{avgMs}: for sustained runaway behavior
\item \texttt{p95Ms}: for severe tail latency spikes
\end{itemize}

Tail latency matters because user experience is dominated by occasional slow frames.

\subsection{Reading the perf-smoke source as a spec}

	exttt{benchmarks/perf-smoke.ts} is worth reading because it shows how the repo uses \textbf{the same harness} in a CI-friendly way.
Here are the key pieces, again in JavaScript form:

\begin{lstlisting}[language=JavaScript, caption={CI perf smoke: run a small fixed suite and enforce wide thresholds (from \texttt{benchmarks/perf-smoke.ts}).}]
import { runBenchmark } from './harness'
import { getBenchmarkProfile } from './profile'
import { text1mHRBR } from './cases/text-1m'
import { list10k1pctHRBR } from './cases/list-10k-1pct'

async function main() {
	const profile = getBenchmarkProfile()
	const host = document.getElementById('host') ?? document.body.appendChild(document.createElement('div'))

	// Force quick-ish runs unless explicitly overridden.
	const smokeProfile = {
		...profile,
		name: profile.name === 'default' ? 'quick' : profile.name,
		warmupFrames: profile.warmupFrames ?? 10,
		measureFrames: profile.measureFrames ?? 30,
		frameBudgetMs: profile.frameBudgetMs ?? 50,
	}

	const rows = []

	host.innerHTML = ''
	rows.push(await runBenchmark(text1mHRBR(smokeProfile), host, smokeProfile))

	host.innerHTML = ''
	rows.push(await runBenchmark(list10k1pctHRBR(smokeProfile), host, smokeProfile))

	// Very wide gate: fail only on catastrophic regressions.
	const MAX_AVG_MS = 250
	const MAX_P95_MS = 500
	for (const row of rows) {
		if (row.avgMs > MAX_AVG_MS || row.p95Ms > MAX_P95_MS) {
			throw new Error(`perf-smoke threshold exceeded for ${row.name}`)
		}
	}
}
\end{lstlisting}

Notice what this spec is saying:

\begin{itemize}
\item The smoke suite is intentionally small.
\item It uses the same harness and the same case shapes.
\item It prints a machine-readable JSON report (helpful for CI artifacts).
\item It gates only on extremely large time values to avoid flakiness.
\end{itemize}

If you ever want a stricter gate, add it in a separate workflow/job.
Don't tighten this one until you have enough historical CI data to understand the noise floor.

\section{Instrumentation and benchmarks: keep them separate}

Optional runtime instrumentation exists in \texttt{runtime/devtools.ts} (Chapter~\ref{ch:devtools}).

Methodology rule: keep instrumentation disabled during benchmarks unless you are explicitly measuring it.
Otherwise, the measurement includes the measurement.

\section{Online resources}

These links are useful if you want to validate (or improve) the methodology beyond this repo:

\begin{itemize}
\item High Resolution Time API (\texttt{performance.now()}):
\href{https://developer.mozilla.org/en-US/docs/Web/API/Performance/now}{https://developer.mozilla.org/en-US/docs/Web/API/Performance/now}
\item \texttt{requestAnimationFrame} timing model:
\href{https://developer.mozilla.org/en-US/docs/Web/API/Window/requestAnimationFrame}{https://developer.mozilla.org/en-US/docs/Web/API/Window/requestAnimationFrame}
\item User Timing API (marks/measures; helpful for deeper profiling):
\href{https://developer.mozilla.org/en-US/docs/Web/API/User\_Timing\_API}{https://developer.mozilla.org/en-US/docs/Web/API/User\_Timing\_API}
\item ``The mental model'' for why p95 matters (latency vs averages):
\href{https://www.brendangregg.com/Latency.html}{https://www.brendangregg.com/Latency.html}
\end{itemize}

In practice, if you want to correlate benchmark results with DOM op counts:

\begin{itemize}
\item run once with instrumentation off to measure baseline
\item run again with instrumentation enabled (and accept the overhead) to measure the profile of operations
\end{itemize}

Comparing these two runs helps you understand whether a regression is dominated by DOM work, scheduling, or allocation overhead.

\section{Checklist: how to run a fair comparison}

A practical checklist grounded in this repo's structure:

\begin{enumerate}
\item Build benchmark bundle in a known mode (dev or prod) and record it.
\item Use a named profile (quick/default/stress) so warmup and frames are consistent.
\item Fix a seed and include it in the report.
\item Run multiple times and compare distributions (p95 and drops), not just averages.
\item If a result looks surprising, isolate the case and profile it (browser performance tools).
\end{enumerate}

\section{Interpreting a real output (example walkthrough)}

The runner output is designed to be copied into issue reports.
It includes enough context to be reproducible (profile + seed + case names).

Here is a condensed example pattern (taken from a typical run):

\begin{quote}
	texttt{\# list-10k-1pct:relax-hrbr\newline frames: 120\newline avg: 1.333 ms\newline p95: 4.000 ms\newline drops: 0}
\end{quote}

How to read it:

\begin{itemize}
\item \textbf{0 drops} means it stayed under budget on every measured tick. It should feel smooth.
\item \textbf{avg 1.333ms} means the typical tick is cheap.
\item \textbf{p95 4ms} means spikes exist but are still far below 16.7ms.
\end{itemize}

And compare to the same case in VDOM mode:

\begin{quote}
	texttt{\# list-10k-1pct:relax-vdom\newline frames: 120\newline avg: 62.558 ms\newline p95: 94.000 ms\newline drops: 120}
\end{quote}

This is a very different story:

\begin{itemize}
\item \textbf{drops 120/120} means every tick exceeded the frame budget. It will look janky.
\item The avg and p95 values suggest the runtime is doing work closer to the ``tens of milliseconds'' per tick.
\end{itemize}

This sort of comparison is where HRBR tends to win: it avoids large VDOM diff/patch overhead by pre-compiling stable DOM addressing.

\section{Why this benchmark approach is better than ``single number'' benchmarks}

This harness is intentionally simple, but it has a few properties that make it more trustworthy than many ``microbenchmark charts'':

\begin{itemize}
\item \textbf{Per-frame pacing:} \texttt{requestAnimationFrame} forces you to respect frame boundaries.
\item \textbf{Tail metrics:} p95 and frame drops capture user-visible jank.
\item \textbf{Warmup window:} avoids measuring purely JIT/startup effects.
\item \textbf{Commit barrier:} optional \texttt{commit()} avoids lying about async batching.
\item \textbf{Seeded randomness:} makes results reproducible without removing realism.
\item \textbf{Isolation:} host DOM is cleared between runs, and each case has a teardown contract.
\end{itemize}

No benchmark is perfect.
But if you follow the checklist above and publish the variables (profile/seed/budget/mode), this harness gives you results that are much harder to accidentally misinterpret.

\section{Online resources}

\begin{itemize}
\item Measuring performance in the browser and why alternatives to \texttt{Date.now()} matter:\\
\href{https://developer.mozilla.org/en-US/docs/Web/API/Performance/now}{MDN: \texttt{performance.now()}}.

\item About frame budgets and why 16.7ms is the common baseline:\\
\href{https://web.dev/articles/rendering-performance}{web.dev: rendering performance}.

\item Why warmup matters (JIT, inline caches, tiered compilation):\\
\href{https://v8.dev/blog}{V8 blog (background on optimization and warmup effects)}.

\item Percentiles and tail latency as a reporting best practice:\\
\href{https://en.wikipedia.org/wiki/Percentile}{Wikipedia: percentile},\\
\href{https://research.google/pubs/pub40801/}{Google SRE: tail latency (The Tail at Scale)}.

\item General benchmark hygiene and common pitfalls:\\
\href{https://benchmarking.sites/}{benchmarking.sites (community notes)}.
\end{itemize}
