\chapter*{Copyright}
\addcontentsline{toc}{chapter}{Copyright}

\noindent\textbf{Relax.js: A Deep Build Guide}\\
\noindent\textcopyright\ \the\year\ Ravi Kishan and contributors\\

\vspace{0.8em}
\noindent\textbf{License.} The Relax.js source code repository is licensed under the MIT License.
See the \texttt{LICENSE} file in the repository.

\vspace{0.8em}
\noindent\textbf{Book text.} Unless stated otherwise, the text of this book (including all chapters, diagrams, and examples as written text) is licensed under the MIT License.
Permission is hereby granted, free of charge, to any person obtaining a copy of this text and associated files (the \textquotedblleft Text\textquotedblright),
to deal in the Text without restriction, including without limitation the rights to use, copy, modify, merge, publish, distribute, sublicense,
and/or sell copies of the Text, and to permit persons to whom the Text is furnished to do so, subject to the following conditions:

\vspace{0.4em}
\noindent The above copyright notice and this permission notice shall be included in all copies or substantial portions of the Text.

\vspace{0.4em}
\noindent THE TEXT IS PROVIDED \textquotedblleft AS IS\textquotedblright, WITHOUT WARRANTY OF ANY KIND, EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO
THE WARRANTIES OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE
FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION WITH THE TEXT
OR THE USE OR OTHER DEALINGS IN THE TEXT.

\vspace{0.8em}
\noindent\textbf{Trademarks.} All product names, logos, and brands are the property of their respective owners.
All company, product, and service names used in this book are for identification purposes only.

\vspace{0.8em}
\noindent\textbf{No warranty.} This book is provided \emph{\textquotedblleft as is\textquotedblright} without warranty of any kind.
Techniques described here may be unsafe in production without additional hardening.

\vspace{0.8em}
\noindent\textbf{Attribution.} Inspiration for the learning journey is acknowledged in the Developer Notes, including the book \textit{Build a Frontend Web Framework (From First Principles)} by ÁNGEL SOLA ORBAICETA.

\vspace{0.8em}
\noindent\textbf{Project links.}\\
\noindent Source: \href{https://github.com/Ravikisha/Relax.js}{https://github.com/Ravikisha/Relax.js}\\
\noindent NPM: \href{https://npmjs.com/package/relaxcore}{https://npmjs.com/package/relaxcore}\\
\noindent Author: \href{https://ravikishan.me}{https://ravikishan.me}\\
\noindent License: MIT
