\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

Modern UI frameworks can feel like magic because they hide the two hard things:
\emph{time} (scheduling) and \emph{state} (reactivity).
This book is about making both explicit.

It is written from a systems perspective.
We will treat the browser DOM as an output device, treat rendering as a commit protocol, and treat scheduling as part of correctness.

Relax.js is the concrete case study, but the goal is broader:
to give you a rigorous blueprint for building deterministic UI runtimes.

\section*{What kind of book this is (and is not)}
\addcontentsline{toc}{section}{What kind of book this is (and is not)}

This is a \textbf{runtime engineering book}.
It teaches the contracts, invariants, and failure modes of rendering engines.
You should finish it able to build (and debug) your own small runtime.

It is also an \textbf{architecture reference}.
Each subsystem is tied to concrete implementations and tests that lock behavior down.

Finally, it is a \textbf{compiler-assisted UI runtime case study}.
Relax.js contains two execution paths---a general VDOM runtime and an optimized compiled path (HRBR)---and this book explains how they coexist,
what each path is good at, and what trade-offs they impose.

What it is \emph{not}: a marketing guide, an API cookbook, or a general survey of frameworks.
It is a deliberate walk through one codebase to teach transferable ideas.

\section*{Reader promise}
\addcontentsline{toc}{section}{Reader promise}

After finishing this book, you will be able to:

\begin{itemize}
  \item \textbf{Design your own UI runtime} with clear contracts for model \textrightarrow mount \textrightarrow patch \textrightarrow destroy.
  \item \textbf{Implement and test deterministic scheduling} (lanes/priority, budgets, and flush boundaries).
  \item \textbf{Build a compiler-assisted rendering path} based on stable templates and precomputed slot writes (and know when it helps).
  \item \textbf{Evaluate framework internals critically}: reason about correctness, performance, and maintainability using evidence (tests and measurements).
  \item \textbf{Debug production failure modes} (infinite update loops, unbounded microtasks, memory retention, DOM ownership bugs) with a checklist mindset.
\end{itemize}

The goal is not memorization.
The goal is capability.

\section*{What you'll build, piece by piece}
\addcontentsline{toc}{section}{What you'll build, piece by piece}

Each chapter corresponds to a subsystem with a contract you can run and verify.
Relax.js gives us a full stack of runtime mechanisms:

\begin{itemize}
  \item A \textbf{virtual node model} you can reason about.
  \item A \textbf{mount/patch/unmount pipeline} with crisp boundaries.
  \item A \textbf{keyed diff} (including sequence/LIS thinking) good enough to be educational and pragmatic.
  \item A \textbf{scheduler} that makes time explicit: ordering, fairness, and flush boundaries.
  \item A \textbf{reactivity layer} (signals) that makes invalidation explicit.
  \item A \textbf{compiler-assisted rendering path} (HRBR) that turns stable templates into faster DOM work.
  \item A serious commitment to \textbf{verification contracts} (tests that lock behavior down).
\end{itemize}

\section*{Project links}
\addcontentsline{toc}{section}{Project links}

If you want to follow along in code:

\begin{itemize}
  \item GitHub: \href{https://github.com/ravikisha/relax.js}{https://github.com/ravikisha/relax.js}
  \item NPM (package name \texttt{relaxcore}): \href{https://www.npmjs.com/package/relaxcore}{https://www.npmjs.com/package/relaxcore}
\end{itemize}

\section*{Who this book is for}
\addcontentsline{toc}{section}{Who this book is for}

You should read this if you have written front-end applications and you want to understand what your framework is doing under the hood.
If you have never built a renderer before, the first chapters will start from the DOM and work upward.
If you have built one before, you will still get value from the test-driven approach and the compiler section.

\section*{How to read it}
\addcontentsline{toc}{section}{How to read it}

This is a guided build with reference value.
Whenever you see a behavioral claim, you should be able to trace it to code and to a test expectation.

A practical way to use the book:

\begin{enumerate}
  \item Read a chapter.
  \item Open the referenced file in the repository.
  \item Run the tests for that chapter.
  \item Make one small change and watch what breaks.
\end{enumerate}

\section*{A note on links and citations}
\addcontentsline{toc}{section}{A note on links and citations}

External URLs are included in-line for convenience.
When you need background on browser APIs, MDN is the fastest path to shared vocabulary.
For example: \href{https://developer.mozilla.org/}{https://developer.mozilla.org/}.

Relax.js began as a build-along inspired by \textit{Build a Frontend Web Framework (From First Principles)} by \textsc{\mbox{\hspace{0.2em}ÁNGEL SOLA ORBAICETA}}.
This book pushes further into scheduler design, determinism, failure modes, and compiler-assisted fast paths.

\section*{Versioning}
\addcontentsline{toc}{section}{Versioning}

Relax.js is a living project.
The intent of this book is to explain principles and the architecture of \emph{this} codebase.
If the repository evolves, treat this book as a snapshot of a particular design.
